---
title: "HeadAndNeckAnalysis"
author: "Adir Gigi"
date: "2024-09-29"
output: html_document
---

##Requirements

```{r}
source("~/AXL-PD1/src/globals/constants.R")
CODE_DIR_PATH <- "~/Thesis/Figure 1 - Bulk"
OUTPUT_DIR_PATH <- "~/Thesis/Figure 1 - Bulk/outputs"
source(file.path(CODE_DIR_PATH, "utils/Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/Seurat.Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/RNA.seq.Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/07.CYBERSORTx/CIBERSORT.R"))

library(Seurat)
library(SeuratWrappers)
library(ggplot2)
library(dplyr)
BiocManager::install("scran")
library(scran)
```

```{r}
load(file.path(DROPBOX_DIR_PATH, "Shared_RDataObjects/GSDB.list.of.lists.RData"))
load(file.path(DROPBOX_DIR_PATH, "Shared_RDataObjects/GSDB.list.inclusive.RData"))
```

# Load Data

```{r}
SC.3CA.Tirosh.Head_And_Neck.Input.Path <- file.path(CANCER_DB_DIR_PATH, "SC.3CA.Tirosh/Head_and_Neck")
SC.3CA.Tirosh.Head_And_Neck.Output.Path <- file.path(DATA_DIR_PATH, "SC.3CA.Tirosh/Head_and_Neck")

Kurten.2021.Data.Path <- file.path(SC.3CA.Tirosh.Head_And_Neck.Input.Path, "Data_KĂĽrten2021_Head and Neck")
Kurten.2021.Seurat <- load.and.create.seurat(mtx_path = file.path(Kurten.2021.Data.Path, "Exp_data_UMIcounts.mtx"),
                                             cells_path = file.path(Kurten.2021.Data.Path, "Cells.csv"),
                                             genes_path = file.path(Kurten.2021.Data.Path, "Genes.txt"),
                                             metadata_path = file.path(Kurten.2021.Data.Path, "Meta-data.csv"))

Kurten.2021.Seurat <- seurat.pipeline(Kurten.2021.Seurat)
Kurten.2021.Seurat <- RunALRA(Kurten.2021.Seurat)
my.save.RData("Kurten.2021.Seurat", myfolder=SC.3CA.Tirosh.Head_And_Neck.Output.Path, is.load.obj = T)


Cillo.2020.Data.Path <- file.path(SC.3CA.Tirosh.Head_And_Neck.Input.Path, "Data_Cillo2020_Head and Neck")
Cillo.2020.Seurat <- load.and.create.seurat(mtx_path = file.path(Cillo.2020.Data.Path, "Exp_data_UMIcounts.mtx"),
                                             cells_path = file.path(Cillo.2020.Data.Path, "Cells.csv"),
                                             genes_path = file.path(Cillo.2020.Data.Path, "Genes.txt"),
                                             metadata_path = file.path(Cillo.2020.Data.Path, "Meta-data.csv"))

Cillo.2020.Seurat <- seurat.pipeline(Cillo.2020.Seurat)
Cillo.2020.Seurat <- RunALRA(Cillo.2020.Seurat)
my.save.RData("Cillo.2020.Seurat", myfolder=SC.3CA.Tirosh.Head_And_Neck.Output.Path, is.load.obj = T)


Puram.2017.Data.Path <- file.path(SC.3CA.Tirosh.Head_And_Neck.Input.Path, "Data_Puram2017_Head and Neck")
Puram.2017.Seurat <- load.and.create.seurat(mtx_path = file.path(Puram.2017.Data.Path, "Exp_data_TPM.mtx"),
                                             cells_path = file.path(Puram.2017.Data.Path, "Cells.csv"),
                                             genes_path = file.path(Puram.2017.Data.Path, "Genes.txt"),
                                             metadata_path = file.path(Puram.2017.Data.Path, "Meta-data.csv"))
Puram.2017.Seurat <- seurat.pipeline(Puram.2017.Seurat, is_tpm = T)
Puram.2017.Seurat <- RunALRA(Puram.2017.Seurat)
my.save.RData("Puram.2017.Seurat", myfolder=SC.3CA.Tirosh.Head_And_Neck.Output.Path, is.save.obj = T)



```



# Initial Analysis & Plots

##UMAP by cell_type
```{r}
DimPlot(Puram.2017.Seurat, reduction = "umap", group.by = "cell_type") +
  labs(title = "Puram 2017 UMAP") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Puram2017-UMAP-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)

DimPlot(Cillo.2020.Seurat, reduction = "umap", group.by = "cell_type") +
  labs(title = "Cillo 2020 UMAP") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Cillo2020-UMAP-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)

DimPlot(Kurten.2021.Seurat, reduction = "umap", group.by = "cell_type") +
  labs(title = "Kurten 2021 UMAP") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Kurten2021-UMAP-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)


```



##AXL by cell-type for each dataset
```{r}
VlnPlot(Kurten.2021.Seurat, features = "AXL", group.by = "cell_type") +
  labs(title = "Kurten 2021 - AXL By Cell Type (Before ALRA)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Kurten2021-AXL-by-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)

DefaultAssay(Kurten.2021.Seurat) <- "alra"
VlnPlot(Kurten.2021.Seurat, features = "AXL", group.by = "cell_type") +
  labs(title = "Kurten 2021 - AXL By Cell Type (After ALRA)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Kurten2021-AXL-by-cell-type-Alra.png", path = OUTPUT_DIR_PATH, width = 10, height=6)

DefaultAssay(Cillo.2020.Seurat) <- "alra"
VlnPlot(Cillo.2020.Seurat, features = "AXL", group.by = "cell_type")  +
  labs(title = "Cillo 2020 - AXL By Cell Type (After ALRA)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Cillo2020-AXL-by-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)

DefaultAssay(Puram.2017.Seurat) <- "alra"
VlnPlot(Puram.2017.Seurat, features = "AXL", group.by = "cell_type")  +
  labs(title = "Puram 2017 - AXL By Cell Type (After ALRA)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
ggsave(filename = "Puram2017-AXL-by-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)
```


##HAVRC2 by cell-type for Kurten2021 Dataset
```{r}
# After pan-can bulk analysis, we have seen that HAVCR2 is highly correlated with AXL, as well as Macrophages and CAFs.
DefaultAssay(Kurten.2021.Seurat) <- "alra"

VlnPlot(Kurten.2021.Seurat, features = "HAVCR2", group.by = "cell_type")
ggsave(filename = "Kurten2021-HAVCR2-by-cell-type.png", path = OUTPUT_DIR_PATH, width = 10, height=6)
```



# Find Cell-Type AXL Signature

##Correlate Pairs - Find Correlated Pairs with AXL
```{r}

Cell.Types.To.Create.Signatures <- c("Fibroblast", "Macrophage", "Malignant", "T_cell")


find_correlated_pairs_per_cell_type <- function(seurat_obj, target_gene = "AXL", top_n = 1000) {
    results <- list()
    
    seurat_obj_cell_types <- unique(seurat_obj$cell_type)
    Idents(seurat_obj) <- seurat_obj$cell_type
    
    for(curr_cell_type in seurat_obj_cell_types) {
      message(paste("Starting ", curr_cell_type))

      if(!(curr_cell_type %in% seurat_obj_cell_types)) {  
        message(paste("Skipping", curr_cell_type, "- not found in metadata"))
        next
      } 
      
      cell_subset <- subset(seurat_obj, subset = cell_type == curr_cell_type)
      target_gene_vector <- rep(target_gene, nrow(cell_subset) - 1)
      other_genes_vector <- rownames(cell_subset)[rownames(cell_subset) != target_gene]
      pairings <- list(target_gene_vector, other_genes_vector)
      
      correlations <- correlatePairs(cell_subset@assays$alra@data, pairings = pairings)
      
      top_correlated <- as.data.frame(correlations) %>%
          dplyr::filter(!is.na(rho) & FDR < 0.1) %>%
          arrange(desc(rho)) %>%
          head(top_n)
      
      results[[curr_cell_type]] <- top_correlated
    }
    
    return (results)
}

Kurten.2021.correlated.pairs.with.AXL.per.cell.type <- find_correlated_pairs_per_cell_type(Kurten.2021.Seurat, "AXL")
Cillo.2020.correlated.pairs.with.AXL.per.cell.type <- find_correlated_pairs_per_cell_type(Cillo.2020.Seurat, "AXL")
Puram.2017.correlated.pairs.with.AXL.per.cell.type <- find_correlated_pairs_per_cell_type(Puram.2017.Seurat, "AXL")

```




## Combine Correlated Pairs Result For Constructing Cell-Type Specific AXL Signature
```{r}
library(VennDiagram)
library(grid)

AXL.signature.per.cell.type <- list(
  Macrophage = Reduce(intersect, list(Kurten.2021.correlated.pairs.with.AXL.per.cell.type$Macrophage$gene2, Cillo.2020.correlated.pairs.with.AXL.per.cell.type$Macrophage$gene2)),
  Fibroblast = Reduce(intersect, list(Kurten.2021.correlated.pairs.with.AXL.per.cell.type$Fibroblast$gene2, Puram.2017.correlated.pairs.with.AXL.per.cell.type$Fibroblast$gene2)),
  Malignant = Reduce(intersect, list(Kurten.2021.correlated.pairs.with.AXL.per.cell.type$Malignant$gene2, Puram.2017.correlated.pairs.with.AXL.per.cell.type$Malignant$gene2)),
   T_cell = Reduce(intersect, list(Kurten.2021.correlated.pairs.with.AXL.per.cell.type$T_cell$gene2, Puram.2017.correlated.pairs.with.AXL.per.cell.type$T_cell$gene2))
)

# Venn Diagram of T_cell-specific-AXL signatures per study
venn.plot <- venn.diagram(
  x = list(Kurten2021 = Kurten.2021.correlated.pairs.with.AXL.per.cell.type$T_cell$gene2,
           Puram2017 = Puram.2017.correlated.pairs.with.AXL.per.cell.type$T_cell$gene2, 
           Cillo2020 = Cillo.2020.correlated.pairs.with.AXL.per.cell.type$T_cell$gene2),
  category.names = c("Kurten 2021", "Puram 2017", "Cillo 2020"),
    fill = c("red", "blue", "green"),  # Colors for the sets
  alpha = 0.5,  # Transparency (optional)
  label.col = "white",  # Color of the labels (optional)
  cex = 1.5,  # Size of the labels (optional)
  cat.cex = 1.5 , # Size of the category labels (optional)
  filename = NULL,  # Explicitly NULL prevents any missing argument errors
  output = TRUE
)
grid.newpage() # Clears any existing plots to prevent overlap
grid.draw(venn.plot)

```

```{r}
# Named list of the correlated AXL gene lists per study
correlated_genes_lists <- list(
  Kurten = Kurten.2021.correlated.pairs.with.AXL.per.cell.type,
  Cillo  = Cillo.2020.correlated.pairs.with.AXL.per.cell.type,
  Puram  = Puram.2017.correlated.pairs.with.AXL.per.cell.type
)

# Function to compute intersections
get_intersections_per_study <- function(studies_markers, correlated_lists) {
  result <- list()

  for (study in names(correlated_lists)) {
    markers <- studies_markers[[study]]
    correlated <- correlated_lists[[study]]
    
    cell_types <- intersect(names(markers), names(correlated))  # Only process shared cell types
    study_result <- list()

    for (cell_type in cell_types) {
      markers_genes <- as.character(markers[[cell_type]])
      correlated_genes <- as.character(correlated[[cell_type]]$gene2)
      common_genes <- intersect(markers_genes, correlated_genes)
      
      study_result[[cell_type]] <- common_genes
    }

    result[[study]] <- study_result
  }

  return(result)
}

# Call the function
top.markers.and.axl.correlated.pairs.intersections <- get_intersections_per_study(Studies.Markers, correlated_genes_lists)
my.save.RData("top.markers.and.axl.correlated.pairs.intersections", file.path(OUTPUT_DIR_PATH, "RData_objects"), is.save.obj = T)
```


##Figure 1C

```{r}
Kurten.2021.Seurat <- AddMetaData(Kurten.2021.Seurat, Kurten.2021.Seurat@meta.data$patient, "Patient")
DefaultAssay(Kurten.2021.Seurat) <- "alra"
Idents(Kurten.2021.Seurat) <- Kurten.2021.Seurat$cell_type


Kurten.2021.Malignant.Cells <- subset(Kurten.2021.Seurat, subset = cell_type == "Malignant")
Kurten.2021.Malignant.Cells.AXL.Expression <- FetchData(Kurten.2021.Malignant.Cells,  vars = "AXL", layer = "data")
Kurten.2021.Malignant.Cells <- AddModuleScore(Kurten.2021.Malignant.Cells, list(AXL.signature.per.cell.type[["Malignant"]]), name="Correlation_AXL_Signature_In_Malignant")


Kurten.2021.AXL.Stats <- data.frame(
  Patient = Kurten.2021.Malignant.Cells@meta.data$Patient,
  AXL = Kurten.2021.Malignant.Cells.AXL.Expression$AXL,
  AXL_By_Signature = Kurten.2021.Malignant.Cells@meta.data$Correlation_AXL_Signature_In_Malignant1
) %>% group_by(Patient) %>%
  dplyr::summarize(mean_AXL = mean(AXL, na.rm = TRUE),
                   median_AXL = median(AXL, na.rm = TRUE),
                   mean_correlation_score_from_axl_signature = mean(AXL_By_Signature, na.rm = TRUE),
                   quantile_0.95_AXL = quantile(AXL, probs = 0.95, na.rm = TRUE),
                   quantile_0.96_AXL = quantile(AXL, probs = 0.96, na.rm = TRUE),
                   quantile_0.97_AXL = quantile(AXL, probs = 0.97, na.rm = TRUE),
                   quantile_0.98_AXL = quantile(AXL, probs = 0.98, na.rm = TRUE),
                   quantile_0.99_AXL = quantile(AXL, probs = 0.99, na.rm = TRUE),
                   quantile_0.995_AXL = quantile(AXL, probs = 0.995, na.rm = TRUE),
                   quantile_0.9995_AXL = quantile(AXL, probs = 0.9995, na.rm = TRUE),
                   max_AXL = max(AXL, na.rm = TRUE)
              )

my.save.RData("Kurten.2021.AXL.Stats", file.path(OUTPUT_DIR_PATH, "RData_objects"))
Signatures.To.Correlate <- list(
  T_Cells = GSDB.list.of.lists$cell.types.main$T.cells,
  T_CD8 = GSDB.list.of.lists$cell.types.main$T.CD8,
  T_CD4 = GSDB.list.of.lists$cell.types.main$T.CD4,
  Tirosh_T_Cell_Exhaustion = GSDB.list.inclusive$Tirosh.T_Cell_exhaustion
)

Kurten.2021.Seurat <- AddModuleScore(Kurten.2021.Seurat, Signatures.To.Correlate, name = names(Signatures.To.Correlate))

Kurten.2021.Signatures.Correlations <- data.frame(
  Patient = Kurten.2021.Seurat@meta.data$Patient,
  T_Cell = Kurten.2021.Seurat@meta.data$T_Cells1,
  T_CD8 = Kurten.2021.Seurat@meta.data$T_CD82,
  T_CD4 = Kurten.2021.Seurat@meta.data$T_CD43,
  Tirosh_T_Cell_Exhaustion = Kurten.2021.Seurat@meta.data$Tirosh_T_Cell_Exhaustion4
) %>% group_by(Patient) %>%
  dplyr::summarize(
    mean_T_Cell_Expression = mean(T_Cell, na.rm = TRUE),
    mean_T_CD8_Expression = mean(T_CD8, na.rm = TRUE),
    mean_T_CD4_Expression = mean(T_CD4, na.rm = TRUE),
    mean_T_Cell_Exhaustion_Expression = mean(Tirosh_T_Cell_Exhaustion, na.rm = TRUE)
  )

my.save.RData("Kurten.2021.Signatures.Correlations", file.path(OUTPUT_DIR_PATH, "RData_objects"))

Kurten.2021.T.Cells <- subset(Kurten.2021.Seurat, subset = cell_type == "T_cell")
Kurten.2021.T.Cells.PD1.Expression <- FetchData(Kurten.2021.T.Cells, vars = "PDCD1", layer = "data")

Kurten.2021.T.Cells.PD1.Stats <- data.frame(
  Patient = Kurten.2021.T.Cells@meta.data$Patient,
  PD1 = Kurten.2021.T.Cells.PD1.Expression$PDCD1
) %>% group_by(Patient) %>%
  dplyr::summarize(mean_PD1_T_Cells = mean(PD1, na.rm = TRUE),
                   median_PD1_T_Cells = median(PD1, na.rm = TRUE),
                   quartile_PD1_T_Cells = quantile(PD1, probs = 0.75, na.rm = TRUE))

# Merging data frames for plotting
Kurten.2021.Merged <- Kurten.2021.AXL.Stats %>%
  inner_join(Kurten.2021.Signatures.Correlations, by = "Patient") %>%
  inner_join(Kurten.2021.T.Cells.PD1.Stats, by = "Patient")
my.save.RData("Kurten.2021.Merged", file.path(OUTPUT_DIR_PATH, "RData_objects"))


# Creating the plot
library(ggplot2)
library(reshape2)

# Reshaping data for easier plotting with multiple Y values
Kurten.2021.Plot.Data <- Kurten.2021.Merged %>%
  select(
    Patient, quantile_0.99_AXL, 
    mean_T_Cell_Expression, mean_T_CD8_Expression, mean_T_CD4_Expression,
    mean_T_Cell_Exhaustion_Expression, mean_PD1_T_Cells, median_PD1_T_Cells, quartile_PD1_T_Cells
  ) %>%
  melt(id.vars = c("Patient", "quantile_0.99_AXL"), 
       variable.name = "Metric", value.name = "Value")

# Plotting
p1 <- ggplot(Kurten.2021.Plot.Data, aes(x = quantile_0.99_AXL, y = Value, color = Metric)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, aes(group = Metric), size = 1) +
  labs(title = "Best-Fitting Lines for Expression Metrics",
       x = "Quantile_0.99_AXL", y = "Expression Value", color = "Expression Metric") +
  theme_minimal()

p2 <- ggplot(Kurten.2021.Plot.Data, aes(x = quantile_0.99_AXL, y = Value, color = Metric)) +
  geom_line(aes(group = Metric), size = 1) +
  geom_point(size = 2) +
  labs(title = "Expression Metrics by Quantile_0.99_AXL",
       x = "Quantile_0.99_AXL", y = "Expression Value", color = "Expression Metric") +
  theme_minimal()

ggsave(file.path(OUTPUT_DIR_PATH, "plot1_bestfit_lines.png"), plot = p1, width = 8, height = 6, dpi = 300)
ggsave(file.path(OUTPUT_DIR_PATH, "plot2_connected_lines.png"), plot = p2, width = 8, height = 6, dpi = 300)
```






# Bulk Analysis - Deconvolution

## Create Signatures Matrix For CIBERSORTx
```{r}
# 1. Determine the cell-types I wish to see in the deconvolution result
# 2. Find cell-type markers of the existing cell-types for each SC dataset
# 3. Intersect cell-type markers of different SC datasets (or another construction) to find cell-specific signature
# 4. From the cell-type specific signature - construct the signature matrix
```

### Load H&N Bulk
```{r}
# 1. Find the cell-types of the bulk dataset I wish to perform deconvolution on

my.save.RData("HNSC.Bulk.Dataset", myfolder = "/local/ofir/OCohenLab/Dropbox/gigiadir/Figure 1 - Bulk/outputs/RData_objects", is.load.obj = T)

Puram.2017.cell.types <- unique(Puram.2017.Seurat$cell_type)
Kurten.2021.cell.types <- unique(Kurten.2021.Seurat$cell_type)
Cillo.2020.cell.types <- unique(Cillo.2020.Seurat$cell_type)

Cell.Types.For.Deconvolution <- Filter(function(x) x != "", unique(c(Puram.2017.cell.types, Kurten.2021.cell.types, Cillo.2020.cell.types)))
```
###Find Cell-Type Markers

```{r}
# 2. Find cell-type markers of the existing cell-types for each SC dataset

find.markers.by.cell.type <- function(seurat, n_markers = 1000) {
  Idents(seurat) <- seurat$cell_type
  markers_presto <- RunPrestoAll(seurat, assay = "alra")
  top_markers <- markers_presto %>% group_by(cluster) %>% top_n(n=n_markers, wt = auc) %>% ungroup()
  genes_by_cluster <- split(top_markers$gene, top_markers$cluster)
  
  return(genes_by_cluster)
}

Kurten.2021.top.markers.by.cell.type <- find.markers.by.cell.type(Kurten.2021.Seurat)
Cillo.2020.top.markers.by.cell.type <- find.markers.by.cell.type(Cillo.2020.Seurat)
Puram.2017.top.markers.by.cell.type <- find.markers.by.cell.type(Puram.2017.Seurat)

my.save.RData("Kurten.2021.top.markers.by.cell.type", "~/Thesis/Figure 1 - Bulk/outputs/RData_objects", is.load.obj = T)
my.save.RData("Cillo.2020.top.markers.by.cell.type", "~/Thesis/Figure 1 - Bulk/outputs/RData_objects", is.load.obj = T)
my.save.RData("Puram.2017.top.markers.by.cell.type", "~/Thesis/Figure 1 - Bulk/outputs/RData_objects", is.load.obj = T)
```


###Construct Cell-Type Signature 

```{r}
# 3. Intersect cell-type markers of different SC datasets (or another construction) to find cell-specific signature
Cell.Type.Signatures <- list()
SIGNATURE_MAX_LENGTH <- 50

Studies.Markers <- list(Puram = Puram.2017.top.markers.by.cell.type,
                        Cillo = Cillo.2020.top.markers.by.cell.type,
                        Kurten = Kurten.2021.top.markers.by.cell.type)

for(cell_type in Cell.Types.For.Deconvolution) {
  markers_per_study <- list()
  
  for(study in names(Studies.Markers)) {
    if (cell_type %in% names(Studies.Markers[[study]])) {
      markers_per_study[[study]] <- Studies.Markers[[study]][[cell_type]]
    }
  }
  
  if(length(markers_per_study) > 1) {
    Cell.Type.Signatures[[cell_type]] <- Reduce(intersect, markers_per_study)
  } else {
    Cell.Type.Signatures[[cell_type]] <- unlist(markers_per_study)
  }
}

# Manually Override Dendritic
{Cell.Type.Signatures}[["Dendritic"]] <- intersect(Kurten.2021.top.markers.by.cell.type$Dendritic,
                                                 Cillo.2020.top.markers.by.cell.type$Dendritic)

# Take the top 50 for each cell-type to construct the signature
Cell.Type.Signatures.For.Matrix <- lapply(Cell.Type.Signatures, function(top_markers_list) top_markers_list[1 : min(length(top_markers_list), SIGNATURE_MAX_LENGTH)])

```

### Construct Signature Matrix
```{r}
# 4. From the cell-type specific signature - construct the signature matrix
CIBERSORTx.Construct.Signature.Matrix <- function(seurat_object) {
  #TODO: IMPLEMENT
}

Kurten.2021.Signature.Matrix <- NULL
my.save.RData("Kurten.2021.Signature.Matrix", "~/Thesis/Figure 1 - Bulk/outputs/RData_objects", is.load.obj = T)
```

## Run Deconvolution
```{r}
Bulk.Path <- "~/Thesis/Figure 1 - Bulk/outputs/HNSC.Bulk.Expression.txt"
write.table(HNSC.Bulk.Dataset$counts, file = Bulk.Path, sep="\t", quote = F, row.names = T, col.names = T, fileEncoding = "UTF-8")

Signature.Matrix.Path <- "~/Thesis/Figure 1 - Bulk/outputs/Kurten.2021.Signature.Matrix.txt"
Bulk.Deconvolution <- CIBERSORT(Signature.Matrix.Path, Bulk.Path)
```

##Create Figure
```{r}
AXL.Signatures.Cell.Types.Bulk.Correlations <- list()
CIBERSORT.Cell.Type.Result <- list()

for(cell_type in names(AXL.signature.per.cell.type)) {
  AXL.Signatures.Cell.Types.Bulk.Correlations[[cell_type]] <- calc.Signatures.AUC(HNSC.Bulk.Dataset$counts,
                     GSDB.list = AXL.signature.per.cell.type[[cell_type]],
                     top.n.percent.of.genes.in.ranking.used = 20)
  
  CIBERSORT.Cell.Type.Result[[cell_type]] <- Bulk.Deconvolution[, cell_type]
}

```

```{r}
library(ggplot2)
library(reshape2)

# Get common cell types and sort them
cell_types <- sort(intersect(names(AXL.Signatures.Cell.Types.Bulk.Correlations),
                             names(CIBERSORT.Cell.Type.Result)))

# Initialize the correlation matrix
cor_matrix <- matrix(NA, nrow = length(cell_types), ncol = length(cell_types),
                     dimnames = list(cell_types, cell_types))

# Compute correlations
for (i in cell_types) {
  for (j in cell_types) {
    x <- as.numeric(AXL.Signatures.Cell.Types.Bulk.Correlations[[i]])
    y <- as.numeric(CIBERSORT.Cell.Type.Result[[j]])
    
    if (length(x) == length(y)) {
      cor_matrix[i, j] <- cor(x, y, method = "pearson")
    } else {
      warning(paste("Skipping:", i, "vs", j, "- mismatched length"))
    }
  }
}

cor_df <- melt(cor_matrix, varnames = c("AXL_Cell_Type", "CIBERSORT_Cell_Type"), value.name = "Correlation")

cor_df$AXL_Cell_Type <- factor(cor_df$AXL_Cell_Type, levels = rev(cell_types))
cor_df$CIBERSORT_Cell_Type <- factor(cor_df$CIBERSORT_Cell_Type, levels = cell_types)

cor_df$label <- ifelse(is.na(cor_df$Correlation), "", sprintf("%.3f", cor_df$Correlation))

# Plot heatmap with labels
p3 <- ggplot(cor_df, aes(x = CIBERSORT_Cell_Type, y = AXL_Cell_Type, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = label), size = 4, color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0,
                       limit = c(-1, 1), name = "Pearson\nCorrelation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 12),
        plot.title = element_text(size = 14, face = "bold")) +
  labs(title = "Correlation Matrix: AXL Signatures Score vs CIBERSORT Score", 
       x = "CIBERSORT Score (Cell Type Proportion)", 
       y = "AXL Signature Cell Type Correlation (AUCell)",
       caption = "Based on 469 patients: Pearson correlation between AXL cell-type signature scores\nand CIBERSORT-inferred cell type score across samples")

ggsave("plot3_correlation_heatmap.png", plot = p3, width = 10, height = 8, dpi = 300)


```


# Figure 1C - Macrophage 
```{r}
Kurten.2021.Seurat <- AddMetaData(Kurten.2021.Seurat, Kurten.2021.Seurat@meta.data$patient, "Patient")
DefaultAssay(Kurten.2021.Seurat) <- "alra"
Idents(Kurten.2021.Seurat) <- Kurten.2021.Seurat$cell_type


Kurten.2021.Macrophage.Cells <- subset(Kurten.2021.Seurat, subset = cell_type == "Macrophage")
Kurten.2021.Macrophage.Cells <- AddModuleScore(Kurten.2021.Macrophage.Cells, list(AXL.signature.per.cell.type[["Macrophage"]]), name="Kurten.2021.Macrophage.Cells")

Kurten.2021.AXL.Stats.Macrophages <- data.frame(
  Patient = Kurten.2021.Macrophage.Cells@meta.data$Patient,
  AXL_By_Signature = Kurten.2021.Macrophage.Cells@meta.data$Kurten.2021.Macrophage.Cells1
) %>% group_by(Patient) %>%
  dplyr::summarize(
                mean_correlation_score_from_axl_signature = mean(AXL_By_Signature, na.rm = TRUE)
              )


Kurten.2021.Signatures.Correlations <- data.frame(
  Patient = Kurten.2021.Seurat@meta.data$Patient,
  T_Cell = Kurten.2021.Seurat@meta.data$T_Cells1,
  T_CD8 = Kurten.2021.Seurat@meta.data$T_CD82,
  T_CD4 = Kurten.2021.Seurat@meta.data$T_CD43,
  Tirosh_T_Cell_Exhaustion = Kurten.2021.Seurat@meta.data$Tirosh_T_Cell_Exhaustion4
) %>% group_by(Patient) %>%
  dplyr::summarize(
    mean_T_Cell_Expression = mean(T_Cell, na.rm = TRUE),
    mean_T_CD8_Expression = mean(T_CD8, na.rm = TRUE),
    mean_T_CD4_Expression = mean(T_CD4, na.rm = TRUE),
    mean_T_Cell_Exhaustion_Expression = mean(Tirosh_T_Cell_Exhaustion, na.rm = TRUE)
  )

Kurten.2021.T.Cells <- subset(Kurten.2021.Seurat, subset = cell_type == "T_cell")
Kurten.2021.T.Cells.PD1.Expression <- FetchData(Kurten.2021.T.Cells, vars = "PDCD1", layer = "data")

Kurten.2021.T.Cells.PD1.Stats <- data.frame(
  Patient = Kurten.2021.T.Cells@meta.data$Patient,
  PD1 = Kurten.2021.T.Cells.PD1.Expression$PDCD1
) %>% group_by(Patient) %>%
  dplyr::summarize(mean_PD1_T_Cells = mean(PD1, na.rm = TRUE),
                   median_PD1_T_Cells = median(PD1, na.rm = TRUE),
                   quartile_PD1_T_Cells = quantile(PD1, probs = 0.75, na.rm = TRUE))

# Merging data frames for plotting
Kurten.2021.Merged <- Kurten.2021.AXL.Stats.Macrophages %>%
  inner_join(Kurten.2021.Signatures.Correlations, by = "Patient") %>%
  inner_join(Kurten.2021.T.Cells.PD1.Stats, by = "Patient")

# Creating the plot
library(ggplot2)
library(reshape2)

# Reshaping data for easier plotting with multiple Y values
Kurten.2021.Plot.Data <- Kurten.2021.Merged %>%
  select(
    Patient, mean_correlation_score_from_axl_signature, 
    mean_T_Cell_Expression, mean_T_CD8_Expression, mean_T_CD4_Expression,
    mean_T_Cell_Exhaustion_Expression, mean_PD1_T_Cells, median_PD1_T_Cells, quartile_PD1_T_Cells
  ) %>%
  melt(id.vars = c("Patient", "mean_correlation_score_from_axl_signature"), 
       variable.name = "Metric", value.name = "Value")

# Plotting
p4 <- ggplot(Kurten.2021.Plot.Data, aes(x = mean_correlation_score_from_axl_signature, y = Value, color = Metric)) +
  geom_point(size = 2) +  # Add points for each data point
  geom_smooth(method = "lm", se = FALSE, aes(group = Metric), size = 1) +  # Add best-fitting lines (linear model)
  labs(
    title = "Best-Fitting Lines for Expression Metrics",
    x = "mean_correlation_score_from_axl_signature",
    y = "Expression Value",
    color = "Expression Metric"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p5 <- ggplot(Kurten.2021.Plot.Data, aes(x = mean_correlation_score_from_axl_signature, y = Value, color = Metric)) +
  geom_line(aes(group = Metric), size = 1) +  # Add lines connecting points for each metric
  geom_point(size = 2) +                     # Add points for each data point
  labs(
    title = "Expression Metrics by mean_correlation_score_from_axl_signature",
    x = "mean_correlation_score_from_axl_signature",
    y = "Expression Value",
    color = "Expression Metric"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

ggsave(file.path(OUTPUT_DIR_PATH, "plot4_bestfit_lines.png"), plot = p4, width = 8, height = 6, dpi = 300)
ggsave(file.path(OUTPUT_DIR_PATH, "plot5_connected_lines.png"), plot = p5, width = 8, height = 6, dpi = 300)
```

