---
title: "CellChatTutorial"
output: html_document
---

#Install Packages
```{r}
source("~/AXL-PD1/src/globals/constants.R")
CODE_DIR_PATH <- "~/Thesis/CellChat/scripts"
install.packages("remotes")
install.packages("fastmap")
remotes::install_github("sqjin/CellChat")
source("~/AXL-PD1/src/requirements.R")
source(file.path(CODE_DIR_PATH, "utils/Utils.FH.R"))
source(file.path(CODE_DIR_PATH, "utils/RNA.seq.Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/CellChat.Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/Seurat.Utils.R"))
source(file.path(CODE_DIR_PATH, "utils/my.utils.R"))

library(CellChat)
library(patchwork)
library(Seurat)

devtools::install_github("DKFZ-E220/HNSCatlas")
```

#Load Seurat Objects
```{r}
SC.3CA.Tirosh.Head_And_Neck.Input.Path <- file.path(DATA_DIR_PATH, "SC.3CA.Tirosh/Head_and_Neck")
SC.3CA.Tirosh.BRCA.Input.Path <- file.path(CANCER_DB_DIR_PATH, "SC.3CA.Tirosh/Breast")
DROPBOX.Seurat.Objects.Dir <- file.path(CANCER_DB_DIR_PATH, "SC.3CA.Tirosh/SeuratObjects_10X")

Bassez.2021.Data.Path <- file.path(SC.3CA.Tirosh.BRCA.Input.Path, "Data_Bassez2021_Breast")
Bassez.2021.Seurat <- load.and.create.seurat(mtx_path = file.path(Bassez.2021.Data.Path, "Exp_data_UMIcounts.mtx"),
                                             cells_path = file.path(Bassez.2021.Data.Path, "Cells.csv"),
                                             genes_path = file.path(Bassez.2021.Data.Path, "Genes.txt"),
                                             metadata_path = file.path(Bassez.2021.Data.Path, "Meta-data.csv"))
Bassez.2021.Seurat <- seurat.pipeline(Bassez.2021.Seurat)
Bassez.2021.Seurat <- RunALRA(Bassez.2021.Seurat)

Lin.2020.Data.Path <- file.path(CANCER_DB_DIR_PATH, "SC.3CA.Tirosh/Pancreas/Data_Lin2020_Pancreas")
Lin.2020.Seurat <- load.and.create.seurat(mtx_path = file.path(Lin.2020.Data.Path, "Exp_data_UMIcounts.mtx"),
                                             cells_path = file.path(Lin.2020.Data.Path, "Cells.csv"),
                                             genes_path = file.path(Lin.2020.Data.Path, "Genes.txt"),
                                             metadata_path = file.path(Lin.2020.Data.Path, "Meta-data.csv"))

HNSCC.Atlas.Path = file.path(CANCER_DB_DIR_PATH, "SingleCell_Data_Misc/HNSCC_Atlas.rds")
HNSCC.Atlas <- readRDS(HNSCC.Atlas.Path)

#save for backup
my.save.RData("Kurten.2021.Seurat", myfolder=SC.3CA.Tirosh.Head_And_Neck.Input.Path, is.load.obj = T)
my.save.RData("Bassez.2021.Seurat", myfolder=SC.3CA.Tirosh.BRCA.Input.Path, is.load.obj = T)
my.save.RData("Peng2019_Pancreas.Seurat", myfolder=DROPBOX.Seurat.Objects.Dir, is.load.obj = T)
my.save.RData("Lin2020_Pancreas.Seurat", myfolder=DROPBOX.Seurat.Objects.Dir, is.load.obj = T)

my.save.RData("GSDB.list.of.lists", myfolder=file.path(DROPBOX_DIR_PATH, "Shared_RDataObjects"), is.load.obj = T)

```



# Initial Playground with CellChat
```{r}
data.input <- GetAssayData(Kurten.2021.Seurat, assay="alra", slot="data")
meta <- data.frame(group = Kurten.2021.Seurat@meta.data$cell_type)
rownames(meta) <- colnames(Kurten.2021.Seurat)

cellchat <- createCellChat(object=data.input, meta=meta, group.by = "group")
# cellchat <- createCellChat(object=Kurten.2021.Seurat, group.by = "ident, assay="alra") Should also work

CellChatDB <- CellChatDB.human

#We can Select the L-R pairs from the selected CellChatDB database for cell-cell communication analysis, we can use subsetDB
CellChatDB.use <- CellChatDB

#Subsetting the expression data matrix using genes relevant to the selected L-R pairs. An updated CellChatObject is outputed with two slots: DB and data.signaling
cellchat@DB <- CellChatDB.use
cellchat <- subsetData(cellchat)

#Identify over-expressed ligands or receptors in each cell group to infer the cell state-specific communication
cellchat <- identifyOverExpressedGenes(cellchat)

#For each overexpressed ligand tne receptor obatined in the previous step, identify over-expressed L-R interactions if either its associated ligand or receptor is over expressed
cellchat <- identifyOverExpressedInteractions(cellchat)

cellchat <- computeCommunProb(cellchat, type="triMean", trim=NULL, raw.use=TRUE)
cellchat <- filterCommunication(cellchat, min.cells = 10)
cellchat <- computeCommunProbPathway(cellchat)
cellchat <- aggregateNet(cellchat)
cellchat <- netAnalysis_computeCentrality(cellchat)

netVisual_circle(cellchat@net$count, vertex.weight = T)




# Plot the heatmap with only those
netAnalysis_signalingRole_heatmap(cellchat, signaling = c(
  "WNT",
  "NOTCH",
  "TGFb",
  "EGF",
  "CXCL",
  "VEGF",
  "TNF",
  "IL6",
  "MIF",
  "SPP1"
))

netVisual_circle(
  cellchat@net$weight,
  vertex.weight = TRUE,
  edge.width.max = 5,
  arrow.size = 0.4,
  label.edge = TRUE,
  sources.use = c("Malignant"),
  title.name = "Malignant → Other Interactions",
  margin = 0.05
)
```




# Differential CCC - V1
```{r}
pathways <- unique(CellChatDB$interaction$pathway_name)
CHUNK_SIZE <- 20
pathways_chunks <- split(pathways, ceiling(seq_along(pathways) / CHUNK_SIZE))
```

## H&N (Kurten 2021) - AXL High vs. Low
```{r}
Kurten.2021.Seurat.AXL.Split <- split_seurat_two_groups(
  Kurten.2021.Seurat,
  mode="gene_expression",
  gene="AXL",
  assay="alra",
  slot="data",
  patient_col = "orig.ident",
  agg_fun = "mean",
  cutoff_mode = "quantile",
  q_low = 0.5
)

High.AXL.CellChat <- create_cell_chat_object(Kurten.2021.Seurat.AXL.Split$group1 ,assay="alra")
Low.AXL.CellChat <- create_cell_chat_object(Kurten.2021.Seurat.AXL.Split$group2 ,assay="alra")
High.Low.AXL.merged.CellChat <- mergeCellChat(c(High.AXL.CellChat, Low.AXL.CellChat), add.names = c("High AXL", "Low AXL"))

```

```{r}
# Ensure output directory exists
 output_dir <- "../outputs/dCCC/AXL"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

for (i in seq_along(pathways_chunks)) {
  pathways_chunk <- pathways_chunks[[i]]
  file_name <- file.path(output_dir, sprintf("Comparison_High_AXL_Low_%02d.png", i))

  # build plot
  p <- rankNet(
    High.Low.AXL.merged.CellChat,
    mode = "comparison",
    measure = "weight",
    signaling = pathways_chunk,
  ) + ggplot2::labs(title = "AXL high vs. low dCCC in signaling pathways")

  # write to file (METHOD A: device + print)
  png(file_name, width = 2000, height = 1500, res = 300)
  print(p)                     # <- this actually draws it
  dev.off()
}
```
## do the same proccess generically
```{r}
plot_dccc_chunks(cellchat_merged = TNBC.Luminal.merged.CellChat,
                 pathway_chunks = pathways_chunks,
                 dataset_label = "Breast",
                 comparison_label = "TNBC vs Luminal")
```

## HNSCC (Kurten 2021) - High vs. Low AXL
```{r}
# Key pathways to analyze in Head & Neck (AXL-high vs AXL-low)
pathways_HNSCC <- c(
  "GAS",        # GAS6–AXL core TAM signaling (EMT, resistance, immune suppression)
  "TGFb",       # EMT and Treg/MDSC induction; major axis up in AXL-high
  "EGF",        # Cross-talk with AXL/EGFR; growth & invasion
  "IL6",        # STAT3 inflammatory loop; tumor–immune communication
  "CCL",        # Myeloid recruitment (TAMs, MDSCs)
  "CXCL",       # Neutrophil and MDSC chemotaxis
  "PD-L1",      # Immune checkpoint (tumor or myeloid → T/NK cells)
  "VEGF",       # Angiogenesis and immune exclusion
  "SPP1",       # Osteopontin; macrophage polarization and invasion
  "GALECTIN"    # Immunosuppressive and ECM remodeling signal
)

png("High_AXL_Signaling_Cell-Type.png", width = 2000, height = 1500, res=300)
netAnalysis_signalingRole_heatmap(High.AXL.CellChat, 
                                  signaling = pathways_HNSCC,
                                  title="High AXL Expressing Patients - Signaling Pathways By Cell-Type")
dev.off()

png("Low_AXL_Signaling_Cell-Type.png", width = 2000, height = 1500, res=300)
netAnalysis_signalingRole_heatmap(Low.AXL.CellChat, 
                                  signaling = pathways_HNSCC,
                                  title="Low AXL Expressing Patients - Signaling Pathways By Cell-Type")
dev.off()

png("Low_AXL_Cells_Interactions.png", width = 2000, height = 1500, res=300)
netVisual_circle(
  Low.AXL.CellChat@net$weight,
  vertex.weight = TRUE,
  edge.width.max = 5,
  arrow.size = 0.4,
  label.edge = TRUE,
  sources.use = c("Malignant"),
  title.name = "Low AXL: Malignant → Other Interactions",
  margin = 0.05
)
dev.off()


png("High_AXL_Cells_Interactions.png", width = 2000, height = 1500, res=300)
netVisual_circle(
  High.AXL.CellChat@net$weight,
  vertex.weight = TRUE,
  edge.width.max = 5,
  arrow.size = 0.4,
  label.edge = TRUE,
  sources.use = c("Malignant"),
  title.name = "High AXL: Malignant → Other Interactions",
  margin = 0.05
)
dev.off()

compareInteractions(High.Low.AXL.merged.CellChat, measure="count")

png("Comparison_High_AXL_Low.png", width = 2000, height = 1500, res=300)
rankNet(High.Low.AXL.merged.CellChat, mode="comparison", measure="weight", signaling=pathways_HNSCC)
def.off()
```


## Breast (Bassez 2021) - TNBC vs. ER+/Luminal
```{r}
Bassez.2021.Seurat.BC_Type.Split <- split_seurat_two_groups(
  Bassez.2021.Seurat,
  slot="data",
  mode="metadata",
  condition_col = "BC_type",
  patient_col = "sample",
  group1_values = c("TNBC"),
  group1_label = "TNBC",
  group2_values = c("ER+", "HER2+"),
  group2_label = "ER+/Luminal",
  majority_threshold = 0.5
)
TNBC.CellChat <- create_cell_chat_object(Bassez.2021.Seurat.BC_Type.Split$group1)
Luminal.CellChat <- create_cell_chat_object(Bassez.2021.Seurat.BC_Type.Split$group2)

pathways_Breast<- c(
  "EGF",        # EGFR signaling — hallmark of TNBC
  "NRG",        # NRG1–ERBB3/4 — luminal paracrine growth
  "IL6",        # Pro-inflammatory, higher in TNBC
  "CSF",        # Myeloid/macrophage recruitment (CSF1/CSF1R)
  "CXCL",       # Neutrophil and immune infiltration
  "SPP1",       # TAM–tumor cross-talk and invasion
  "PD-L1",      # Immune checkpoint — TNBC enriched
  "VEGF",       # Angiogenesis and immune suppression
  "FGF",        # Stromal growth factor; luminal/stromal paracrine loop
  "TNF"         # Inflammatory cytokine driving NF-κB signaling
)


png("Breast-TNBC-Signaling-By-Cell-Type.png", width = 2000, height = 1500, res=300)
netAnalysis_signalingRole_heatmap(TNBC.CellChat, 
                                  signaling = pathways_Breast,
                                  title="TNBC - Signaling Pathways By Cell-Type")

png("Breast-Luminal-Signaling-By-Cell-Type.png", width = 2000, height = 1500, res=300)

netAnalysis_signalingRole_heatmap(Luminal.CellChat, 
                                  signaling = pathways_Breast,
                                  title="Luminal - Signaling Pathways By Cell-Type")


TNBC.Luminal.merged.CellChat <- mergeCellChat(c(TNBC.CellChat, Luminal.CellChat), add.names = c("TNBC", "Luminal"))
compareInteractions(TNBC.Luminal.merged.CellChat, measure="weight")
compareInteractions(TNBC.Luminal.merged.CellChat, measure="count")

png("Comparison_TNBC_Luminal.png", width = 2000, height = 1500, res=300)
rankNet(TNBC.Luminal.merged.CellChat, mode="comparison", measure="weight", signaling=pathways_Breast)




```


## dCCC pipeline
```{r}
res_axl <- dCCC.Pipeline(
  seurat_obj = Kurten.2021.Seurat,
  split_args = list(
    mode = "gene_expression",
    gene = "AXL",
    assay = "alra",
    slot  = "data",
    patient_col = "orig.ident",
    agg_fun = "mean",
    cutoff_mode = "quantile",
    q_low = 0.5
  ),
  cellchat_args = list(assay = "alra", slot = "data"),
  add_names = c("High AXL", "Low AXL"),
  pathway_chunks = pathways_chunks,
  dataset_label = "HNSCC",
  comparison_label = "AXL high vs low",
  output_base_dir = "../outputs/dCCC",
  file_format = "png",
  width_px = 2000, height_px = 1500, dpi = 300
)
```


```{r}
Low.AXL.Pathways.IF <- CellChat.Utils.Get.InfoFlow.Pathway(Low.AXL.CellChat)
```




# Differential CCC - V2
```{r}
HNSCC.Atlas.Patients.Objects <- list()
HNSCC.Atlas.AXL.Expression <- add_gene_expression_group(HNSCC.Atlas,
                                                        gene="AXL",
                                                        assay="RNA",
                                                        patient_col_options=c("Patient"))

HNSCC.Atlas.Seurat <- HNSCC.Atlas.AXL.Expression$seurat_obj
HNSCC.Atlas.Patients <- sort(unique(HNSCC.Atlas.Seurat@meta.data$Patient))
```

```{r}

patients.to.run <- setdiff(HNSCC.Atlas.Patients, names(HNSCC.Atlas.Patients.Objects))
                           
for(patient in patients.to.run) {
  message("Processing patient: ", patient)

  patient_cells <- rownames(HNSCC.Atlas.Seurat@meta.data)[
    HNSCC.Atlas.Seurat@meta.data$Patient == patient
  ]

  patient_seurat <- subset(HNSCC.Atlas.Seurat, cells = patient_cells)
  patient_cellchat <- create_cell_chat_object(patient_seurat,
                                              assay="RNA",
                                              cell_type_col = "Cell_Type")
  patient_info_pathways_vec <- CellChat.Utils.Get.InfoFlow.Pathway(patient_cellchat)
  patient_info_lr_vec <- CellChat.Utils.Get.InfoFlow.LR(patient_cellchat)

  patient_summary <- list(
    seurat = patient_seurat,
    cellchat = patient_cellchat,
    infoflow_pathways = patient_info_pathways_vec,
    infoflow_lr = patient_info_lr_vec,
    n_cells = ncol(patient_seurat),
    celltype_counts = table(patient_seurat@meta.data$Cell_Type)
  )

  HNSCC.Atlas.Patients.Objects[[patient]] <- patient_summary
}

my.save.RData("HNSCC.Atlas.Patients.Objects", myfolder="~/Thesis/CellChat/outputs", is.load.obj = T)

```


## Pathways Aggregation
```{r}
infoflow_pathway_list <- lapply(
  HNSCC.Atlas.Patients.Objects,
  function(x) x$infoflow_pathways
)

# All pathways (should be your 192)
all_pathways <- sort(unique(unlist(lapply(infoflow_pathway_list, names))))

# Matrix: rows = pathways, cols = patients
infoflow_pathway_mat <- sapply(infoflow_pathway_list, function(v) {
  x <- v[all_pathways]
  x[is.na(x)] <- 0
  x
})
rownames(infoflow_pathway_mat) <- all_pathways

# Data frame version
infoflow_pathway_df <- as.data.frame(infoflow_pathway_mat)
infoflow_pathway_df$pathway <- rownames(infoflow_pathway_df)
infoflow_pathway_df <- infoflow_pathway_df[, c("pathway", setdiff(colnames(infoflow_pathway_df), "pathway"))]

```


## LR Aggregation
```{r}
infoflow_lr_list <- lapply(
    HNSCC.Atlas.Patients.Objects,
    function(x) x$infoflow_lr
)

all_lr <- sort(unique(unlist(lapply(infoflow_lr_list, names))))

# Matrix: rows = pathways, cols = patients
infoflow_lr_mat <- sapply(infoflow_lr_list, function(v) {
  x <- v[all_lr]
  x[is.na(x)] <- 0
  x
})
rownames(infoflow_lr_mat) <- all_lr
infoflow_lr_df <- as.data.frame(infoflow_lr_mat)
infoflow_lr_df$lr <- rownames(infoflow_lr_df)
infoflow_lr_df <- infoflow_lr_df[, c("lr", setdiff(colnames(infoflow_lr_df), "lr"))]

```

```{r}
library(limma)

High.AXL.Patients <- HNSCC.Atlas.AXL.Expression$patient_summary %>%
  filter(AXL_EXP == "HIGH") %>% select(patient_id) %>% pull(patient_id) %>% intersect(colnames(infoflow_pathway_df))
Low.AXL.Patients <- HNSCC.Atlas.AXL.Expression$patient_summary %>%
  filter(AXL_EXP == "LOW") %>% select(patient_id)  %>% pull(patient_id)%>% intersect(colnames(infoflow_pathway_df))


fit <- diff.expression.limma(
  mat.expr = infoflow_pathway_df,
  is_log_transform = F,
  is_normalize = T,
  is_voom = F,
  robust = F,
  trend = F,
  proportion = 0.01,
  test.cols.numbers = High.AXL.Patients,
  ctrl.cols.numbers = Low.AXL.Patients
)
Pathways.Limma.Fit <- fit

Pathways.Top.Table <- process.topTable(Pathways.Limma.Fit)
Diff.Pathways <- rownames(Pathways.Top.Table)

LR.Limma.Fit <- diff.expression.limma(
  mat.expr = infoflow_lr_df,
  is_log_transform = F,
  is_normalize = T,
  is_voom = F,
  robust = F,
  trend = F,
  proportion = 0.01,
  test.cols.numbers = Low.AXL.Patients,
  ctrl.cols.numbers = High.AXL.Patients
)

LR.Top.Table <- process.topTable(LR.Limma.Fit, number = Inf)
Diff.LR <- rownames(LR.Top.Table)
```

```{r}
HNSCC.Atlas.High.AXL.Seurat <- subset(HNSCC.Atlas.Seurat, subset = AXL_EXP == "HIGH")
HNSCC.Atlas.Low.AXL.Seurat <- subset(HNSCC.Atlas.Seurat, subset = AXL_EXP == "LOW")


High.AXL.CellChat <- create_cell_chat_object(HNSCC.Atlas.High.AXL.Seurat ,assay="RNA", cell_type_col = "Cell_Type")
Low.AXL.CellChat <- create_cell_chat_object(HNSCC.Atlas.Low.AXL.Seurat ,assay="RNA", cell_type_col = "Cell_Type")
High.Low.AXL.merged.CellChat <- mergeCellChat(c(High.AXL.CellChat, Low.AXL.CellChat), add.names = c("High AXL", "Low AXL"))

rankNet(High.Low.AXL.merged.CellChat, mode="comparison", measure="weight", signaling=Diff.LR)
```

## Parse CellChatDB 
```{r}
library(tidyr)

CellChatDB.Ligands.df <- CellChatDB$interaction %>%
  select(interaction_name, interaction_name_2) %>%
  separate(interaction_name_2, into = c("gene", "junk"), sep = " - ", remove = FALSE) %>%
  mutate(role = "ligand") %>%
  select(interaction_name, gene, role)

CellChatDB.Receptors.df <- CellChatDB$interaction %>%
  select(interaction_name, interaction_name_2) %>%
  mutate(gene = str_extract(interaction_name_2, "(?<=\\().*(?=\\))")) %>%
  mutate(gene = ifelse(is.na(gene), str_remove(interaction_name_2, ".* - "), gene)) %>%
  separate_rows(gene, sep = "\\+") %>%
  mutate(role = "receptor") %>%
  select(interaction_name, gene, role)

CellChatDB.Genes.Pathways <- bind_rows(CellChatDB.Ligands.df, CellChatDB.Receptors.df) %>%
  arrange(interaction_name)
```

## Parse GSDB.list.of.lists$GO.biological.processes.v6.2
```{r}
library(dplyr)
GO.df <- stack(GSDB.list.of.lists$GO.biological.processes.v6.2) %>%
  rename(gene = values, pathway = ind)

Interaction.Pathways.Map <- CellChatDB.Genes.Pathways %>%
  inner_join(GO.df, by="gene", relationship = "many-to-many")

Significant.LR.Interactions <- subset(LR.Top.Table, minus.log.q.val > 0.60)

CellChat.GO.Aggregation <- Interaction.Pathways.Map %>%
  inner_join(Significant.LR.Interactions, by = c("interaction_name" = "Gene"), relationship = "many-to-many") %>%
  group_by(pathway) %>%
  summarize(
    n_interactions = n_distinct(interaction_name),
    sum_logFC = sum(logFC, na.rm = TRUE),
    genes_involved = paste(unique(gene), collapse = ", "),
    .groups = 'drop'
  ) %>%
  arrange(desc(sum_logFC))

my.save.RData("CellChat.GO.Aggregation", file.path("~/Thesis/CellChat/outputs/dCCC"), is.load.obj = T)
```



# Differential CCC - V3 (POC)
```{r}
V3.Pipeline.Config <- list(
  #Seurat Parsing Config
  n.var.genes = 5000
)
DefaultAssay(HNSCC.Atlas) <- "RNA"

HNSCC.Atlas.Malignant.Seurat <- subset(HNSCC.Atlas, Cell_Type == "Epithelial")
HNSCC.Atlas.Malignant.Seurat <- FindVariableFeatures(HNSCC.Atlas.Malignant.Seurat,
                                                     nfeatures = V3.Pipeline.Config$n.var.genes)
HNSCC.Atlas.Malignant.Top.5000.Genes <- VariableFeatures(HNSCC.Atlas.Malignant.Seurat)


my.save.RData("Bassez.2021.Seurat", myfolder=file.path(DROPBOX_DIR_PATH, "Cancer_DB/SC.3CA.Tirosh/Breast"), is.load.obj = T)
Bassez.2021.Malignant.Seurat <- subset(Bassez.2021.Seurat, cell_type == "Malignant")
Bassez.2021.Malignant.Seurat <- FindVariableFeatures(Bassez.2021.Malignant.Seurat,
                                                     nfeatures = V3.Pipeline.Config$n.var.genes)
Bassez.2021.Malignant.Top.5000.Genes <- VariableFeatures(Bassez.2021.Malignant.Seurat)
```

