---
title: "scDiffCom-PostAnalysis"
output: html_document
---

# Analysis Globals & Helpers
```{r setup, message=FALSE, warning=FALSE}
# source("~/Thesis/CCC/scripts/utils/Utils.R")
# source("~/Thesis/CCC/scripts/utils/my.utils.R")
# install.packages("SuperRanker")
source("~/Thesis/CCC/scripts/utils/Utils.R")

library(tidyverse) # Loads dplyr, tidyr, ggplot2, purrr, stringr, etc.
library(RColorBrewer)
library(pheatmap)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(scDiffCom)
library(ggplot2)
#library(ggrepel)
library(SuperRanker)


# BASE_RESULTS_DIR <- "~/Thesis/CCC/outputs/scDiffCom/scDiffComs/Consensus_Cell_Type"
BASE_RESULTS_DIR <- "~/CCC-scDiffCom/results/split-by-rank-genes"
# MALIGNANT_CELLTYPE <- c("Epithelial", "Malignant", "Tumor", "Cancer")
MALIGNANT_CELLTYPE <- c("Tumor")
USE_COSINE         <- TRUE           # TRUE → cosine distance instead of correlation
MIN_GENES_PER_CCI  <- 2              # CCIs must appear in ≥ N genes (filters sparse rows)
# OUTPUT_DIR         <- "~/Thesis/CCC/outputs/scDiffCom/plots/v2"            # where to save output PNGs (v2 — original, no unknown-type filter)
OUTPUT_DIR         <- "~/Thesis/CCC/outputs/scDiffCom/plots/v5-split-by-rank"
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

LRI_GO <- scDiffCom::LRI_human$LRI_curated_GO
BP_GO_INFO <- scDiffCom::gene_ontology_level %>%
  filter(ASPECT == "biological_process") %>%
  select(ID, NAME)
LRI_GO_BP <- LRI_GO %>%
  inner_join(BP_GO_INFO, by = c("GO_ID" = "ID"))
```


# Analysis Utils
```{r utils}
load.dataset.scDiffComs <- function(dataset_name,
                                    results_dir = BASE_RESULTS_DIR,
                                    do_switch   = FALSE) {
  message("Loading scDiffComs results for dataset: ", dataset_name)
  rds_dir   <- file.path(results_dir, dataset_name)
  rds_files <- list.files(
    rds_dir,
    pattern   = paste0("_", dataset_name, "_scDiffCom\\.rds$"),
    full.names = TRUE
  )

  scDiffCom_list <- list()

  for (rds_path in rds_files) {
    fname        <- basename(rds_path)
    gene         <- str_remove(fname, paste0("_", dataset_name, "_scDiffCom\\.rds$"))
    message("  Loading gene: ", gene)
    scDiffCom_obj <- readRDS(rds_path)
    scDiffCom_list[[gene]] <- scDiffCom_obj
  }

  return(scDiffCom_list)
}

filter.scDiffCom.cci_table_detected.for.malignant <- function(scDiffCom_obj,
                                                               malignant_celltype = MALIGNANT_CELLTYPE) {
  scDiffCom_obj@cci_table_detected %>%
    filter(
      (EMITTER_CELLTYPE %in% malignant_celltype | RECEIVER_CELLTYPE %in% malignant_celltype),
      IS_CCI_DE == TRUE,
      is.finite(LOGFC)
    )
}

filter.scDiffCom.cci_table_detected.for.celltypes <- function(scDiffCom_obj,
                                                               emitter_celltype,
                                                               receiver_celltype) {
  scDiffCom_obj@cci_table_detected %>%
    filter(
      EMITTER_CELLTYPE == emitter_celltype,
      RECEIVER_CELLTYPE == receiver_celltype,
      IS_CCI_DE == TRUE,
      is.finite(LOGFC)
    )
}
```


# Load Data
```{r load-data}
# _step4 datasets: conditions are correctly oriented — do_switch = FALSE (default)

# ── Breast ────────────────────────────────────────────────────────────────────
# Bassez2021_Breast_step4.scDiffComs <- load.dataset.scDiffComs("Bassez2021_Breast_step4")
# Qian2020_Breast_step4.scDiffComs   <- load.dataset.scDiffComs("Qian2020_Breast_step4")
# Wu2021_Breast_step4.scDiffComs     <- load.dataset.scDiffComs("Wu2021_Breast_step4")
# 
# Bassez2021_Breast_step4.malignant <- lapply(
#   Bassez2021_Breast_step4.scDiffComs,
#   filter.scDiffCom.cci_table_detected.for.malignant
# )
# Qian2020_Breast_step4.malignant <- lapply(
#   Qian2020_Breast_step4.scDiffComs,
#   filter.scDiffCom.cci_table_detected.for.malignant
# )
# Wu2021_Breast_step4.malignant <- lapply(
#   Wu2021_Breast_step4.scDiffComs,
#   filter.scDiffCom.cci_table_detected.for.malignant
# )
# 
# my.save.RData("Bassez2021_Breast_step4.malignant", "~/Thesis/CCC/outputs/scDiffCom/RData_objects", is.load.obj = T)
# my.save.RData("Qian2020_Breast_step4.malignant", "~/Thesis/CCC/outputs/scDiffCom/RData_objects", is.load.obj = T)
# my.save.RData("Wu2021_Breast_step4.malignant", "~/Thesis/CCC/outputs/scDiffCom/RData_objects", is.load.obj = T)


# ── Lung ──────────────────────────────────────────────────────────────────────
# Chan2021_Lung_step4.scDiffComs     <- load.dataset.scDiffComs("Chan2021_Lung_step4")
# Laughney2020_Lung_step4.scDiffComs <- load.dataset.scDiffComs("Laughney2020_Lung_step4")
# Xing2021_Lung_step4.scDiffComs     <- load.dataset.scDiffComs("Xing2021_Lung_step4")
#
# Chan2021_Lung_step4.malignant <- lapply(
#   Chan2021_Lung_step4.scDiffComs,
#   filter.scDiffCom.cci_table_detected.for.malignant
# )
# Laughney2020_Lung_step4.malignant <- lapply(
#   Laughney2020_Lung_step4.scDiffComs,
#   filter.scDiffCom.cci_table_detected.for.malignant
# )
# Xing2021_Lung_step4.malignant <- lapply(
#   Xing2021_Lung_step4.scDiffComs,
#   filter.scDiffCom.cci_table_detected.for.malignant
# )

# ── Head & Neck (HNSC) ────────────────────────────────────────────────────────
Kurten_HNSC.scDiffComs <- load.dataset.scDiffComs("Kurten_HNSC")
Puram_HNSC.scDiffComs  <- load.dataset.scDiffComs("Puram_HNSC")
Choi_HNSC.scDiffComs   <- load.dataset.scDiffComs("Choi_HNSC")
Bill_HNSC.scDiffComs   <- load.dataset.scDiffComs("Bill_HNSC")

Kurten_HNSC.malignant <- lapply(
  Kurten_HNSC.scDiffComs,
  filter.scDiffCom.cci_table_detected.for.malignant
)
Puram_HNSC.malignant <- lapply(
  Puram_HNSC.scDiffComs,
  filter.scDiffCom.cci_table_detected.for.malignant
)
Choi_HNSC.malignant <- lapply(
  Choi_HNSC.scDiffComs,
  filter.scDiffCom.cci_table_detected.for.malignant
)
Bill_HNSC.malignant <- lapply(
  Bill_HNSC.scDiffComs,
  filter.scDiffCom.cci_table_detected.for.malignant
)

# my.save.RData("Kurten_HNSC.malignant", "~/Thesis/CCC/outputs/scDiffCom/RData_objects")
# my.save.RData("Puram_HNSC.malignant",  "~/Thesis/CCC/outputs/scDiffCom/RData_objects")
# my.save.RData("Choi_HNSC.malignant",   "~/Thesis/CCC/outputs/scDiffCom/RData_objects")
# my.save.RData("Bill_HNSC.malignant",   "~/Thesis/CCC/outputs/scDiffCom/RData_objects")

```



## Filter Unknown Cell Types + Non-Common Genes
```{r filter-unknown-celltypes}
# Remove interactions where EMITTER or RECEIVER cell type matches "unknown", "other",
# or "equivocal" (case-insensitive). Original .malignant lists are preserved with
# an .orig suffix so the unfiltered data remain accessible in the session.

is_unknown_celltype <- function(ct) {
  grepl("unknown|other|equivocal", ct, ignore.case = TRUE)
}

filter_unknown_celltypes <- function(cci_df) {
  cci_df %>%
    filter(
      !is_unknown_celltype(EMITTER_CELLTYPE),
      !is_unknown_celltype(RECEIVER_CELLTYPE)
    )
}

# Preserve originals
# Bassez2021_Breast_step4.malignant.orig   <- Bassez2021_Breast_step4.malignant
# Qian2020_Breast_step4.malignant.orig     <- Qian2020_Breast_step4.malignant
# Wu2021_Breast_step4.malignant.orig       <- Wu2021_Breast_step4.malignant
# Chan2021_Lung_step4.malignant.orig       <- Chan2021_Lung_step4.malignant
# Laughney2020_Lung_step4.malignant.orig   <- Laughney2020_Lung_step4.malignant
# Xing2021_Lung_step4.malignant.orig       <- Xing2021_Lung_step4.malignant
Kurten_HNSC.malignant.orig <- Kurten_HNSC.malignant
Puram_HNSC.malignant.orig  <- Puram_HNSC.malignant
Choi_HNSC.malignant.orig   <- Choi_HNSC.malignant
Bill_HNSC.malignant.orig   <- Bill_HNSC.malignant

# Overwrite .malignant lists — all downstream chunks pick up the filtered data
Bassez2021_Breast_step4.malignant   <- lapply(Bassez2021_Breast_step4.malignant.orig,   filter_unknown_celltypes)
Qian2020_Breast_step4.malignant     <- lapply(Qian2020_Breast_step4.malignant.orig,     filter_unknown_celltypes)
Wu2021_Breast_step4.malignant       <- lapply(Wu2021_Breast_step4.malignant.orig,       filter_unknown_celltypes)
# Chan2021_Lung_step4.malignant       <- lapply(Chan2021_Lung_step4.malignant.orig,       filter_unknown_celltypes)
# Laughney2020_Lung_step4.malignant   <- lapply(Laughney2020_Lung_step4.malignant.orig,   filter_unknown_celltypes)
# Xing2021_Lung_step4.malignant       <- lapply(Xing2021_Lung_step4.malignant.orig,       filter_unknown_celltypes)
Kurten_HNSC.malignant <- lapply(Kurten_HNSC.malignant.orig, filter_unknown_celltypes)
Puram_HNSC.malignant  <- lapply(Puram_HNSC.malignant.orig,  filter_unknown_celltypes)
Choi_HNSC.malignant   <- lapply(Choi_HNSC.malignant.orig,   filter_unknown_celltypes)
Bill_HNSC.malignant   <- lapply(Bill_HNSC.malignant.orig,   filter_unknown_celltypes)

# For breast datasets - keep only genes present in all datasets, then select
# the top 100 by mean CCI count (computed once, applied identically to all objects)
Common.Breast.Malignant.Genes <- Reduce(intersect, list(
  names(Bassez2021_Breast_step4.malignant),
  names(Qian2020_Breast_step4.malignant),
  names(Wu2021_Breast_step4.malignant)
))

TOP_N_FILTER_GENES <- 100

Original.Genes.List <- unique(c(
  c("CDH1", "CDH2", "CDH11", "CTNNB1", "VWF", "COL1A1", "COL1A2", "VEGFA", "IGF1", "IGF2", "CSF1", "CSF1R"),
  c("AXL", "ERBB2", "EGFR", "HLA-A", "HLA-B", "HLA-C", "ESR1", "MKI67", "GATA3", "CTLA4")
  ))

rest <- head(intersect(setdiff(Complexes.Oncogenes.OncoKB, gene_list), Common.Breast.Malignant.Genes), TOP_N_FILTER_GENES - length(Original.Genes.List))
Top.N.Genes <- c(Original.Genes.List, rest)

message("Shared gene filter: ", length(Top.N.Genes),
        " genes selected (top by mean CCI count across all breast datasets).")

Bassez2021_Breast_step4.malignant <- Bassez2021_Breast_step4.malignant[Top.N.Genes]
Qian2020_Breast_step4.malignant   <- Qian2020_Breast_step4.malignant[Top.N.Genes]
Wu2021_Breast_step4.malignant     <- Wu2021_Breast_step4.malignant[Top.N.Genes]

# For H&N datasets - keep only genes present in all four H&N datasets, then select
# the same top 100 gene list (anchored on Original.Genes.List + oncogene supplement)
Common.HN.Malignant.Genes <- Reduce(intersect, list(
  names(Kurten_HNSC.malignant),
  names(Puram_HNSC.malignant),
  names(Choi_HNSC.malignant),
  names(Bill_HNSC.malignant)
))

rest.HN <- head(intersect(setdiff(Complexes.Oncogenes.OncoKB, gene_list), Common.HN.Malignant.Genes),
                TOP_N_FILTER_GENES - length(Original.Genes.List))
Top.N.HN.Genes <- Common.HN.Malignant.Genes

message("H&N shared gene filter: ", length(Top.N.HN.Genes),
        " genes selected (top by mean CCI count across all H&N datasets).")

Kurten_HNSC.malignant <- Kurten_HNSC.malignant[Top.N.HN.Genes]
Puram_HNSC.malignant  <- Puram_HNSC.malignant[Top.N.HN.Genes]
Choi_HNSC.malignant   <- Choi_HNSC.malignant[Top.N.HN.Genes]
Bill_HNSC.malignant   <- Bill_HNSC.malignant[Top.N.HN.Genes]

report_filter <- function(orig, filt, label) {
  n_orig <- sum(sapply(orig, nrow))
  n_filt <- sum(sapply(filt, nrow))
  message(sprintf("  %-35s  %d → %d rows (removed %d)", label, n_orig, n_filt, n_orig - n_filt))
}
message("Unknown-celltype filter applied:")
report_filter(Bassez2021_Breast_step4.malignant.orig,  Bassez2021_Breast_step4.malignant,  "Bassez2021_Breast_step4")
report_filter(Qian2020_Breast_step4.malignant.orig,    Qian2020_Breast_step4.malignant,    "Qian2020_Breast_step4")
report_filter(Wu2021_Breast_step4.malignant.orig,      Wu2021_Breast_step4.malignant,      "Wu2021_Breast_step4")
# report_filter(Chan2021_Lung_step4.malignant.orig,      Chan2021_Lung_step4.malignant,      "Chan2021_Lung_step4")
# report_filter(Laughney2020_Lung_step4.malignant.orig,  Laughney2020_Lung_step4.malignant,  "Laughney2020_Lung_step4")
# report_filter(Xing2021_Lung_step4.malignant.orig,      Xing2021_Lung_step4.malignant,      "Xing2021_Lung_step4")
report_filter(Kurten_HNSC.malignant.orig, Kurten_HNSC.malignant, "Kurten_HNSC")
report_filter(Puram_HNSC.malignant.orig,  Puram_HNSC.malignant,  "Puram_HNSC")
report_filter(Choi_HNSC.malignant.orig,   Choi_HNSC.malignant,   "Choi_HNSC")
report_filter(Bill_HNSC.malignant.orig,   Bill_HNSC.malignant,   "Bill_HNSC")
```





# Data Quality Checks
```{r celltype-noise-diagnostics}
# Uses .malignant.orig (all interactions, before unknown-celltype removal)
# to quantify how much "other/unknown/equivocal" noise each dataset contains.
OUTPUT_DIR         <- "~/Thesis/CCC/outputs/scDiffCom/plots/v2/qualityCheck"

# ── Local dataset metadata ────────────────────────────────────────────────────
noise_ds_meta <- data.frame(
  ds_full     = c("Bassez2021_Breast_step4", "Qian2020_Breast_step4",
                  "Wu2021_Breast_step4",     "Chan2021_Lung_step4",
                  "Laughney2020_Lung_step4", "Xing2021_Lung_step4",
                  "Kurten_HNSC", "Puram_HNSC", "Choi_HNSC", "Bill_HNSC"),
  ds_short    = c("Bassez2021", "Qian2020", "Wu2021",
                  "Chan2021",   "Laughney2020", "Xing2021",
                  "Kurten", "Puram", "Choi", "Bill"),
  Cancer_Type = c("Breast", "Breast", "Breast", "Lung", "Lung", "Lung",
                  "H&N", "H&N", "H&N", "H&N"),
  stringsAsFactors = FALSE
)

malignant_orig_catalog <- list(
  # Bassez2021_Breast_step4   = Bassez2021_Breast_step4.malignant.orig,
  # Qian2020_Breast_step4     = Qian2020_Breast_step4.malignant.orig,
  # Wu2021_Breast_step4       = Wu2021_Breast_step4.malignant.orig,
  # Chan2021_Lung_step4       = Chan2021_Lung_step4.malignant.orig,
  # Laughney2020_Lung_step4   = Laughney2020_Lung_step4.malignant.orig,
  # Xing2021_Lung_step4       = Xing2021_Lung_step4.malignant.orig,
  Kurten_HNSC               = Kurten_HNSC.malignant.orig,
  Puram_HNSC                = Puram_HNSC.malignant.orig,
  Choi_HNSC                 = Choi_HNSC.malignant.orig,
  Bill_HNSC                 = Bill_HNSC.malignant.orig
)

# ── Count interactions per gene per dataset ───────────────────────────────────
noise_counts_gene <- purrr::imap_dfr(malignant_orig_catalog, function(gene_list, ds_full) {
  purrr::imap_dfr(gene_list, function(df, gene) {
    if (nrow(df) == 0) return(NULL)
    n_other <- sum(is_unknown_celltype(df$EMITTER_CELLTYPE) |
                   is_unknown_celltype(df$RECEIVER_CELLTYPE))
    data.frame(Dataset = ds_full, Gene = gene,
               Clean = nrow(df) - n_other, Contains_Other = n_other,
               stringsAsFactors = FALSE)
  })
})

# ── Aggregate to dataset level ─────────────────────────────────────────────────
noise_counts <- noise_counts_gene %>%
  group_by(Dataset) %>%
  summarise(Clean = sum(Clean), Contains_Other = sum(Contains_Other), .groups = "drop")

# ── Summary table ─────────────────────────────────────────────────────────────
table_celltype_noise_stats <- noise_counts %>%
  left_join(noise_ds_meta, by = c("Dataset" = "ds_full")) %>%
  mutate(
    Total_Interactions       = Clean + Contains_Other,
    Other_Interactions_Count = Contains_Other,
    Clean_Interactions_Count = Clean,
    Percent_Noise            = round(100 * Contains_Other / Total_Interactions, 2)
  ) %>%
  select(Dataset, Cancer_Type, Total_Interactions,
         Other_Interactions_Count, Clean_Interactions_Count, Percent_Noise) %>%
  arrange(Cancer_Type, Dataset)

print(table_celltype_noise_stats)
View(table_celltype_noise_stats)

# ── Per-gene summary table ─────────────────────────────────────────────────────
table_celltype_noise_gene_stats <- noise_counts_gene %>%
  left_join(noise_ds_meta, by = c("Dataset" = "ds_full")) %>%
  mutate(
    Total_Interactions       = Clean + Contains_Other,
    Other_Interactions_Count = Contains_Other,
    Clean_Interactions_Count = Clean,
    Percent_Noise            = round(100 * Contains_Other / Total_Interactions, 2)
  ) %>%
  select(Dataset, Cancer_Type, Gene, Total_Interactions,
         Other_Interactions_Count, Clean_Interactions_Count, Percent_Noise) %>%
  arrange(Cancer_Type, Dataset, Gene)

print(table_celltype_noise_gene_stats)
View(table_celltype_noise_gene_stats)

# ── Long format for plotting ───────────────────────────────────────────────────
noise_long <- noise_counts %>%
  left_join(noise_ds_meta, by = c("Dataset" = "ds_full")) %>%
  pivot_longer(c(Clean, Contains_Other),
               names_to = "Category", values_to = "Count") %>%
  mutate(
    Category = factor(
      dplyr::recode(Category, Contains_Other = "Contains Other"),
      levels = c("Clean", "Contains Other")
    ),
    ds_short = factor(ds_short, levels = noise_ds_meta$ds_short)
  )

# ── Plot aesthetics ────────────────────────────────────────────────────────────
fill_palette <- c("Clean" = "#4DAF4A", "Contains Other" = "#E41A1C")
# X-axis label colours follow cancer type: Breast = red, Lung = blue, H&N = darkgreen
x_label_cols <- dplyr::case_when(
  noise_ds_meta$Cancer_Type == "Breast" ~ "#E41A1C",
  noise_ds_meta$Cancer_Type == "Lung"   ~ "#377EB8",
  noise_ds_meta$Cancer_Type == "H&N"    ~ "#33A02C",
  TRUE ~ "black"
)

# ── Plot 1: Absolute counts ────────────────────────────────────────────────────
p_noise_counts <- ggplot(noise_long,
                         aes(x = ds_short, y = Count, fill = Category)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(values = fill_palette) +
  scale_y_continuous(labels = scales::comma) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, colour = x_label_cols),
    legend.position = "right"
  ) +
  labs(
    title    = "Celltype Noise in Malignant Interaction Pool",
    subtitle = "X-axis labels: red = Breast, blue = Lung, green = H&N",
    x = NULL, y = "Number of Interactions", fill = "Category"
  )

print(p_noise_counts)
ggsave(file.path(OUTPUT_DIR, "celltype_noise_counts.png"),
       plot = p_noise_counts, width = 9, height = 6, dpi = 300)
message("Saved celltype_noise_counts.png")

# ── Plot 2: Proportions ────────────────────────────────────────────────────────
p_noise_props <- ggplot(noise_long,
                        aes(x = ds_short, y = Count, fill = Category)) +
  geom_col(position = "fill", width = 0.7) +
  scale_fill_manual(values = fill_palette) +
  scale_y_continuous(labels = scales::percent) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, colour = x_label_cols),
    legend.position = "right"
  ) +
  labs(
    title    = "Proportion of Celltype Noise in Malignant Interaction Pool",
    subtitle = "X-axis labels: red = Breast, blue = Lung, green = H&N",
    x = NULL, y = "Proportion of Interactions", fill = "Category"
  )

print(p_noise_props)
ggsave(file.path(OUTPUT_DIR, "celltype_noise_proportions.png"),
       plot = p_noise_props, width = 9, height = 6, dpi = 300)
message("Saved celltype_noise_proportions.png")

# ── Plot 3: Per-gene % noise heatmap (Gene × Dataset) ─────────────────────────
noise_gene_tile <- noise_counts_gene %>%
  left_join(noise_ds_meta, by = c("Dataset" = "ds_full")) %>%
  mutate(
    Percent_Noise = 100 * Contains_Other / (Clean + Contains_Other),
    ds_short      = factor(ds_short, levels = noise_ds_meta$ds_short)
  )

n_genes_tile <- length(unique(noise_gene_tile$Gene))

p_gene_heatmap <- ggplot(noise_gene_tile,
                         aes(x = ds_short, y = Gene, fill = Percent_Noise)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "#4DAF4A", mid = "#ffffbf", high = "#E41A1C",
    midpoint = 50, limits = c(0, 100), name = "% Noise"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = x_label_cols),
    axis.text.y = element_text(size = max(4, 180 / n_genes_tile))
  ) +
  labs(
    title    = "Celltype Noise per Gene per Dataset",
    subtitle = "Color = % interactions containing 'other/unknown/equivocal' cell type\nX-axis labels: red = Breast, blue = Lung, green = H&N",
    x = NULL, y = "Gene"
  )

print(p_gene_heatmap)
ggsave(file.path(OUTPUT_DIR, "celltype_noise_gene_heatmap.png"),
       plot = p_gene_heatmap,
       width  = 9,
       height = max(6, n_genes_tile * 0.25),
       dpi    = 300)
message("Saved celltype_noise_gene_heatmap.png")
```


## Quality Check — Cell-Type Interaction Distributions
```{r qc-celltype-distributions, message=FALSE, warning=FALSE}
library(ggalluvial)

# ── 1. Data Preparation ───────────────────────────────────────────────────────
# Flatten malignant_orig_catalog (list of datasets → list of genes → data frames)
# into one tidy data frame: Dataset | Gene | EMITTER | RECEIVER
qc_interactions <- purrr::imap_dfr(malignant_orig_catalog, function(gene_list, ds_full) {
  ds_short <- noise_ds_meta$ds_short[noise_ds_meta$ds_full == ds_full]
  purrr::imap_dfr(gene_list, function(df, gene) {
    if (is.null(df) || nrow(df) == 0) return(NULL)
    data.frame(
      Dataset  = ds_short,
      Gene     = gene,
      EMITTER  = df$EMITTER_CELLTYPE,
      RECEIVER = df$RECEIVER_CELLTYPE,
      stringsAsFactors = FALSE
    )
  })
})

# Dataset factor ordered as in noise_ds_meta
qc_interactions$Dataset <- factor(qc_interactions$Dataset,
                                  levels = noise_ds_meta$ds_short)

# Aggregate: count interactions per Dataset × EMITTER × RECEIVER
qc_counts <- qc_interactions %>%
  count(Dataset, EMITTER, RECEIVER, name = "n_interactions")

# Consistent cell-type order (most active first) for both axes
ct_order <- qc_interactions %>%
  count(EMITTER, name = "n") %>%
  arrange(desc(n)) %>%
  pull(EMITTER) %>%
  unique()

qc_counts <- qc_counts %>%
  mutate(
    EMITTER  = factor(EMITTER,  levels = rev(ct_order)),  # y-axis: top = most active
    RECEIVER = factor(RECEIVER, levels = ct_order)
  )

message("QC data prepared: ", nrow(qc_counts),
        " Dataset × EMITTER × RECEIVER combinations")

# ── 2. Faceted Heatmap ────────────────────────────────────────────────────────
p_qc_heatmap <- ggplot(qc_counts,
                       aes(x = RECEIVER, y = EMITTER, fill = n_interactions)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#ccd9ff", high = "#08306b",
                      name = "# interactions",
                      labels = scales::comma) +
  facet_wrap(~ Dataset, ncol = 3) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y      = element_text(size = 7),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    strip.text       = element_text(face = "bold"),
    panel.spacing    = unit(0.5, "lines")
  ) +
  labs(
    title    = "Cell-Type Interaction Counts per Dataset (unfiltered)",
    subtitle = "'Other/Unknown/Equivocal' cell types included — rows = Emitter, cols = Receiver",
    x = "Receiver Cell Type", y = "Emitter Cell Type"
  )

print(p_qc_heatmap)
ggsave(file.path(OUTPUT_DIR, "qc_celltype_heatmap.png"),
       plot = p_qc_heatmap, width = 18, height = 10, dpi = 300)
message("Saved qc_celltype_heatmap.png")

# ── 3. Sankey / Alluvial Diagram (aggregated across all datasets) ─────────────
TOP_PAIRS_SANKEY <- 20   # keep top N Emitter→Receiver pairs

qc_alluvial_data <- qc_interactions %>%
  count(EMITTER, RECEIVER, name = "n_interactions") %>%
  arrange(desc(n_interactions)) %>%
  slice_head(n = TOP_PAIRS_SANKEY)

# sqrt(n) transform: compresses Epithelial's dominant band while keeping
# smaller categories (Hematopoietic, Myeloid, …) visually distinct.
p_qc_sankey <- ggplot(
  qc_alluvial_data,
  aes(axis1 = EMITTER, axis2 = RECEIVER, y = sqrt(n_interactions))
) +
  # Sigmoid curves look cleaner than the default splines
  geom_alluvium(aes(fill = EMITTER), width = 1/10, alpha = 0.65,
                curve_type = "sigmoid") +
  geom_stratum(width = 1/4, fill = "grey92", colour = "grey55",
               linewidth = 0.4) +
  # hjust = "outward" floats every label outside its stratum box so labels
  # for small categories don't collide with the flows or each other.
  geom_text(
    stat     = "stratum",
    aes(label = after_stat(stratum)),
    hjust    = "outward",
    size     = 3.2,
    fontface = "bold"
  ) +
  scale_x_discrete(
    limits = c("Emitter", "Receiver"),
    expand = c(0.35, 0.35)   # extra horizontal room for outward labels
  ) +
  theme_void(base_size = 13) +
  theme(
    axis.text.x     = element_text(face = "bold", size = 13,
                                   margin = margin(b = 6)),
    plot.title      = element_text(face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(size = 10, colour = "grey40", hjust = 0.5),
    plot.margin     = margin(12, 40, 12, 40),
    legend.position = "none"
  ) +
  labs(
    title    = sprintf(
      "Cell-Type Interaction Flow — Top %d Emitter→Receiver Pairs (All Datasets)",
      TOP_PAIRS_SANKEY)  )

print(p_qc_sankey)
ggsave(file.path(OUTPUT_DIR, "qc_celltype_sankey.png"),
       plot = p_qc_sankey, width = 14, height = 9, dpi = 300)
message("Saved qc_celltype_sankey.png")

# ── 4. Top-10 Summary Table ───────────────────────────────────────────────────
qc_top10 <- qc_interactions %>%
  mutate(ER_CELLTYPES = paste0(EMITTER, "_", RECEIVER)) %>%
  count(ER_CELLTYPES, EMITTER, RECEIVER, name = "n_interactions") %>%
  arrange(desc(n_interactions)) %>%
  slice_head(n = 10)

message("\nTop 10 most frequent Emitter→Receiver pairs (across all datasets):")
print(qc_top10)
View(qc_top10)
```








# Shared Config, Helpers & CCI Set Construction
```{r cci-sets}
# ── GLOBAL CONFIG ─────────────────────────────────────────────────────────────
TOP_N_CCI    <- 500        # top CCIs per gene per dataset (ranked by |LOGFC|) — used by Jaccard
GENE_OF_INT  <- "AXL"   # gene for the 6×6 case-study heatmap
SEED         <- 42

DATASET_LABELS <- c(
  "Bassez2021_Breast_step4",
  "Qian2020_Breast_step4",
  "Wu2021_Breast_step4",
  "Chan2021_Lung_step4",
  "Laughney2020_Lung_step4",
  "Xing2021_Lung_step4",
  "Kurten_HNSC",
  "Puram_HNSC",
  "Choi_HNSC",
  "Bill_HNSC"
)

# ── DATASET METADATA MAPS ─────────────────────────────────────────────────────
# Short display names (used as point labels / row IDs in UMAP)
ds_short_map <- c(
  Bassez2021_Breast_step4   = "Bassez2021",
  Qian2020_Breast_step4     = "Qian2020",
  Wu2021_Breast_step4       = "Wu2021",
  Chan2021_Lung_step4       = "Chan2021",
  Laughney2020_Lung_step4   = "Laughney2020",
  Xing2021_Lung_step4       = "Xing2021",
  Kurten_HNSC               = "Kurten",
  Puram_HNSC                = "Puram",
  Choi_HNSC                 = "Choi",
  Bill_HNSC                 = "Bill"
)

# Cancer type per dataset (used for UMAP shape aesthetic)
ds_cancer_map <- c(
  Bassez2021_Breast_step4   = "Breast",
  Qian2020_Breast_step4     = "Breast",
  Wu2021_Breast_step4       = "Breast",
  Chan2021_Lung_step4       = "Lung",
  Laughney2020_Lung_step4   = "Lung",
  Xing2021_Lung_step4       = "Lung",
  Kurten_HNSC               = "H&N",
  Puram_HNSC                = "H&N",
  Choi_HNSC                 = "H&N",
  Bill_HNSC                 = "H&N"
)

# ── PLOT AESTHETICS ───────────────────────────────────────────────────────────
# Colours per source dataset (UMAP) and shapes per cancer type
source_colors <- c(
  Bassez2021   = "#E41A1C",
  Qian2020     = "#FF7F00",
  Wu2021       = "#4DAF4A",
  Chan2021     = "#377EB8",
  Laughney2020 = "#984EA3",
  Xing2021     = "#A65628",
  Kurten       = "#F781BF",
  Puram        = "#999999",
  Choi         = "#66C2A5",
  Bill         = "#FC8D62"
)
cancer_shapes <- c(Breast = 16, Lung = 17, `H&N` = 15)  # circle, triangle, square

# ── HELPER FUNCTIONS ──────────────────────────────────────────────────────────

# build_gene_cci_sets: for each gene in a .malignant list, return CCIs ranked by
# |LOGFC|. top_n = NULL returns all interactions (used by Average Overlap); an
# integer truncates to the top-N (used by Jaccard).
build_gene_cci_sets <- function(malignant_list, top_n = TOP_N_CCI) {
  map(malignant_list, function(df) {
    df <- df %>%
      as.data.frame() %>%
      arrange(-abs(LOGFC))
    if (!is.null(top_n)) df <- slice_head(df, n = top_n)
    pull(df, CCI) %>% unique()
  })
}

build_cci_logfc_mat <- function(malignant_list, min_genes = MIN_GENES_PER_CCI) {
  cci_list <- lapply(names(malignant_list), function(gene) {
    malignant_list[[gene]] %>%
      select(CCI, LOGFC) %>%
      group_by(CCI) %>%
      summarise(LOGFC = mean(LOGFC, na.rm = TRUE), .groups = "drop") %>%
      mutate(Gene = gene)
  })

  all_data <- bind_rows(cci_list)
  mat_long <- all_data %>%
    pivot_wider(names_from = Gene, values_from = LOGFC)

  mat <- as.matrix(mat_long[, -1])
  rownames(mat) <- mat_long$CCI
  message("  Matrix dimensions (CCIs x Genes): ", nrow(mat), " x ", ncol(mat))

  n_non_na <- rowSums(!is.na(mat))
  mat <- mat[n_non_na >= min_genes, , drop = FALSE]
  message("  After filtering sparse CCIs: ", nrow(mat), " x ", ncol(mat))
  mat
}

# Build Gene×Dataset × CCI logFC matrix (NA where CCI absent for that gene×dataset).
build_gene_dataset_cci_feat_mat <- function(cci_mats, cancer_map) {
  all_ccis <- sort(unique(unlist(lapply(cci_mats, rownames))))
  gd_rows <- purrr::imap(cci_mats, function(mat, ds) {
    gene_names <- colnames(mat)
    lapply(gene_names, function(g) {
      vec <- setNames(rep(NA_real_, length(all_ccis)), all_ccis)
      present <- rownames(mat)[!is.na(mat[, g])]
      vec[present] <- mat[present, g]
      list(
        label   = paste0(g, "_", ds),
        gene    = g,
        dataset = ds,
        cancer  = cancer_map[[ds]],
        vec     = vec
      )
    })
  }) %>% purrr::flatten()

  non_empty <- sapply(gd_rows, function(x) any(!is.na(x$vec)))
  gd_rows   <- gd_rows[non_empty]

  feat_mat <- do.call(rbind, lapply(gd_rows, `[[`, "vec"))
  rownames(feat_mat) <- vapply(gd_rows, `[[`, "", "label")

  list(
    feat_mat    = feat_mat,
    all_ccis    = all_ccis,
    labels_gd   = vapply(gd_rows, `[[`, "", "label"),
    genes_gd    = vapply(gd_rows, `[[`, "", "gene"),
    datasets_gd = vapply(gd_rows, `[[`, "", "dataset"),
    cancer_gd   = vapply(gd_rows, `[[`, "", "cancer")
  )
}

# HVG CCIs by cross Gene×Dataset logFC variance (matches post-analysis-hvg-ccis).
select_hvg_ccis <- function(malignant_by_ds, cancer_map,
                            top_pct = 0.05, min_obs = 10) {
  cci_mats <- lapply(malignant_by_ds, build_cci_logfc_mat)
  feat_mat <- build_gene_dataset_cci_feat_mat(cci_mats, cancer_map)$feat_mat
  n_obs_per_cci <- colSums(!is.na(feat_mat))
  cci_var <- apply(feat_mat, 2, function(x) var(x, na.rm = TRUE))
  cci_var[n_obs_per_cci < min_obs] <- NA_real_
  eligible <- !is.na(cci_var)
  if (!any(eligible)) {
    stop("No CCIs passed min_obs filter for HVG selection.")
  }
  var_cutoff <- quantile(cci_var[eligible], probs = 1 - top_pct, na.rm = TRUE)
  hvg <- names(cci_var)[eligible & cci_var >= var_cutoff]
  list(
    hvg         = hvg,
    n_eligible  = sum(eligible),
    var_cutoff  = var_cutoff,
    cci_var     = cci_var
  )
}

restrict_ccis_to_hvg <- function(ccis, hvg_ccis) {
  ccis[ccis %in% hvg_ccis]
}

# jaccard_dist: scalar Jaccard distance between two character-vector CCI sets.
jaccard_dist <- function(set_a, set_b) {
  inter <- length(intersect(set_a, set_b))
  union <- length(union(set_a, set_b))
  if (union == 0) return(NA_real_)
  1 - inter / union
}

# build_gene_dataset_jaccard: N×N Jaccard distance matrix + pairwise CCI intersections
# for one gene across N datasets.
# Returns a list:
#   $dist_mat      — N×N numeric Jaccard distance matrix
#   $intersections — N×N list-matrix of shared CCI character vectors (diagonal = own CCIs)
build_gene_dataset_jaccard <- function(gene, all_sets, labels = DATASET_LABELS) {
  sets    <- map(all_sets, function(ds_sets) ds_sets[[gene]])
  present <- map_lgl(sets, Negate(is.null))
  if (!all(present)) {
    warning(sprintf("Gene '%s' absent in: %s",
                    gene, paste(labels[!present], collapse = ", ")))
  }
  n            <- length(labels)
  dist_mat     <- matrix(NA_real_, nrow = n, ncol = n, dimnames = list(labels, labels))
  intersections <- matrix(vector("list", n * n), nrow = n, ncol = n,
                          dimnames = list(labels, labels))
  for (i in seq_len(n)) for (j in seq_len(n)) {
    if (i == j) {
      dist_mat[i, j]      <- NA_real_
      intersections[[i, j]] <- sets[[i]]
    } else if (!is.null(sets[[i]]) && !is.null(sets[[j]])) {
      dist_mat[i, j]      <- jaccard_dist(sets[[i]], sets[[j]])
      intersections[[i, j]] <- intersect(sets[[i]], sets[[j]])
    }
  }
  list(dist_mat = dist_mat, intersections = intersections)
}


# pad_list / dist_fun_superranker_overlap: Average Overlap distance helpers.
# pad_list fills a ranked list to target_len with unique placeholder tokens so
# that SuperRanker::average_overlap receives two same-length rank matrices.
pad_list <- function(lst, target_len) {
  if (length(lst) < target_len) {
    padding <- paste0("MISSING_", seq_len(target_len - length(lst)), "__")
    return(c(lst, padding))
  }
  return(lst[1:target_len])
}

dist_fun_superranker_overlap <- function(list1, list2, k = NULL) {
  if (is.null(k)) k <- max(length(list1), length(list2))
  list1     <- pad_list(list1, k)
  list2     <- pad_list(list2, k)
  all_items <- unique(c(list1, list2))
  ranks1    <- match(all_items, list1);  ranks1[is.na(ranks1)] <- k + 1
  ranks2    <- match(all_items, list2);  ranks2[is.na(ranks2)] <- k + 1
  rank_matrix    <- cbind(ranks1, ranks2)
  overlap_values <- SuperRanker::average_overlap(rank_matrix)
  similarity     <- mean(overlap_values[1:min(k, length(overlap_values))], na.rm = TRUE)
  1 - similarity
}

# build_gene_dataset_ao: N×N Average Overlap distance matrix for one gene across
# N datasets. Uses full ranked CCI lists (no top-N truncation).
build_gene_dataset_ao <- function(gene, all_ranked_sets, labels = DATASET_LABELS) {
  sets    <- map(all_ranked_sets, function(ds_sets) ds_sets[[gene]])
  present <- map_lgl(sets, Negate(is.null))
  if (!all(present)) {
    warning(sprintf("Gene '%s' absent in: %s",
                    gene, paste(labels[!present], collapse = ", ")))
  }
  n        <- length(labels)
  dist_mat <- matrix(NA_real_, nrow = n, ncol = n, dimnames = list(labels, labels))
  for (i in seq_len(n)) for (j in seq_len(n)) {
    if (i == j) {
      dist_mat[i, j] <- NA_real_
    } else if (!is.null(sets[[i]]) && !is.null(sets[[j]])) {
      dist_mat[i, j] <- dist_fun_superranker_overlap(sets[[i]], sets[[j]])
    }
  }
  dist_mat
}

# ── BUILD PER-DATASET CCI SETS ────────────────────────────────────────────────
# Bassez_top_cci_sets   <- build_gene_cci_sets(Bassez2021_Breast_step4.malignant,   top_n = TOP_N_CCI)
# Qian_top_cci_sets     <- build_gene_cci_sets(Qian2020_Breast_step4.malignant,     top_n = TOP_N_CCI)
# Wu_top_cci_sets       <- build_gene_cci_sets(Wu2021_Breast_step4.malignant,       top_n = TOP_N_CCI)
# Chan_top_cci_sets     <- build_gene_cci_sets(Chan2021_Lung_step4.malignant,       top_n = TOP_N_CCI)
# Laughney_top_cci_sets <- build_gene_cci_sets(Laughney2020_Lung_step4.malignant,   top_n = TOP_N_CCI)
# Xing_top_cci_sets     <- build_gene_cci_sets(Xing2021_Lung_step4.malignant,       top_n = TOP_N_CCI)
Kurten_top_cci_sets   <- build_gene_cci_sets(Kurten_HNSC.malignant,               top_n = TOP_N_CCI)
Puram_top_cci_sets    <- build_gene_cci_sets(Puram_HNSC.malignant,                top_n = TOP_N_CCI)
Choi_top_cci_sets     <- build_gene_cci_sets(Choi_HNSC.malignant,                 top_n = TOP_N_CCI)
Bill_top_cci_sets     <- build_gene_cci_sets(Bill_HNSC.malignant,                 top_n = TOP_N_CCI)

# message("Genes in Bassez:   ", length(Bassez_top_cci_sets))
# message("Genes in Qian:     ", length(Qian_top_cci_sets))
# message("Genes in Wu:       ", length(Wu_top_cci_sets))
# message("Genes in Chan:     ", length(Chan_top_cci_sets))
# message("Genes in Laughney: ", length(Laughney_top_cci_sets))
# message("Genes in Xing:     ", length(Xing_top_cci_sets))
message("Genes in Kurten:   ", length(Kurten_top_cci_sets))
message("Genes in Puram:    ", length(Puram_top_cci_sets))
message("Genes in Choi:     ", length(Choi_top_cci_sets))
message("Genes in Bill:     ", length(Bill_top_cci_sets))

# ── COMBINED CCI SET LIST (used by all downstream chunks) ─────────────────────
all_top_cci_sets <- list(
  # Bassez2021_Breast_step4   = Bassez_top_cci_sets,
  # Qian2020_Breast_step4     = Qian_top_cci_sets,
  # Wu2021_Breast_step4       = Wu_top_cci_sets,
  # Chan2021_Lung_step4       = Chan_top_cci_sets,
  # Laughney2020_Lung_step4   = Laughney_top_cci_sets,
  # Xing2021_Lung_step4       = Xing_top_cci_sets,
  Kurten_HNSC               = Kurten_top_cci_sets,
  Puram_HNSC                = Puram_top_cci_sets,
  Choi_HNSC                 = Choi_top_cci_sets,
  Bill_HNSC                 = Bill_top_cci_sets
)

# ── RANKED CCI SETS (full list, no truncation — used by Average Overlap) ─────
# Bassez_cci_sets_ranked   <- build_gene_cci_sets(Bassez2021_Breast_step4.malignant,   top_n = NULL)
# Qian_cci_sets_ranked     <- build_gene_cci_sets(Qian2020_Breast_step4.malignant,     top_n = NULL)
# Wu_cci_sets_ranked       <- build_gene_cci_sets(Wu2021_Breast_step4.malignant,       top_n = NULL)
# Chan_cci_sets_ranked     <- build_gene_cci_sets(Chan2021_Lung_step4.malignant,       top_n = NULL)
# Laughney_cci_sets_ranked <- build_gene_cci_sets(Laughney2020_Lung_step4.malignant,   top_n = NULL)
# Xing_cci_sets_ranked     <- build_gene_cci_sets(Xing2021_Lung_step4.malignant,       top_n = NULL)
Kurten_cci_sets_ranked   <- build_gene_cci_sets(Kurten_HNSC.malignant,               top_n = NULL)
Puram_cci_sets_ranked    <- build_gene_cci_sets(Puram_HNSC.malignant,                top_n = NULL)
Choi_cci_sets_ranked     <- build_gene_cci_sets(Choi_HNSC.malignant,                 top_n = NULL)
Bill_cci_sets_ranked     <- build_gene_cci_sets(Bill_HNSC.malignant,                 top_n = NULL)

all_cci_sets_ranked <- list(
  Bassez2021_Breast_step4   = Bassez_cci_sets_ranked,
  Qian2020_Breast_step4     = Qian_cci_sets_ranked,
  Wu2021_Breast_step4       = Wu_cci_sets_ranked,
  Chan2021_Lung_step4       = Chan_cci_sets_ranked,
  Laughney2020_Lung_step4   = Laughney_cci_sets_ranked,
  Xing2021_Lung_step4       = Xing_cci_sets_ranked,
  Kurten_HNSC               = Kurten_cci_sets_ranked,
  Puram_HNSC                = Puram_cci_sets_ranked,
  Choi_HNSC                 = Choi_cci_sets_ranked,
  Bill_HNSC                 = Bill_cci_sets_ranked
)

# ── GENE-CCI INSPECTION TABLE ─────────────────────────────────────────────────
# One tidy data frame per dataset: Gene | n_CCIs | CCI (one row per gene-CCI pair).
# Usage: View(gene_cci_top500_summary$Bassez2021_Breast_step4)

gene_cci_top500_summary <- purrr::imap(all_top_cci_sets, function(ds_sets, ds_name) {
  
  # Step 1: Create the un-arranged data frame
  df <- purrr::imap_dfr(ds_sets, function(cci_vec, gene) {
    if (length(cci_vec) == 0) return(NULL)
    
    data.frame(
      Gene = gene, 
      n_CCIs = length(cci_vec), 
      CCI = cci_vec,
      stringsAsFactors = FALSE
    )
  })
  
  # Step 2: Check if the resulting data frame is empty
  # If it is, return an empty data frame with the correct columns to prevent errors
  if (nrow(df) == 0 || !("Gene" %in% names(df))) {
    return(data.frame(
      Gene = character(), 
      n_CCIs = integer(), 
      CCI = character(), 
      stringsAsFactors = FALSE
    ))
  }
  
  # Step 3: If not empty, arrange and return
  df %>% arrange(.data$Gene, .data$CCI)
})

message("gene_cci_top500_summary ready — datasets: ",
        paste(names(gene_cci_top500_summary), collapse = ", "))

# ── Quick inspection (uncomment to run interactively) ─────────────────────────
# View(gene_cci_top500_summary$Bassez2021_Breast_step4)
# View(gene_cci_top500_summary$Qian2020_Breast_step4)
# View(gene_cci_top500_summary$Wu2021_Breast_step4)
# View(gene_cci_top500_summary$Chan2021_Lung_step4)
# View(gene_cci_top500_summary$Laughney2020_Lung_step4)
# View(gene_cci_top500_summary$Xing2021_Lung_step4)
```





# Gene-of-Interest — 6×6 Jaccard Heatmap Across All Datasets
```{r gene-gene-jaccard}
OUTPUT_DIR <- "~/Thesis/CCC/outputs/scDiffCom/plots/v3"
GENE_OF_INT <- "AXL"
goi_result        <- build_gene_dataset_jaccard(GENE_OF_INT, all_top_cci_sets)
goi_jaccard_mat   <- goi_result$dist_mat
goi_intersections <- goi_result$intersections   # list-matrix: [i,j] → shared CCI strings

message(sprintf("\n%s — Jaccard distance matrix (top %d CCIs by |LOGFC|):",
                GENE_OF_INT, TOP_N_CCI))
print(round(goi_jaccard_mat, 3))

# NA cells arise when a gene is absent in a dataset; display as "N/A", colour as 1.
# Diagonal is NA (self-distance undefined); off-diagonal NAs → 1 for colour scale.
goi_display_mat <- goi_jaccard_mat
goi_display_mat[is.na(goi_display_mat) & row(goi_display_mat) != col(goi_display_mat)] <- 1

goi_clust_mat <- goi_jaccard_mat
goi_clust_mat[is.na(goi_clust_mat)] <- 1
diag(goi_clust_mat) <- 0

cell_annot <- matrix(
  ifelse(row(goi_jaccard_mat) == col(goi_jaccard_mat), "",
         ifelse(is.na(goi_jaccard_mat), "N/A", sprintf("%.3f", goi_jaccard_mat))),
  nrow = nrow(goi_jaccard_mat),
  dimnames = dimnames(goi_jaccard_mat)
)

goi_heatmap_file <- file.path(OUTPUT_DIR, paste0(GENE_OF_INT, "_jaccard_umap.png"))
png(goi_heatmap_file, width = 5000, height = 3000, res = 300)

# Dynamic color bounds for the plotted matrix
goi_min_val <- min(goi_display_mat, na.rm = TRUE)
goi_max_val <- max(goi_display_mat, na.rm = TRUE)
goi_breaks  <- seq(goi_min_val, goi_max_val, length.out = 101)
goi_leg_breaks <- pretty(c(goi_min_val, goi_max_val), n = 5)
goi_leg_breaks <- goi_leg_breaks[goi_leg_breaks >= goi_min_val & goi_leg_breaks <= goi_max_val]
goi_leg_labels <- format(round(goi_leg_breaks, 3), trim = TRUE)

pheatmap(
  goi_display_mat,
  color             = colorRampPalette(c("#4575b4", "#ffffbf", "#d73027"))(100),
  breaks            = goi_breaks,
  na_col            = "grey80",
  cluster_rows      = TRUE,
  cluster_cols      = TRUE,
  clustering_distance_rows = as.dist(goi_clust_mat),
  clustering_distance_cols = as.dist(goi_clust_mat),
  display_numbers   = cell_annot,
  number_color      = "black",
  fontsize_number   = 10,
  fontsize_row      = 10,
  fontsize_col      = 10,
  angle_col         = "45",
  main              = sprintf(
    "%s — Jaccard Distance across Breast & Lung Datasets\n(top %d CCIs by |LOGFC|)",
    GENE_OF_INT, TOP_N_CCI
  ),
  legend_breaks     = goi_leg_breaks,
  legend_labels     = goi_leg_labels
)

dev.off()
# Expose for inspection
matrix_goi_jaccard   <- goi_jaccard_mat    # numeric distance matrix
matrix_goi_intersect <- goi_intersections  # list-matrix of shared CCI strings

# View(as.data.frame(matrix_goi_jaccard))
# goi_intersections[["Bassez2021_Breast_step4", "Chan2021_Lung_step4"]]  # example lookup

# ── AO: 6×6 distance heatmap for the same gene (full ranked lists) ────────────
message(sprintf("\n%s — Average Overlap distance matrix (full ranked CCI lists):",
                GENE_OF_INT))
goi_ao_mat <- build_gene_dataset_ao(GENE_OF_INT, all_cci_sets_ranked)
print(round(goi_ao_mat, 3))

goi_ao_display_mat <- goi_ao_mat
goi_ao_display_mat[is.na(goi_ao_display_mat) & row(goi_ao_display_mat) != col(goi_ao_display_mat)] <- 1

goi_ao_clust_mat <- goi_ao_mat
goi_ao_clust_mat[is.na(goi_ao_clust_mat)] <- 1
diag(goi_ao_clust_mat) <- 0

ao_cell_annot <- matrix(
  ifelse(row(goi_ao_mat) == col(goi_ao_mat), "",
         ifelse(is.na(goi_ao_mat), "N/A", sprintf("%.3f", goi_ao_mat))),
  nrow = nrow(goi_ao_mat),
  dimnames = dimnames(goi_ao_mat)
)

goi_ao_heatmap_file <- file.path(OUTPUT_DIR, paste0(GENE_OF_INT, "_ao_heatmap.png"))
png(goi_ao_heatmap_file, width = 5000, height = 3000, res = 300)

# Dynamic color bounds for the plotted matrix
goi_ao_min_val <- min(goi_ao_display_mat, na.rm = TRUE)
goi_ao_max_val <- max(goi_ao_display_mat, na.rm = TRUE)
goi_ao_breaks  <- seq(goi_ao_min_val, goi_ao_max_val, length.out = 101)
goi_ao_leg_breaks <- pretty(c(goi_ao_min_val, goi_ao_max_val), n = 5)
goi_ao_leg_breaks <- goi_ao_leg_breaks[goi_ao_leg_breaks >= goi_ao_min_val & goi_ao_leg_breaks <= goi_ao_max_val]
goi_ao_leg_labels <- format(round(goi_ao_leg_breaks, 3), trim = TRUE)

pheatmap(
  goi_ao_display_mat,
  color             = colorRampPalette(c("#4575b4", "#ffffbf", "#d73027"))(100),
  breaks            = goi_ao_breaks,
  na_col            = "grey80",
  cluster_rows      = TRUE,
  cluster_cols      = TRUE,
  clustering_distance_rows = as.dist(goi_ao_clust_mat),
  clustering_distance_cols = as.dist(goi_ao_clust_mat),
  display_numbers   = ao_cell_annot,
  number_color      = "black",
  fontsize_number   = 10,
  fontsize_row      = 10,
  fontsize_col      = 10,
  angle_col         = "45",
  main              = sprintf(
    "%s — Average Overlap Distance across Breast & Lung Datasets\n(full ranked CCIs by |LOGFC|)",
    GENE_OF_INT
  ),
  legend_breaks     = goi_ao_leg_breaks,
  legend_labels     = goi_ao_leg_labels
)
dev.off()

matrix_goi_ao <- goi_ao_mat
```








# Global Jaccard UMAP — All Genes × All Six Datasets
```{r global-jaccard-umap}
# if (!requireNamespace("batchelor", quietly = TRUE))
#   stop("Install batchelor via: BiocManager::install('batchelor')")
# library(batchelor)

# ── CONFIG (UMAP-specific) ────────────────────────────────────────────────────
K_UMAP        <- 15
MIN_DIST_UMAP <- 0.1
UMAP_CANCER_FILTER_OPTIONS <- c("H&N") # c("All", "Breast", "Lung", "H&N")

for (UMAP_CANCER_FILTER in UMAP_CANCER_FILTER_OPTIONS) {
message("\n===== UMAP_CANCER_FILTER: ", UMAP_CANCER_FILTER, " =====")

# ── 1. CCI vocabulary (restricted to included datasets when filtering) ────────
{
  included_ds_full <- if (UMAP_CANCER_FILTER == "All") {
    names(all_top_cci_sets)
  } else {
    names(all_top_cci_sets)[ds_cancer_map[names(all_top_cci_sets)] == UMAP_CANCER_FILTER]
  }
  all_top_ccis <- unique(unlist(lapply(all_top_cci_sets[included_ds_full],
                                       function(ds) unlist(unname(ds)))))
}
message("Unique CCIs in included datasets (", UMAP_CANCER_FILTER, "): ", length(all_top_ccis))

# ── 2. Gene_Dataset entries with binary CCI membership vectors ────────────────
gene_dataset_rows <- purrr::imap(all_top_cci_sets, function(ds_sets, ds_full) {
  ds_short <- ds_short_map[[ds_full]]
  cancer   <- ds_cancer_map[[ds_full]]
  purrr::imap(ds_sets, function(cci_vec, gene) {
    list(label       = paste0(gene, "_", ds_short),
         gene        = gene,
         dataset     = ds_short,
         cancer_type = cancer,
         ccis        = cci_vec,                           # raw CCI strings retained
         vec         = as.integer(all_top_ccis %in% cci_vec))
  })
}) %>% purrr::flatten()

# ── 2b. Filter by cancer type ─────────────────────────────────────────────────
if (!UMAP_CANCER_FILTER %in% c("All", "Breast", "Lung", "H&N"))
  stop("UMAP_CANCER_FILTER must be one of: \"All\", \"Breast\", \"Lung\", \"H&N\"")
if (UMAP_CANCER_FILTER != "All") {
  gene_dataset_rows <- gene_dataset_rows[
    sapply(gene_dataset_rows, function(x) x$cancer_type == UMAP_CANCER_FILTER)
  ]
  message("After cancer-type filter (", UMAP_CANCER_FILTER, "): ",
          length(gene_dataset_rows), " rows")
}

non_empty         <- sapply(gene_dataset_rows, function(x) sum(x$vec) > 0)
gene_dataset_rows <- gene_dataset_rows[non_empty]
message("Gene_Dataset pairs (non-empty): ", length(gene_dataset_rows))

# ── 3. Binary matrix (Gene_Dataset) × CCIs ────────────────────────────────────
labels_gd      <- sapply(gene_dataset_rows, `[[`, "label")
genes_gd       <- sapply(gene_dataset_rows, `[[`, "gene")
datasets_gd    <- sapply(gene_dataset_rows, `[[`, "dataset")
cancer_type_gd <- sapply(gene_dataset_rows, `[[`, "cancer_type")

bin_mat <- do.call(rbind, lapply(gene_dataset_rows, `[[`, "vec"))
rownames(bin_mat) <- labels_gd
message("Binary matrix: ", nrow(bin_mat), " rows × ", ncol(bin_mat), " CCI columns")

# ── 3b. CCI set lookup & pairwise intersection object ─────────────────────────
# Named list: Gene_Dataset label → character vector of its CCIs
gene_ds_ccis <- setNames(
  lapply(gene_dataset_rows, `[[`, "ccis"),
  labels_gd
)

# Helper: return CCIs shared between any two Gene_Dataset labels
get_mutual_ccis <- function(gd1, gd2) {
  s1 <- gene_ds_ccis[[gd1]]
  s2 <- gene_ds_ccis[[gd2]]
  if (is.null(s1) || is.null(s2)) stop("Label not found. Available: ", paste(names(gene_ds_ccis), collapse = ", "))
  intersect(s1, s2)
}

# Comprehensive pairwise intersection table (long format, list-column for CCIs)
# Builds upper-triangle only (n*(n-1)/2 rows) to avoid duplication.
{
  n   <- length(labels_gd)
  idx <- which(upper.tri(matrix(0, n, n)), arr.ind = TRUE)
  list_gene_cci_intersections <- tibble::tibble(
    GD1          = labels_gd[idx[, 1]],
    GD2          = labels_gd[idx[, 2]],
    mutual_ccis  = mapply(
      function(i, j) intersect(gene_ds_ccis[[labels_gd[i]]], gene_ds_ccis[[labels_gd[j]]]),
      idx[, 1], idx[, 2],
      SIMPLIFY = FALSE
    ),
    n_mutual     = sapply(
      seq_len(nrow(idx)),
      function(k) length(intersect(gene_ds_ccis[[labels_gd[idx[k, 1]]]],
                                   gene_ds_ccis[[labels_gd[idx[k, 2]]]]))
    )
  )
}
message("Pairwise intersection table: ", nrow(list_gene_cci_intersections), " pairs")

# ── 4. Jaccard distance (efficient matrix multiplication) ─────────────────────
cross     <- tcrossprod(bin_mat)
row_sums  <- rowSums(bin_mat)
union_mat <- outer(row_sums, row_sums, "+") - cross
union_mat[union_mat == 0] <- NA_real_

jac_dist_mat <- 1 - cross / union_mat
diag(jac_dist_mat) <- 0
jac_dist_mat[is.na(jac_dist_mat)] <- 1

# ── 5. UMAP ───────────────────────────────────────────────────────────────────
set.seed(SEED)
k_eff       <- min(K_UMAP, nrow(jac_dist_mat) - 1)
umap_global <- uwot::umap(as.dist(jac_dist_mat),
                           n_neighbors = k_eff, min_dist = MIN_DIST_UMAP)

# ── 6. Plot data frame ────────────────────────────────────────────────────────
umap_global_df <- data.frame(
  UMAP1       = umap_global[, 1],
  UMAP2       = umap_global[, 2],
  Label       = labels_gd,
  Gene        = genes_gd,
  Source      = datasets_gd,
  Cancer_Type = cancer_type_gd,
  stringsAsFactors = FALSE
)

# ── 7. Plot ───────────────────────────────────────────────────────────────────
umap_title_suffix <- if (UMAP_CANCER_FILTER == "All") "All Cancer Types" else
                       paste0(UMAP_CANCER_FILTER, " Only")
umap_out_tag      <- tolower(UMAP_CANCER_FILTER)   # "all" / "breast" / "lung"

p_global <- ggplot(umap_global_df,
                   aes(x = UMAP1, y = UMAP2,
                       colour = Source, shape = Cancer_Type)) +
  geom_point(size = 1.8, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = Gene),
    size = 2.5, max.overlaps = Inf,
    box.padding = 0.3, segment.size = 0.2, segment.alpha = 0.4,
    show.legend = FALSE) +
  scale_colour_manual(values = source_colors) +
  scale_shape_manual(values = cancer_shapes) +
  theme_classic(base_size = 13) +
  labs(
    title    = paste0("Global Gene UMAP — Jaccard Distance on CCI Profiles (",
                      umap_title_suffix, ")"),
    subtitle = sprintf("Each point = Gene × Dataset | top %d CCIs by |LOGFC|", TOP_N_CCI),
    colour = "Source", shape = "Cancer Type", x = "UMAP 1", y = "UMAP 2"
  )

print(p_global)
ggsave(file.path(OUTPUT_DIR, sprintf("global_jaccard_umap_%s.png", umap_out_tag)),
       plot = p_global, width = 13, height = 9, dpi = 300)
message("Global UMAP saved (", UMAP_CANCER_FILTER, ")")

# ── 5b. MNN-corrected UMAP (Jaccard) ─────────────────────────────────────────
# bin_mat (Gene_Dataset x CCIs) → fastMNN applies internal PCA then finds
# Mutual Nearest Neighbors across dataset batches → UMAP on corrected embedding.
message("Applying MNN batch correction on Jaccard binary matrix ...")
k_mnn_jac <- min(20L, min(table(factor(datasets_gd))) - 1L)
mnn_jac <- batchelor::fastMNN(
  t(bin_mat),
  batch  = factor(datasets_gd),
  d      = min(50L, nrow(bin_mat) - 1L),
  k      = k_mnn_jac
)
corrected_jac <- as.matrix(SingleCellExperiment::reducedDim(mnn_jac, "corrected"))
rownames(corrected_jac) <- labels_gd
message("MNN (Jaccard) correction complete; corrected dim: ",
        nrow(corrected_jac), " x ", ncol(corrected_jac))

set.seed(SEED)
k_eff_mnn_jac <- min(K_UMAP, nrow(corrected_jac) - 1L)
umap_mnn_jac  <- uwot::umap(corrected_jac,
                              n_neighbors = k_eff_mnn_jac,
                              min_dist    = MIN_DIST_UMAP)

umap_mnn_jac_df <- data.frame(
  UMAP1       = umap_mnn_jac[, 1],
  UMAP2       = umap_mnn_jac[, 2],
  Label       = labels_gd,
  Gene        = genes_gd,
  Source      = datasets_gd,
  Cancer_Type = cancer_type_gd,
  stringsAsFactors = FALSE
)

p_global_mnn_jac <- ggplot(umap_mnn_jac_df,
                            aes(x = UMAP1, y = UMAP2,
                                colour = Source, shape = Cancer_Type)) +
  geom_point(size = 1.8, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = Gene),
    size = 2.5, max.overlaps = Inf,
    box.padding = 0.3, segment.size = 0.2, segment.alpha = 0.4,
    show.legend = FALSE) +
  scale_colour_manual(values = source_colors) +
  scale_shape_manual(values = cancer_shapes) +
  theme_classic(base_size = 13) +
  labs(
    title    = paste0("Global Gene UMAP — Jaccard + MNN Batch Correction (",
                      umap_title_suffix, ")"),
    subtitle = sprintf(
      "Each point = Gene x Dataset | top %d CCIs by |LOGFC| | batchelor::fastMNN corrected",
      TOP_N_CCI),
    colour = "Source", shape = "Cancer Type", x = "UMAP 1", y = "UMAP 2"
  )

print(p_global_mnn_jac)
ggsave(file.path(OUTPUT_DIR, sprintf("global_jaccard_mnn_umap_%s.png", umap_out_tag)),
       plot = p_global_mnn_jac, width = 13, height = 9, dpi = 300)
message("Global Jaccard MNN UMAP saved (", UMAP_CANCER_FILTER, ")")

# ── Expose for inspection ────────────────────────────────────────────────────
matrix_umap_input <- bin_mat   # numeric: rows = Gene_Dataset, cols = CCIs (binary)
# list_gene_cci_intersections   # tibble with columns GD1, GD2, mutual_ccis (list), n_mutual
# gene_ds_ccis                  # named list: label → character vector of CCIs
# get_mutual_ccis(gd1, gd2)     # helper: returns shared CCI strings for any two labels

# ── AO UMAP (full ranked lists, same gene×dataset pairs) ─────────────────────
 # Build a reverse map from short dataset name to full dataset key
ds_full_map <- setNames(names(ds_short_map), ds_short_map)

# Retrieve the full ranked CCI list for each gene×dataset pair already in labels_gd
gene_ds_ccis_ranked <- setNames(
  lapply(gene_dataset_rows, function(x) {
    ds_full <- ds_full_map[[x$dataset]]
    all_cci_sets_ranked[[ds_full]][[x$gene]]
  }),
  labels_gd
)

message("Computing pairwise Average Overlap distances for ", length(labels_gd),
        " Gene×Dataset pairs — this may take several minutes...")
n_gd <- length(labels_gd)
ao_dist_mat <- matrix(0, nrow = n_gd, ncol = n_gd, dimnames = list(labels_gd, labels_gd))
for (i in seq_len(n_gd - 1)) {
  for (j in (i + 1):n_gd) {
    d <- dist_fun_superranker_overlap(gene_ds_ccis_ranked[[labels_gd[i]]],
                                     gene_ds_ccis_ranked[[labels_gd[j]]])
    ao_dist_mat[i, j] <- d
    ao_dist_mat[j, i] <- d
  }
}
ao_dist_mat[is.na(ao_dist_mat)] <- 1
message("AO distance matrix complete.")

set.seed(SEED)
k_eff_ao  <- min(K_UMAP, nrow(ao_dist_mat) - 1)
umap_ao   <- uwot::umap(as.dist(ao_dist_mat),
                         n_neighbors = k_eff_ao, min_dist = MIN_DIST_UMAP)

umap_ao_df <- data.frame(
  UMAP1       = umap_ao[, 1],
  UMAP2       = umap_ao[, 2],
  Label       = labels_gd,
  Gene        = genes_gd,
  Source      = datasets_gd,
  Cancer_Type = cancer_type_gd,
  stringsAsFactors = FALSE
)

p_global_ao <- ggplot(umap_ao_df,
                      aes(x = UMAP1, y = UMAP2,
                          colour = Source, shape = Cancer_Type)) +
  geom_point(size = 1.8, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = Gene),
    size = 2.5, max.overlaps = Inf,
    box.padding = 0.3, segment.size = 0.2, segment.alpha = 0.4,
    show.legend = FALSE) +
  scale_colour_manual(values = source_colors) +
  scale_shape_manual(values = cancer_shapes) +
  theme_classic(base_size = 13) +
  labs(
    title    = paste0("Global Gene UMAP — Average Overlap Distance on CCI Profiles (",
                      umap_title_suffix, ")"),
    subtitle = "Each point = Gene × Dataset | full ranked CCIs by |LOGFC|",
    colour = "Source", shape = "Cancer Type", x = "UMAP 1", y = "UMAP 2"
  )

print(p_global_ao)
ggsave(file.path(OUTPUT_DIR, sprintf("global_ao_umap_%s.png", umap_out_tag)),
       plot = p_global_ao, width = 13, height = 9, dpi = 300)
message("Global AO UMAP saved (", UMAP_CANCER_FILTER, ")")

# ── AO: MNN-corrected UMAP (Average Overlap) ─────────────────────────────────
# ao_dist_mat has no feature matrix, so classical MDS is used first to produce
# Euclidean coordinates; fastMNN then corrects those coordinates across batches.
message("Applying MNN batch correction on Average Overlap distance matrix ...")
n_mds_ao      <- min(50L, nrow(ao_dist_mat) - 1L)
mds_ao_coords <- cmdscale(as.dist(ao_dist_mat), k = n_mds_ao)

k_mnn_ao <- min(20L, min(table(factor(datasets_gd))) - 1L)
mnn_ao <- batchelor::fastMNN(
  t(mds_ao_coords),
  batch  = factor(datasets_gd),
  d      = min(30L, n_mds_ao),
  k      = k_mnn_ao
)
corrected_ao <- as.matrix(SingleCellExperiment::reducedDim(mnn_ao, "corrected"))
rownames(corrected_ao) <- labels_gd
message("MNN (AO) correction complete; corrected dim: ",
        nrow(corrected_ao), " x ", ncol(corrected_ao))

set.seed(SEED)
k_eff_mnn_ao <- min(K_UMAP, nrow(corrected_ao) - 1L)
umap_mnn_ao  <- uwot::umap(corrected_ao,
                             n_neighbors = k_eff_mnn_ao,
                             min_dist    = MIN_DIST_UMAP)

umap_mnn_ao_df <- data.frame(
  UMAP1       = umap_mnn_ao[, 1],
  UMAP2       = umap_mnn_ao[, 2],
  Label       = labels_gd,
  Gene        = genes_gd,
  Source      = datasets_gd,
  Cancer_Type = cancer_type_gd,
  stringsAsFactors = FALSE
)

p_global_mnn_ao <- ggplot(umap_mnn_ao_df,
                           aes(x = UMAP1, y = UMAP2,
                               colour = Source, shape = Cancer_Type)) +
  geom_point(size = 1.8, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = Gene),
    size = 2.5, max.overlaps = Inf,
    box.padding = 0.3, segment.size = 0.2, segment.alpha = 0.4,
    show.legend = FALSE) +
  scale_colour_manual(values = source_colors) +
  scale_shape_manual(values = cancer_shapes) +
  theme_classic(base_size = 13) +
  labs(
    title    = paste0("Global Gene UMAP — Average Overlap + MNN Batch Correction (",
                      umap_title_suffix, ")"),
    subtitle = "Each point = Gene x Dataset | full ranked CCIs by |LOGFC| | batchelor::fastMNN corrected",
    colour = "Source", shape = "Cancer Type", x = "UMAP 1", y = "UMAP 2"
  )

print(p_global_mnn_ao)
ggsave(file.path(OUTPUT_DIR, sprintf("global_ao_mnn_umap_%s.png", umap_out_tag)),
       plot = p_global_mnn_ao, width = 13, height = 10, dpi = 300)
message("Global AO MNN UMAP saved (", UMAP_CANCER_FILTER, ")")

} # end for (UMAP_CANCER_FILTER in c("All", "Breast", "Lung"))
# NOTE: labels_gd, gene_ds_ccis, list_gene_cci_intersections below reflect
# the last loop iteration ("Lung"). Re-run that block manually for other filters.

# ── Gene-of-interest cross-dataset CCI analysis (set GENE_OF_INT in config) ──
goi_labels <- grep(paste0("^", GENE_OF_INT, "_"), labels_gd, value = TRUE)
message(GENE_OF_INT, " found in datasets: ", paste(goi_labels, collapse = ", "))

if (length(goi_labels) >= 2) {
  goi_sets   <- gene_ds_ccis[goi_labels]
  goi_all    <- Reduce(union,     goi_sets)
  goi_shared <- Reduce(intersect, goi_sets)
  message(GENE_OF_INT, " union CCIs (any dataset): ",      length(goi_all))
  message(GENE_OF_INT, " conserved CCIs (all datasets): ", length(goi_shared))

  # Per-label: CCIs unique to that dataset (not present in any other dataset for this gene)
  goi_unique <- setNames(
    lapply(goi_labels, function(lbl) {
      setdiff(goi_sets[[lbl]], Reduce(union, goi_sets[setdiff(goi_labels, lbl)]))
    }),
    goi_labels
  )

  # Summary table
  goi_summary <- data.frame(
    Label       = goi_labels,
    N_total     = sapply(goi_sets,   length),
    N_conserved = sapply(goi_sets,   function(s) length(intersect(s, goi_shared))),
    N_unique    = sapply(goi_unique, length),
    stringsAsFactors = FALSE
  )
  print(goi_summary)

  if (length(goi_shared) > 0) {
    message("\nCCIs conserved across ALL ", GENE_OF_INT, " datasets:")
    print(sort(goi_shared))
  } else {
    message("No CCIs conserved across ALL ", GENE_OF_INT, " datasets.")
  }

  # Pairwise intersections between all datasets for this gene
  goi_pairwise <- list_gene_cci_intersections %>%
    dplyr::filter(GD1 %in% goi_labels & GD2 %in% goi_labels) %>%
    dplyr::arrange(dplyr::desc(n_mutual))
  if (nrow(goi_pairwise) > 0) print(goi_pairwise[, c("GD1", "GD2", "n_mutual")])
} else {
  message(GENE_OF_INT, " not found (or only in one dataset) — check GENE_OF_INT in config.")
}

# Quick-access convenience (uncomment to open in RStudio viewer)
# View(umap_global_df)
# View(as.data.frame(matrix_umap_input[1:50, 1:20]))
# View(list_gene_cci_intersections)
# get_mutual_ccis(paste0(GENE_OF_INT, "_Lung"), paste0(GENE_OF_INT, "_Breast"))
```





# Gene-Gene Consensus Jaccard Matrix
```{r gene-gene-clustering}
# ── CONFIG ────────────────────────────────────────────────────────────────────
TOP_N_GENES <- 400  # genes ranked by total union CCI count
GENE_GENE_TOP_VAR_CCI_PCT <- 0.05   # top X% CCIs by Gene×Dataset logFC variance (HVG)
GENE_GENE_MIN_OBS_PER_CCI <- 10      # min non-NA Gene×Dataset rows per CCI for HVG ranking
GENE_GENE_CANCER_FILTER_OPTIONS <- c("H&N") # c("All", "Breast", "Lung", "H&N")
for (GENE_GENE_CANCER_FILTER in GENE_GENE_CANCER_FILTER_OPTIONS) {
message("\n===== GENE_GENE_CANCER_FILTER: ", GENE_GENE_CANCER_FILTER, " =====")

# ── 1. Union CCI set per gene (restricted to included datasets) ───────────────
if (!GENE_GENE_CANCER_FILTER %in% c("All", "Breast", "Lung", "H&N"))
  stop("GENE_GENE_CANCER_FILTER must be one of: \"All\", \"Breast\", \"Lung\", \"H&N\"")
{
  included_cci_sets <- if (GENE_GENE_CANCER_FILTER == "All") {
    all_top_cci_sets
  } else {
    all_top_cci_sets[ds_cancer_map[names(all_top_cci_sets)] == GENE_GENE_CANCER_FILTER]
  }
}
message("Gene–Gene matrix using datasets (", GENE_GENE_CANCER_FILTER, "): ",
        paste(names(included_cci_sets), collapse = ", "))

# HVG CCIs from logFC space (same logic as post-analysis-hvg-ccis).
malignant_for_hvg <- setNames(
  lapply(names(included_cci_sets), function(ds) {
    obj_name <- paste0(ds, ".malignant")
    if (!exists(obj_name, inherits = TRUE)) {
      stop("Missing ", obj_name, " — required for gene–gene HVG CCI selection.")
    }
    get(obj_name, inherits = TRUE)
  }),
  names(included_cci_sets)
)
hvg_sel <- select_hvg_ccis(
  malignant_for_hvg,
  ds_cancer_map[names(malignant_for_hvg)],
  top_pct = GENE_GENE_TOP_VAR_CCI_PCT,
  min_obs = GENE_GENE_MIN_OBS_PER_CCI
)
gene_gene_hvg_ccis <- hvg_sel$hvg
message(sprintf(
  "Gene–gene HVG CCIs: %d / %d eligible (top %.0f%% by variance, cutoff = %.4f)",
  length(gene_gene_hvg_ccis), hvg_sel$n_eligible,
  GENE_GENE_TOP_VAR_CCI_PCT * 100, hvg_sel$var_cutoff
))

all_gene_names <- sort(unique(unlist(lapply(included_cci_sets, names))))
gene_cci_union <- setNames(
  lapply(all_gene_names, function(g) {
    unique(unlist(lapply(included_cci_sets, function(ds) ds[[g]])))
  }),
  all_gene_names
)
gene_cci_union <- lapply(gene_cci_union, restrict_ccis_to_hvg, hvg_ccis = gene_gene_hvg_ccis)
gene_cci_union <- Filter(function(x) length(x) > 0, gene_cci_union)
message("Genes with at least one HVG CCI (union): ", length(gene_cci_union))

# ── 2. Top N genes by union CCI count ─────────────────────────────────────────
cci_counts <- sapply(gene_cci_union, length)
top_genes  <- sort(names(sort(cci_counts, decreasing = TRUE))[
  seq_len(min(TOP_N_GENES, length(cci_counts)))])
gene_cci_top <- gene_cci_union[top_genes]
message("Top ", length(top_genes), " genes selected")

# ── 3. Gene × CCI binary matrix ───────────────────────────────────────────────
all_ccis_top <- unique(unlist(gene_cci_top))
gene_bin_mat <- do.call(rbind, lapply(gene_cci_top, function(v) as.integer(all_ccis_top %in% v)))
rownames(gene_bin_mat) <- top_genes
message("Gene binary matrix: ", nrow(gene_bin_mat), " × ", ncol(gene_bin_mat))

# ── 3b. Gene-level pairwise CCI intersection object ───────────────────────────
# Helper: return CCIs shared between any two top genes (union-across-datasets sets)
get_mutual_ccis_genes <- function(gene1, gene2) {
  s1 <- gene_cci_top[[gene1]]
  s2 <- gene_cci_top[[gene2]]
  if (is.null(s1) || is.null(s2)) stop("Gene not found. Available: ", paste(top_genes, collapse = ", "))
  intersect(s1, s2)
}

# Pairwise intersection table for all top-gene pairs (upper-triangle, long format)
{
  ng  <- length(top_genes)
  idx_g <- which(upper.tri(matrix(0, ng, ng)), arr.ind = TRUE)
  list_gene_pairwise_intersections <- tibble::tibble(
    GeneA        = top_genes[idx_g[, 1]],
    GeneB        = top_genes[idx_g[, 2]],
    mutual_ccis  = mapply(
      function(i, j) intersect(gene_cci_top[[top_genes[i]]], gene_cci_top[[top_genes[j]]]),
      idx_g[, 1], idx_g[, 2],
      SIMPLIFY = FALSE
    ),
    n_mutual     = sapply(
      seq_len(nrow(idx_g)),
      function(k) length(intersect(gene_cci_top[[top_genes[idx_g[k, 1]]]],
                                   gene_cci_top[[top_genes[idx_g[k, 2]]]]))
    )
  )
}
message("Gene pairwise intersection table: ", nrow(list_gene_pairwise_intersections), " pairs")

# ── 4. Consensus Jaccard distance (Gene × Gene) ───────────────────────────────
cross_g <- tcrossprod(gene_bin_mat)
rsums_g <- rowSums(gene_bin_mat)
union_g <- outer(rsums_g, rsums_g, "+") - cross_g
union_g[union_g == 0] <- NA_real_

gene_jac_dist <- 1 - cross_g / union_g
diag(gene_jac_dist) <- NA_real_
gene_jac_dist_clust <- gene_jac_dist
gene_jac_dist_clust[is.na(gene_jac_dist_clust)] <- 1
diag(gene_jac_dist_clust) <- 0
gene_jac_dist[is.na(gene_jac_dist) & row(gene_jac_dist) != col(gene_jac_dist)] <- 1

matrix_global_gene_distance <- gene_jac_dist

# ── 5. Heatmap ────────────────────────────────────────────────────────────────
{
  disp_nums_jac <- matrix(sprintf("%.2f", gene_jac_dist),
                          nrow = nrow(gene_jac_dist), dimnames = dimnames(gene_jac_dist))
  diag(disp_nums_jac) <- ""
}
heatmap_args <- list(
  gene_jac_dist,
  color           = colorRampPalette(c("#4575b4", "#ffffbf", "#d73027"))(100),
  breaks          = seq(min(gene_jac_dist, na.rm = TRUE),
                        max(gene_jac_dist, na.rm = TRUE),
                        length.out = 101),
  na_col          = "grey80",
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  # clustering_distance_rows = as.dist(gene_jac_dist_clust),
  # clustering_distance_cols = as.dist(gene_jac_dist_clust),
  #display_numbers = disp_nums_jac,
  #number_color    = "black",
  #fontsize_number = 8,
  fontsize_row    = max(4, 120 / length(top_genes)),
  fontsize_col    = max(4, 120 / length(top_genes)),
  angle_col       = "45",
  main            = sprintf(
    "Gene–Gene Jaccard Distance — HVG CCIs only (%s)",
    if (GENE_GENE_CANCER_FILTER == "All") "consensus across all datasets"
    else paste0(GENE_GENE_CANCER_FILTER, " datasets only"))
)

do.call(pheatmap, heatmap_args)

gene_heatmap_file <- file.path(OUTPUT_DIR,
  sprintf("gene_gene_jaccard_heatmap_%s.png", tolower(GENE_GENE_CANCER_FILTER)))
do.call(pheatmap, c(heatmap_args, list(filename = gene_heatmap_file,
                                       width    = 3000 / 300,
                                       height   = 3000 / 300)))
message("Heatmap saved to ", gene_heatmap_file)

# ── RBO: Gene × Gene distance matrix (same top genes, full ranked lists) ─────
# For each top gene, build a consensus ranked CCI list across included datasets:
# CCIs appearing in more datasets rank first; ties broken by mean rank position.
{
  included_cci_sets_ranked <- if (GENE_GENE_CANCER_FILTER == "All") {
    all_cci_sets_ranked
  } else {
    all_cci_sets_ranked[ds_cancer_map[names(all_cci_sets_ranked)] == GENE_GENE_CANCER_FILTER]
  }
}

gene_cci_consensus_ranked <- setNames(
  lapply(top_genes, function(g) {
    rank_data <- purrr::imap_dfr(included_cci_sets_ranked, function(ds_sets, ds_name) {
      cci_vec <- ds_sets[[g]]
      if (is.null(cci_vec) || length(cci_vec) == 0) return(NULL)
      data.frame(CCI = cci_vec, rank_pos = seq_along(cci_vec), stringsAsFactors = FALSE)
    })
    if (nrow(rank_data) == 0) return(character(0))
    rank_data %>%
      group_by(CCI) %>%
      summarise(n_datasets = n(), mean_rank = mean(rank_pos), .groups = "drop") %>%
      arrange(desc(n_datasets), mean_rank) %>%
      pull(CCI) %>%
      restrict_ccis_to_hvg(gene_gene_hvg_ccis)
  }),
  top_genes
)


# ── AO: Gene × Gene distance matrix (same top genes, full ranked lists) ──────
message("Computing gene–gene Average Overlap distances for ", length(top_genes), " genes...")
ng <- length(top_genes)
gene_ao_dist <- matrix(0, nrow = ng, ncol = ng,
                       dimnames = list(top_genes, top_genes))
for (i in seq_len(ng - 1)) {
  for (j in (i + 1):ng) {
    d <- dist_fun_superranker_overlap(gene_cci_consensus_ranked[[top_genes[i]]],
                                     gene_cci_consensus_ranked[[top_genes[j]]])
    gene_ao_dist[i, j] <- d
    gene_ao_dist[j, i] <- d
  }
}
gene_ao_dist_clust <- gene_ao_dist
gene_ao_dist_clust[is.na(gene_ao_dist_clust)] <- 1
diag(gene_ao_dist) <- NA_real_
matrix_global_gene_ao_distance <- gene_ao_dist
message("Gene–gene AO matrix complete.")

{
  disp_nums_ao <- matrix(sprintf("%.2f", gene_ao_dist),
                         nrow = nrow(gene_ao_dist), dimnames = dimnames(gene_ao_dist))
  diag(disp_nums_ao) <- ""
}
ao_heatmap_args <- list(
  gene_ao_dist,
  color           = colorRampPalette(c("#4575b4", "#ffffbf", "#d73027"))(100),
  breaks          = seq(min(gene_ao_dist, na.rm = TRUE),
                        max(gene_ao_dist, na.rm = TRUE),
                        length.out = 101),
  na_col          = "grey80",
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  # clustering_distance_rows = as.dist(gene_ao_dist_clust),
  # clustering_distance_cols = as.dist(gene_ao_dist_clust),
 # display_numbers = disp_nums_ao,
  #number_color    = "black",
  #fontsize_number = 8,
  fontsize_row    = max(4, 120 / length(top_genes)),
  fontsize_col    = max(4, 120 / length(top_genes)),
  angle_col       = "45",
  main            = sprintf(
    "Gene–Gene Average Overlap Distance — HVG CCIs only (%s)",
    if (GENE_GENE_CANCER_FILTER == "All") "consensus across all datasets"
    else paste0(GENE_GENE_CANCER_FILTER, " datasets only"))
)

do.call(pheatmap, ao_heatmap_args)

gene_ao_heatmap_file <- file.path(OUTPUT_DIR,
  sprintf("gene_gene_ao_heatmap_%s.png", tolower(GENE_GENE_CANCER_FILTER)))
do.call(pheatmap, c(ao_heatmap_args, list(filename = gene_ao_heatmap_file,
                                           width    = 3000 / 300,
                                           height   = 3000 / 300)))
message("AO heatmap saved to ", gene_ao_heatmap_file)

} # end for (GENE_GENE_CANCER_FILTER in c("All", "Breast", "Lung"))
# NOTE: matrix_global_gene_* and gene_cci_* below reflect the last iteration ("Lung").

# ── Expose for inspection ────────────────────────────────────────────────────
# matrix_global_gene_distance          # Gene × Gene Jaccard distance matrix (top-500 lists)
# matrix_global_gene_ao_distance       # Gene × Gene AO distance matrix (full ranked lists)
# list_gene_pairwise_intersections      # tibble: GeneA, GeneB, mutual_ccis (list), n_mutual
# gene_cci_top                          # named list: gene → union CCI character vector (Jaccard)
# gene_cci_consensus_ranked             # named list: gene → consensus ranked CCI vector (AO)
# get_mutual_ccis_genes(gene1, gene2)   # helper: shared CCIs for any two top genes

# Quick-access convenience (uncomment to open in RStudio viewer)
# View(as.data.frame(matrix_global_gene_distance))
# View(list_gene_pairwise_intersections)
# get_mutual_ccis_genes("CTLA4", "CD274")
```


# GENES x LRI Heatmap & UMAP
```{r genes-lri}
library(umap)

build_lri_logfc_mat <- function(malignant_list, min_genes = MIN_GENES_PER_CCI) {
  lri_list <- lapply(names(malignant_list), function(gene) {
    malignant_list[[gene]] %>%
      select(LRI, LOGFC) %>%
      group_by(LRI) %>%
      summarise(LOGFC = mean(LOGFC, na.rm = TRUE), .groups = "drop") %>%
      mutate(Gene = gene)
  })

  all_data <- bind_rows(lri_list)
  mat_long <- all_data %>%
    pivot_wider(names_from = Gene, values_from = LOGFC)

  mat <- as.matrix(mat_long[, -1])
  rownames(mat) <- mat_long$LRI
  message("  Matrix dimensions (LRIs x Genes): ", nrow(mat), " x ", ncol(mat))

  n_non_na <- rowSums(!is.na(mat))
  mat <- mat[n_non_na >= min_genes, , drop = FALSE]
  message("  After filtering sparse LRIs: ", nrow(mat), " x ", ncol(mat))
  mat
}

plot_lri_heatmap_umap <- function(malignant_list, dataset_label,
                                  k_clusters = 5,
                                  min_genes  = MIN_GENES_PER_CCI) {
  message("\n=== GENES x LRI: ", dataset_label, " ===")

  mat     <- build_lri_logfc_mat(malignant_list, min_genes)
  sim_mat <- cosine_sim_na(mat)
  diag(sim_mat) <- NA

  gene_order <- sort(rownames(sim_mat))
  sim_mat    <- sim_mat[gene_order, gene_order]
  n_genes    <- nrow(sim_mat)

  sim_mat_clust <- sim_mat
  diag(sim_mat_clust) <- 1
  sim_mat_clust[is.na(sim_mat_clust)] <- 0
  dist_for_clust <- as.dist(1 - sim_mat_clust)

  # Heatmap
  heatmap_file <- file.path(OUTPUT_DIR, paste0(dataset_label, "_lri_heatmap.png"))
  png(heatmap_file,
      width  = max(4000, n_genes * 20),
      height = max(4000, n_genes * 20),
      res    = 300)
  pheatmap(
    sim_mat,
    color        = colorRampPalette(c("red", "white", "blue"))(100),
    breaks       = seq(-1, 1, length.out = 101),
    na_col       = "grey80",
    fontsize_row = max(4, 200 / n_genes),
    fontsize_col = max(4, 200 / n_genes),
    angle_col    = "45",
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    clustering_distance_rows = dist_for_clust,
    clustering_distance_cols = dist_for_clust,
    main         = paste0(dataset_label, " – Gene-Gene Cosine Similarity (LRI)")
  )
  dev.off()
  message("  Heatmap saved to ", heatmap_file)
# 
#   # UMAP
#   mat_genes         <- t(mat)
#   mat_genes_imputed <- mat_genes
#   mat_genes_imputed[is.na(mat_genes_imputed)] <- 0
# 
#   dist_mat   <- as.dist(1 - sim_mat)
#   hclust_res <- hclust(dist_mat, method = "complete")
#   clusters   <- cutree(hclust_res, k = k_clusters)
# 
#   set.seed(42)
#   umap_config             <- umap.defaults
#   umap_config$n_neighbors <- min(15, nrow(mat_genes_imputed) - 1)
#   umap_config$min_dist    <- 0.1
#   umap_config$metric      <- "cosine"
# 
#   umap_result <- umap(mat_genes_imputed, config = umap_config)
# 
#   umap_df <- data.frame(
#     Gene    = rownames(mat_genes_imputed),
#     UMAP1   = umap_result$layout[, 1],
#     UMAP2   = umap_result$layout[, 2],
#     n_LRIs  = rowSums(!is.na(mat_genes)),
#     Cluster = factor(clusters[rownames(mat_genes_imputed)])
#   )
# 
#   p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = Cluster, label = Gene)) +
#     geom_point(size = 3, alpha = 0.9) +
#     geom_text_repel(size = 2.8, max.overlaps = 40,
#                     segment.size = 0.2, segment.alpha = 0.5, show.legend = FALSE) +
#     scale_colour_brewer(palette = "Set1", name = "Cluster") +
#     labs(
#       title    = paste0(dataset_label, " – Gene UMAP (LRI LOGFC space)"),
#       subtitle = paste0("Hierarchical clustering (k = ", k_clusters, ", cosine distance)"),
#       x = "UMAP 1", y = "UMAP 2"
#     ) +
#     theme_bw(base_size = 11) +
#     theme(plot.title = element_text(face = "bold"))
# 
#   umap_file <- file.path(OUTPUT_DIR, paste0(dataset_label, "_lri_gene_umap.png"))
#   ggsave(umap_file, plot = p_umap, width = 10, height = 8, dpi = 300)
#   message("  UMAP saved to ", umap_file)
# 
#   cluster_df <- umap_df %>%
#     select(Gene, Cluster, n_LRIs) %>%
#     arrange(Cluster, Gene)
#   print(cluster_df)
# 
#   cluster_summary <- cluster_df %>%
#     group_by(Cluster) %>%
#     summarise(n_genes = n(), genes = paste(sort(Gene), collapse = ", ")) %>%
#     arrange(Cluster)
#   print(cluster_summary)
# 
#   invisible(list(umap_df = umap_df, clusters = clusters))
}

# plot_lri_heatmap_umap(Bassez2021_Breast_step4.malignant, "Bassez2021_Breast_step4")
# plot_lri_heatmap_umap(Qian2020_Breast_step4.malignant,   "Qian2020_Breast_step4")
# plot_lri_heatmap_umap(Wu2021_Breast_step4.malignant,     "Wu2021_Breast_step4")

# plot_lri_heatmap_umap(Chan2021_Lung_step4.malignant,     "Chan2021_Lung_step4")
# plot_lri_heatmap_umap(Xing2021_Lung_step4.malignant,     "Xing2021_Lung_step4")
# plot_lri_heatmap_umap(Laughney2020_Lung_step4.malignant, "Laughney2020_Lung_step4")

plot_lri_heatmap_umap(Kurten_HNSC.malignant, "Kurten_HNSC")
plot_lri_heatmap_umap(Puram_HNSC.malignant,  "Puram_HNSC")
plot_lri_heatmap_umap(Choi_HNSC.malignant,   "Choi_HNSC")
plot_lri_heatmap_umap(Bill_HNSC.malignant,   "Bill_HNSC")
```


# Gene × Dataset UMAP — Combined LRI LogFC Space
```{r gene-dataset-lri-umap}
# ── CONFIG ────────────────────────────────────────────────────────────────────
K_LRI_UMAP  <- 15
MINDIST_LRI <- 0.1

LRI_UMAP_CANCER_FILTER_OPTIONS <- c("H&N") # c("All", "Breast", "Lung", "H&N")
for (LRI_UMAP_CANCER_FILTER in LRI_UMAP_CANCER_FILTER_OPTIONS) {
message("\n===== LRI_UMAP_CANCER_FILTER: ", LRI_UMAP_CANCER_FILTER, " =====")

# Full catalogue — filter is applied below
.lri_all_datasets <- list(
  # Bassez2021_Breast_step4  = Bassez2021_Breast_step4.malignant,
  # Qian2020_Breast_step4    = Qian2020_Breast_step4.malignant,
  # Wu2021_Breast_step4      = Wu2021_Breast_step4.malignant,
  # Chan2021_Lung_step4      = Chan2021_Lung_step4.malignant,
  # Laughney2020_Lung_step4  = Laughney2020_Lung_step4.malignant,
  # Xing2021_Lung_step4      = Xing2021_Lung_step4.malignant,
  Kurten_HNSC              = Kurten_HNSC.malignant,
  Puram_HNSC               = Puram_HNSC.malignant,
  Choi_HNSC                = Choi_HNSC.malignant,
  Bill_HNSC                = Bill_HNSC.malignant
)
.lri_all_cancer <- c(
  # Bassez2021_Breast_step4  = "Breast",
  # Qian2020_Breast_step4    = "Breast",
  # Wu2021_Breast_step4      = "Breast",
  # Chan2021_Lung_step4      = "Lung",
  # Laughney2020_Lung_step4  = "Lung",
  # Xing2021_Lung_step4      = "Lung",
  Kurten_HNSC              = "H&N",
  Puram_HNSC               = "H&N",
  Choi_HNSC                = "H&N",
  Bill_HNSC                = "H&N"
)
.lri_all_colors <- c(
  # Bassez2021_Breast_step4  = "#E41A1C",
  # Qian2020_Breast_step4    = "#FF7F00",
  # Wu2021_Breast_step4      = "#4DAF4A",
  # Chan2021_Lung_step4      = "#377EB8",
  # Laughney2020_Lung_step4  = "#984EA3",
  # Xing2021_Lung_step4      = "#A65628",
  Kurten_HNSC              = "#F781BF",
  Puram_HNSC               = "#999999",
  Choi_HNSC                = "#66C2A5",
  Bill_HNSC                = "#FC8D62"
)

# Apply filter
if (!LRI_UMAP_CANCER_FILTER %in% c("All", "Breast", "Lung", "H&N"))
  stop("LRI_UMAP_CANCER_FILTER must be one of: \"All\", \"Breast\", \"Lung\", \"H&N\"")
{
  keep_ds <- if (LRI_UMAP_CANCER_FILTER == "All") names(.lri_all_datasets) else
               names(.lri_all_datasets)[.lri_all_cancer[names(.lri_all_datasets)] == LRI_UMAP_CANCER_FILTER]
  LRI_UMAP_DATASETS <- .lri_all_datasets[keep_ds]
  LRI_UMAP_CANCER   <- .lri_all_cancer[keep_ds]
  LRI_UMAP_COLORS   <- .lri_all_colors[keep_ds]
}
message("LRI UMAP datasets (", LRI_UMAP_CANCER_FILTER, "): ", paste(names(LRI_UMAP_DATASETS), collapse = ", "))

# ── 1. Build per-dataset LRI × Gene LogFC matrices ───────────────────────────
lri_mats <- lapply(LRI_UMAP_DATASETS, build_lri_logfc_mat)

# ── 2. Union LRI space ────────────────────────────────────────────────────────
all_lris <- sort(unique(unlist(lapply(lri_mats, rownames))))
message("Union LRI space: ", length(all_lris), " LRIs")

# ── 3. Build Gene × Dataset feature matrix (rows = Gene_Dataset, cols = LRIs) ─
gd_rows <- purrr::imap(lri_mats, function(mat, ds) {
  # mat is LRI × Gene — transpose to Gene × LRI, then align to union LRI space
  gene_names <- colnames(mat)
  lapply(gene_names, function(g) {
    vec <- setNames(rep(NA_real_, length(all_lris)), all_lris)
    present <- rownames(mat)[!is.na(mat[, g])]
    vec[present] <- mat[present, g]
    list(label   = paste0(g, "_", ds),
         gene    = g,
         dataset = ds,
         cancer  = LRI_UMAP_CANCER[[ds]],
         vec     = vec)
  })
}) %>% purrr::flatten()

# Remove rows that are entirely NA (gene absent in that dataset)
non_empty_gd <- sapply(gd_rows, function(x) any(!is.na(x$vec)))
gd_rows      <- gd_rows[non_empty_gd]
message("Gene × Dataset points (non-empty): ", length(gd_rows))

labels_gd_lri  <- sapply(gd_rows, `[[`, "label")
genes_gd_lri   <- sapply(gd_rows, `[[`, "gene")
datasets_gd_lri <- sapply(gd_rows, `[[`, "dataset")
cancer_gd_lri  <- sapply(gd_rows, `[[`, "cancer")

# ── 4. Impute NAs with 0 for UMAP input ───────────────────────────────────────
feat_mat <- do.call(rbind, lapply(gd_rows, `[[`, "vec"))
rownames(feat_mat) <- labels_gd_lri
feat_mat[is.na(feat_mat)] <- 0
message("Feature matrix: ", nrow(feat_mat), " rows × ", ncol(feat_mat), " LRI columns")

# ── 5. UMAP ───────────────────────────────────────────────────────────────────
set.seed(SEED)
k_eff_lri   <- min(K_LRI_UMAP, nrow(feat_mat) - 1)
umap_lri    <- uwot::umap(feat_mat,
                           n_neighbors = k_eff_lri,
                           min_dist    = MINDIST_LRI,
                           metric      = "cosine")

umap_lri_df <- data.frame(
  UMAP1   = umap_lri[, 1],
  UMAP2   = umap_lri[, 2],
  Label   = labels_gd_lri,
  Gene    = genes_gd_lri,
  Dataset = datasets_gd_lri,
  Cancer  = cancer_gd_lri,
  stringsAsFactors = FALSE
)

# ── 6. Plot ───────────────────────────────────────────────────────────────────
lri_umap_title_suffix <- if (LRI_UMAP_CANCER_FILTER == "All") "All Cancer Types" else
                           paste0(LRI_UMAP_CANCER_FILTER, " Only")
lri_umap_out_tag      <- tolower(LRI_UMAP_CANCER_FILTER)   # "all" / "breast" / "lung"

p_lri_umap <- ggplot(umap_lri_df,
                     aes(x = UMAP1, y = UMAP2,
                         colour = Dataset, shape = Cancer)) +
  geom_point(size = 1.8, alpha = 0.5) +
  ggrepel::geom_text_repel(aes(label = Gene),
    size = 2.5, max.overlaps = Inf,
    box.padding = 0.3, segment.size = 0.2, segment.alpha = 0.4,
    show.legend = FALSE) +
  scale_colour_manual(values = LRI_UMAP_COLORS) +
  scale_shape_manual(values = cancer_shapes) +
  theme_classic(base_size = 13) +
  labs(
    title    = paste0("Gene × Dataset UMAP — LRI LogFC Space (",
                      lri_umap_title_suffix, ")"),
    subtitle = sprintf("Each point = Gene × Dataset | %d union LRIs | cosine distance",
                       length(all_lris)),
    colour = "Dataset", shape = "Cancer Type", x = "UMAP 1", y = "UMAP 2"
  )

print(p_lri_umap)
ggsave(file.path(OUTPUT_DIR, sprintf("gene_dataset_lri_umap_%s.png", lri_umap_out_tag)),
       plot = p_lri_umap, width = 10, height = 7, dpi = 300)
message("Gene × Dataset LRI UMAP saved (", LRI_UMAP_CANCER_FILTER, ")")

} # end for (LRI_UMAP_CANCER_FILTER in c("All", "Breast", "Lung"))
# NOTE: umap_lri_df / feat_mat below reflect the last iteration ("Lung").

# Expose for inspection
# View(umap_lri_df)
# feat_mat  — raw Gene × Dataset LogFC matrix (imputed NAs = 0)
```



# Post Analysis — Highly Variable CCIs (CCI logFC space)
```{r post-analysis-hvg-ccis}
# ── CONFIG ────────────────────────────────────────────────────────────────────
TOP_VAR_CCI_PCT  <- 0.1   # keep top X% CCIs by cross Gene×Dataset variance
MIN_OBS_PER_CCI  <- 10     # minimum non-NA rows per CCI to enter HVG ranking

CCI_POST_CANCER_FILTER_OPTIONS <- c("H&N") # c("All", "Breast", "Lung", "H&N")
for (CCI_POST_CANCER_FILTER in CCI_POST_CANCER_FILTER_OPTIONS) {
message("\n===== CCI_POST_CANCER_FILTER: ", CCI_POST_CANCER_FILTER, " =====")

.cci_all_datasets <- list(
  Kurten_HNSC = Kurten_HNSC.malignant,
  Puram_HNSC  = Puram_HNSC.malignant,
  Choi_HNSC   = Choi_HNSC.malignant,
  Bill_HNSC   = Bill_HNSC.malignant
)
.cci_all_cancer <- c(
  Kurten_HNSC = "H&N",
  Puram_HNSC  = "H&N",
  Choi_HNSC   = "H&N",
  Bill_HNSC   = "H&N"
)

if (!CCI_POST_CANCER_FILTER %in% c("All", "Breast", "Lung", "H&N"))
  stop('CCI_POST_CANCER_FILTER must be one of: "All", "Breast", "Lung", "H&N"')
{
  keep_ds <- if (CCI_POST_CANCER_FILTER == "All") {
    names(.cci_all_datasets)
  } else {
    names(.cci_all_datasets)[.cci_all_cancer[names(.cci_all_datasets)] == CCI_POST_CANCER_FILTER]
  }
  CCI_POST_DATASETS <- .cci_all_datasets[keep_ds]
  CCI_POST_CANCER   <- .cci_all_cancer[keep_ds]
}
message("CCI post-analysis datasets (", CCI_POST_CANCER_FILTER, "): ",
        paste(names(CCI_POST_DATASETS), collapse = ", "))

# ── 1. Per-dataset CCI × Gene logFC matrices ──────────────────────────────────
cci_mats <- lapply(CCI_POST_DATASETS, build_cci_logfc_mat)
message("Union CCI space (before HVG): ", length(unique(unlist(lapply(cci_mats, rownames)))), " CCIs")

# ── 2. Cross-cohort Gene×Dataset × CCI matrix (NA = absent) ───────────────────
cci_feat <- build_gene_dataset_cci_feat_mat(cci_mats, CCI_POST_CANCER)
feat_mat_cci      <- cci_feat$feat_mat
all_ccis_union    <- cci_feat$all_ccis
labels_gd_cci     <- cci_feat$labels_gd
genes_gd_cci      <- cci_feat$genes_gd
datasets_gd_cci   <- cci_feat$datasets_gd
cancer_gd_cci     <- cci_feat$cancer_gd
message("Gene×Dataset × CCI matrix: ", nrow(feat_mat_cci), " rows × ",
        ncol(feat_mat_cci), " CCI columns")

# ── 3. HVG selection (variance on non-NA values only) ───────────────────────
n_obs_per_cci <- colSums(!is.na(feat_mat_cci))
cci_var <- apply(feat_mat_cci, 2, function(x) var(x, na.rm = TRUE))
cci_var[n_obs_per_cci < MIN_OBS_PER_CCI] <- NA_real_

eligible <- !is.na(cci_var)
if (!any(eligible)) {
  stop("No CCIs passed MIN_OBS_PER_CCI filter for HVG selection.")
}

var_cutoff <- quantile(cci_var[eligible], probs = 1 - TOP_VAR_CCI_PCT, na.rm = TRUE)
top_var_ccis <- names(cci_var)[eligible & cci_var >= var_cutoff]
feat_mat_hvg <- feat_mat_cci[, top_var_ccis, drop = FALSE]

message(sprintf(
  "HVG CCIs: %d / %d eligible (top %.0f%% by variance, cutoff = %.4f)",
  length(top_var_ccis), sum(eligible), TOP_VAR_CCI_PCT * 100, var_cutoff
))

cci_var_df <- data.frame(
  CCI = names(cci_var),
  variance = cci_var,
  n_obs = n_obs_per_cci,
  is_hvg = names(cci_var) %in% top_var_ccis,
  stringsAsFactors = FALSE
) %>%
  arrange(desc(variance))

cci_post_out_tag <- tolower(CCI_POST_CANCER_FILTER)
write.csv(
  cci_var_df,
  file.path(OUTPUT_DIR, sprintf("cci_hvg_variance_ranking_%s.csv", cci_post_out_tag)),
  row.names = FALSE
)
message("CCI variance table saved.")

} # end for (CCI_POST_CANCER_FILTER)
# NOTE: feat_mat_cci, feat_mat_hvg, top_var_ccis reflect the last filter iteration.
# Run post-analysis-limma-batch next (uses objects from the last iteration above).
```


# Post Analysis — limma batch correction on CCI logFC matrix
```{r post-analysis-limma-batch}
library(limma)

# ── CONFIG ────────────────────────────────────────────────────────────────────
K_CCI_UMAP  <- 15
MINDIST_CCI <- 0.1

# Re-run full CCI post pipeline per cancer filter (HVG + limma + UMAPs).
CCI_POST_CANCER_FILTER_OPTIONS <- c("H&N") # c("All", "Breast", "Lung", "H&N")
TOP_VAR_CCI_PCT  <- 0.05
MIN_OBS_PER_CCI  <- 10

impute_col_mean <- function(mat) {
  mat_imp <- mat
  for (j in seq_len(ncol(mat_imp))) {
    na_idx <- is.na(mat_imp[, j])
    if (any(na_idx)) {
      col_mean <- mean(mat_imp[, j], na.rm = TRUE)
      if (!is.finite(col_mean)) col_mean <- 0
      mat_imp[na_idx, j] <- col_mean
    }
  }
  mat_imp
}

plot_gene_dataset_cci_umap <- function(feat_mat_input, labels, genes, datasets, cancer,
                                      colors, title, subtitle, out_file) {
  mat <- feat_mat_input
  mat[is.na(mat)] <- 0
  if (nrow(mat) < 3L || ncol(mat) < 2L) {
    message("Skipping UMAP (", out_file, "): insufficient dimensions.")
    return(invisible(NULL))
  }
  set.seed(SEED)
  k_eff <- min(K_CCI_UMAP, nrow(mat) - 1L)
  umap_coords <- uwot::umap(mat,
                            n_neighbors = k_eff,
                            min_dist    = MINDIST_CCI,
                            metric      = "cosine")
  umap_df <- data.frame(
    UMAP1   = umap_coords[, 1],
    UMAP2   = umap_coords[, 2],
    Label   = labels,
    Gene    = genes,
    Dataset = datasets,
    Cancer  = cancer,
    stringsAsFactors = FALSE
  )
  p <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = Dataset, shape = Cancer)) +
    geom_point(size = 1.8, alpha = 0.5) +
    ggrepel::geom_text_repel(aes(label = Gene),
      size = 2.5, max.overlaps = Inf,
      box.padding = 0.3, segment.size = 0.2, segment.alpha = 0.4,
      show.legend = FALSE) +
    scale_colour_manual(values = colors) +
    scale_shape_manual(values = cancer_shapes) +
    theme_classic(base_size = 13) +
    labs(title = title, subtitle = subtitle,
         colour = "Dataset", shape = "Cancer Type", x = "UMAP 1", y = "UMAP 2")
  print(p)
  ggsave(out_file, plot = p, width = 14, height = 10, dpi = 300)
  message("UMAP saved to ", out_file)
  invisible(umap_df)
}

.cci_post_colors <- c(
  Kurten_HNSC = "#F781BF",
  Puram_HNSC  = "#999999",
  Choi_HNSC   = "#66C2A5",
  Bill_HNSC   = "#FC8D62"
)
.cci_all_datasets <- list(
  Kurten_HNSC = Kurten_HNSC.malignant,
  Puram_HNSC  = Puram_HNSC.malignant,
  Choi_HNSC   = Choi_HNSC.malignant,
  Bill_HNSC   = Bill_HNSC.malignant
)
.cci_all_cancer <- c(
  Kurten_HNSC = "H&N",
  Puram_HNSC  = "H&N",
  Choi_HNSC   = "H&N",
  Bill_HNSC   = "H&N"
)

for (CCI_POST_CANCER_FILTER in CCI_POST_CANCER_FILTER_OPTIONS) {
message("\n===== limma batch correction: ", CCI_POST_CANCER_FILTER, " =====")

if (!CCI_POST_CANCER_FILTER %in% c("All", "Breast", "Lung", "H&N"))
  stop('CCI_POST_CANCER_FILTER must be one of: "All", "Breast", "Lung", "H&N"')
{
  keep_ds <- if (CCI_POST_CANCER_FILTER == "All") {
    names(.cci_all_datasets)
  } else {
    names(.cci_all_datasets)[.cci_all_cancer[names(.cci_all_datasets)] == CCI_POST_CANCER_FILTER]
  }
  CCI_POST_DATASETS <- .cci_all_datasets[keep_ds]
  CCI_POST_CANCER   <- .cci_all_cancer[keep_ds]
}
cci_mats <- lapply(CCI_POST_DATASETS, build_cci_logfc_mat)
cci_feat <- build_gene_dataset_cci_feat_mat(cci_mats, CCI_POST_CANCER)
feat_mat_cci    <- cci_feat$feat_mat
labels_gd_cci   <- cci_feat$labels_gd
genes_gd_cci    <- cci_feat$genes_gd
datasets_gd_cci <- cci_feat$datasets_gd
cancer_gd_cci   <- cci_feat$cancer_gd

n_obs_per_cci <- colSums(!is.na(feat_mat_cci))
cci_var <- apply(feat_mat_cci, 2, function(x) var(x, na.rm = TRUE))
cci_var[n_obs_per_cci < MIN_OBS_PER_CCI] <- NA_real_
eligible <- !is.na(cci_var)
var_cutoff <- quantile(cci_var[eligible], probs = 1 - TOP_VAR_CCI_PCT, na.rm = TRUE)
top_var_ccis <- names(cci_var)[eligible & cci_var >= var_cutoff]
feat_mat_hvg <- feat_mat_cci[, top_var_ccis, drop = FALSE]
message(sprintf("HVG CCIs for limma: %d / %d eligible", length(top_var_ccis), sum(eligible)))

cci_post_out_tag <- tolower(CCI_POST_CANCER_FILTER)
cci_umap_title_suffix <- if (CCI_POST_CANCER_FILTER == "All") {
  "All Cancer Types"
} else {
  paste0(CCI_POST_CANCER_FILTER, " Only")
}
cci_umap_colors <- .cci_post_colors[unique(datasets_gd_cci)]

# ── 1. limma::removeBatchEffect (column-mean imputation for NA only) ──────────
feat_mat_hvg_na <- feat_mat_hvg
feat_mat_hvg_imp <- impute_col_mean(feat_mat_hvg)

y <- t(feat_mat_hvg_imp)
batch <- factor(datasets_gd_cci, levels = unique(datasets_gd_cci))
mod <- model.matrix(~ 0 + factor(genes_gd_cci))
colnames(mod) <- make.names(colnames(mod))

corrected_y <- tryCatch(
  limma::removeBatchEffect(y, batch = batch, mod = mod),
  error = function(e) {
    message("removeBatchEffect with gene mod failed: ", conditionMessage(e),
            " — falling back to mod = NULL")
    limma::removeBatchEffect(y, batch = batch)
  }
)
feat_mat_cci_corrected <- t(corrected_y)
rownames(feat_mat_cci_corrected) <- labels_gd_cci
message("Batch-corrected matrix: ", nrow(feat_mat_cci_corrected), " × ",
        ncol(feat_mat_cci_corrected))

saveRDS(
  list(
    feat_mat_cci           = feat_mat_cci,
    feat_mat_hvg           = feat_mat_hvg,
    feat_mat_hvg_na        = feat_mat_hvg_na,
    feat_mat_cci_corrected = feat_mat_cci_corrected,
    top_var_ccis           = top_var_ccis,
    batch                  = batch,
    labels_gd              = labels_gd_cci,
    genes_gd               = genes_gd_cci,
    datasets_gd            = datasets_gd_cci
  ),
  file.path(OUTPUT_DIR, sprintf("cci_post_analysis_matrices_%s.rds", cci_post_out_tag))
)
message("Post-analysis matrices saved to RDS.")

# ── 2. Validation UMAPs: raw / HVG / batch-corrected ────────────────────────
n_cci_raw <- ncol(feat_mat_cci)
n_cci_hvg <- ncol(feat_mat_hvg)

plot_gene_dataset_cci_umap(
  feat_mat_cci, labels_gd_cci, genes_gd_cci, datasets_gd_cci, cancer_gd_cci,
  cci_umap_colors,
  title    = paste0("Gene × Dataset UMAP — CCI LogFC (", cci_umap_title_suffix, ")"),
  subtitle = sprintf("Raw | %d union CCIs | NA imputed as 0 for UMAP only", n_cci_raw),
  out_file = file.path(OUTPUT_DIR, sprintf("gene_dataset_cci_umap_raw_%s.png", cci_post_out_tag))
)

plot_gene_dataset_cci_umap(
  feat_mat_hvg, labels_gd_cci, genes_gd_cci, datasets_gd_cci, cancer_gd_cci,
  cci_umap_colors,
  title    = paste0("Gene × Dataset UMAP — HVG CCIs (", cci_umap_title_suffix, ")"),
  subtitle = sprintf("Top %.0f%% variable CCIs (%d features) | NA imputed as 0 for UMAP",
                     TOP_VAR_CCI_PCT * 100, n_cci_hvg),
  out_file = file.path(OUTPUT_DIR, sprintf("gene_dataset_cci_umap_hvg_%s.png", cci_post_out_tag))
)

plot_gene_dataset_cci_umap(
  feat_mat_cci_corrected, labels_gd_cci, genes_gd_cci, datasets_gd_cci, cancer_gd_cci,
  cci_umap_colors,
  title    = paste0("Gene × Dataset UMAP — Batch-Corrected CCI LogFC (",
                    cci_umap_title_suffix, ")"),
  subtitle = sprintf("HVG (%d CCIs) + limma::removeBatchEffect(batch=dataset) | NA→0 for UMAP",
                     n_cci_hvg),
  out_file = file.path(OUTPUT_DIR, sprintf("gene_dataset_cci_umap_corrected_%s.png",
                                          cci_post_out_tag))
)

} # end for (CCI_POST_CANCER_FILTER)

# Expose for inspection
# View(feat_mat_cci)
# View(feat_mat_hvg)
# View(feat_mat_cci_corrected)
# top_var_ccis
```
