---
title: "scDiffCom-Preprocessing-Checks"
output: html_document
---

# Load Seurats
```{r}
Load.Seurat.With.Patients.Data <- function(ds_path, ps_path, gene_col, patient_col_in_seurat) {
     message("Loading raw Seurat object...")
    ext <- tolower(tools::file_ext(ds_path))
    if (ext == "rds") {
      obj <- readRDS(ds_path)
    } else if (ext %in% c("rdata", "rda")) {
      miceadds::load.Rdata(ds_path, "obj")
    } else {
      stop(paste("Unsupported file format:", ext, ". Supported formats: .rds, .RData, .rda"))
    }
    
    message("Loading patient summary (preprocess output)...")
    ps <- readRDS(ps_path) 
    
    if (!patient_col_in_seurat %in% colnames(obj@meta.data)) {
      stop(sprintf("Error: Column '%s' not found in Seurat metadata.", patient_col_in_seurat))
    }
    
    message(sprintf("Merging %s classifications...", gene_col))
    
    # Map patient IDs to their HIGH/LOW/MID status
    mapping_vec <- setNames(ps[[gene_col]], ps$patient_id)
    
    # Apply mapping to the Seurat metadata
    obj@meta.data[[gene_col]] <- mapping_vec[as.character(obj@meta.data[[patient_col_in_seurat]])]
    
    return(list(
      seurat = obj, 
      patient_summary = as.data.frame(ps)
    ))
}
```

# Load Data
```{r}
# Qian.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Qian2020_Breast_step4.RDS",
#                                             ps_path = "~/outputs/Qian2020_Breast_step4/AXL_Qian2020_Breast_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")
# 
# Chan.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Chan2021_Lung_step4.RDS",
#                                             ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Chan2021_Lung_step4/AXL_Chan2021_Lung_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")
# 
# Laughney.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Laughney2020_Lung_step4.RDS",
#                                             ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Laughney2020_Lung_step4/AXL_Laughney2020_Lung_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")
# 
# Bassez.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Bassez2021_Breast_step4.RDS",
#                                             ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Bassez2021_Breast_step4/AXL_Bassez2021_Breast_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")

Bill_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Bill_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Bill_HNSC/AXL_Bill_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")
Choi_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Choi_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Choi_HNSC/AXL_Choi_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")
Kurten_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Kurten_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Kurten_HNSC/AXL_Kurten_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")
Puram_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Puram_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Puram_HNSC/AXL_Puram_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")

```

# Check Seurat object Cell-Type-Distribution
```{r}

library(dplyr)
library(ggplot2)
library(Seurat)

plot_patient_composition <- function(obj, patient_summary, gene_col_exp_name, ds_name,
                                     cell_type_column = "Consensus_Cell_Type",
                                     patient_col_in_seurat = "ident",
                                     use_proportions = FALSE) {

  gene_col <- grep("_EXP$", colnames(patient_summary), value = TRUE)

  # Pull metadata, extract patient column
  meta <- obj@meta.data
  meta$patient <- as.character(meta[[patient_col_in_seurat]])
  meta <- meta[!is.na(meta$patient), ]

  # Order all patients by mean_expr descending; patients not in summary sort last
  all_patients <- unique(meta$patient)
  ordered_patients <- data.frame(patient = all_patients, stringsAsFactors = FALSE) %>%
    left_join(patient_summary[, c("patient_id", "mean_expr")],
              by = c("patient" = "patient_id")) %>%
    arrange(desc(mean_expr)) %>%
    pull(patient)

  # Build x-axis labels: "patient\nHIGH/MID/LOW" for labeled patients, else just patient
  label_map <- setNames(patient_summary[[gene_col]], patient_summary$patient_id)
  x_labels <- setNames(
    ifelse(ordered_patients %in% names(label_map),
           paste0(ordered_patients, "\n", label_map[ordered_patients]),
           ordered_patients),
    ordered_patients
  )

  cell_counts <- meta %>%
    dplyr::filter(!is.na(.data[[cell_type_column]])) %>%
    # Rename the dynamic column to "cell_type" to safely pass it down the pipe
    dplyr::rename(cell_type = dplyr::all_of(cell_type_column)) %>%
    # Explicitly use dplyr::count to prevent package masking
    dplyr::count(patient, cell_type) %>%
    dplyr::group_by(patient) %>%
    dplyr::mutate(prop = n / sum(n)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(cell_type = factor(cell_type,
                                     levels = c("Epithelial",
                                                    sort(setdiff(unique(cell_type), "Epithelial")))))

  y_var   <- if (use_proportions) "prop" else "n"
  y_label <- if (use_proportions) "Proportion" else "Cell Count"

  p <- ggplot(cell_counts, aes(x = patient, y = .data[[y_var]], fill = cell_type)) +
    geom_bar(stat = "identity")

  if (!use_proportions)
    p <- p + geom_hline(yintercept = 50, linetype = "dashed", alpha = 0.4) +
             scale_y_continuous(breaks = function(x) sort(unique(c(pretty(x), 50))))

  p +
    scale_x_discrete(limits = ordered_patients, labels = x_labels) +
    labs(x = "Patient", y = y_label,
         title = paste(ds_name, "Cell Type Distribution:", gene_col_exp_name),
         fill = "Cell Type") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.ticks.x = element_blank(),
      legend.title = element_text(size = 10),
      panel.grid = element_blank(),
      panel.background = element_blank()
    )
}

# Qian.AXL.plot <- plot_patient_composition(obj = Qian.AXL$seurat,
#                                            patient_summary =Qian.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Qian2020 - Breast",
#                                            patient_col_in_seurat = "ident",
#                                           cell_type_column = "cell_type")
# 
# Chan.AXL.plot <- plot_patient_composition(obj = Chan.AXL$seurat,
#                                            patient_summary =Chan.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Chan2021 - Lung",
#                                            patient_col_in_seurat = "ident",
#                                           cell_type_column = "cell_type")
# 
# Laughney.AXL.plot <- plot_patient_composition(obj = Laughney.AXL$seurat,
#                                            patient_summary =Laughney.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Laughney2020 - Lung",
#                                            patient_col_in_seurat = "ident",
#                                            cell_type_column = "cell_type")
# 
# Bassez.AXL.plot <- plot_patient_composition(obj = Bassez.AXL$seurat,
#                                            patient_summary =Bassez.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Bassez2021 - Breast",
#                                            patient_col_in_seurat = "ident",
#                                            cell_type_column = "cell_type",
#                                            use_proportions = TRUE)

Kurten_HNSC.AXL.plot <- plot_patient_composition(obj = Kurten_HNSC.AXL$seurat,
                                           patient_summary =Kurten_HNSC.AXL$patient_summary,
                                           gene_col_exp_name = "AXL_EXP",
                                           ds_name = "Kurten - H&N",
                                           patient_col_in_seurat = "Patient",
                                           cell_type_column = "Cell_Type",
                                           use_proportions = TRUE)

out_dir <- "~/Thesis/CCC/outputs/scDiffCom/QC"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ggsave(file.path(out_dir, "Qian2020_Breast_AXL_cell_type_distribution_original_cell_type.png"),    Qian.AXL.plot,    width = 20, height = 6)
# ggsave(file.path(out_dir, "Chan2021_Lung_AXL_cell_type_distribution.png"),      Chan.AXL.plot,    width = 20, height = 6)
# ggsave(file.path(out_dir, "Laughney2020_Lung_AXL_cell_type_distribution.png"),  Laughney.AXL.plot, width = 20, height = 6)
# ggsave(file.path(out_dir, "Bassez2021_Breast_AXL_cell_type_distribution.png"),  Bassez.AXL.plot,  width = 20, height = 6)
ggsave(file.path(out_dir, "Kurten_HNSC_AXL_cell_Type_distribution.png"), Kurten_HNSC.AXL.plot, width = 20, height = 6)
```


# Merge + ComBat Batch Correction + UMAP Integration
```{r merge-combat-umap, message=FALSE, warning=FALSE}
# Dependencies ─────────────────────────────────────────────────────────────────
if (!requireNamespace("sva",       quietly = TRUE)) stop("BiocManager::install('sva')")
if (!requireNamespace("patchwork", quietly = TRUE)) stop("install.packages('patchwork')")
if (!requireNamespace("Matrix",    quietly = TRUE)) stop("install.packages('Matrix')")

library(Seurat)
library(sva)
library(ggplot2)
library(dplyr)
library(patchwork)
library(Matrix)

# Config ───────────────────────────────────────────────────────────────────────
SEURAT_DIR <- path.expand("~/scObjects")
OUTPUT_DIR <- path.expand("~/Thesis/CCC/outputs/scDiffCom/integration")
N_FEATURES <- 3000
N_PCS      <- 30
CT_COL     <- "Cell_Type"
SEURAT_V5  <- utils::packageVersion("Seurat") >= "5.0.0"
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# ── 1. Load objects and stamp source label ────────────────────────────────────
stamp_source <- function(fname, label) {
   message("Loading raw Seurat object...")
    path <- file.path(SEURAT_DIR, fname)
    ext <- tolower(tools::file_ext(path))
    if (ext == "rds") { 
      obj <- readRDS(path)
    } else if (ext %in% c("rdata", "rda")) {
      miceadds::load.Rdata(path, "obj")
    } else {
      stop(paste("Unsupported file format:", ext, ". Supported formats: .rds, .RData, .rda"))
    }
  obj$source <- label
  message("Loaded ", fname, " — ", ncol(obj), " cells")
  obj
}

HNSC.Atlas <- stamp_source("HNSCC.Atlas.RData", "HNSC_Atlas")
puram <- stamp_source("Puram_HNSC.RData", "Puram_HNSC")
choi   <- stamp_source("Choi_HNSC.RData",    "Choi_HNSC")
kurten <- stamp_source("Kurten_HNSC.RData",  "Kurten_HNSC")
bill   <- stamp_source("Bill_HNSC.RData",    "Bill_HNSC")

# ── 2. Individual UMAPs (runs full Seurat pipeline if UMAP absent) ────────────
ensure_umap <- function(obj, label) {
  if ("umap" %in% names(obj@reductions)) {
    message(label, ": UMAP already present — skipping preprocessing.")
    return(obj)
  }
  message(label, ": no UMAP found — running NormalizeData → PCA → UMAP ...")
  obj <- NormalizeData(obj, verbose = FALSE)
  obj <- FindVariableFeatures(obj, nfeatures = N_FEATURES, verbose = FALSE)
  obj <- ScaleData(obj, verbose = FALSE)
  obj <- RunPCA(obj, npcs = N_PCS, verbose = FALSE)
  obj <- FindNeighbors(obj, dims = 1:N_PCS, verbose = FALSE)
  obj <- RunUMAP(obj, dims = 1:N_PCS, verbose = FALSE)
  obj
}
puram  <- ensure_umap(puram,  "Puram_HNSC")
choi   <- ensure_umap(choi,   "Choi_HNSC")
kurten <- ensure_umap(kurten, "Kurten_HNSC")
bill   <- ensure_umap(bill,   "Bill_HNSC")

ind_umap_plot <- function(obj, label) {
  col <- if (CT_COL %in% colnames(obj@meta.data)) CT_COL else "seurat_clusters"
  DimPlot(obj, reduction = "umap", group.by = col, label = TRUE, repel = TRUE,
          pt.size = 0.3) +
    ggtitle(label) +
    theme(plot.title      = element_text(face = "bold", size = 11),
          legend.text     = element_text(size = 7),
          legend.key.size = unit(0.35, "cm"))
}
p_ind <- ((ind_umap_plot(puram,  "Puram_HNSC")  |
           ind_umap_plot(choi,   "Choi_HNSC")) /
          (ind_umap_plot(kurten, "Kurten_HNSC") |
           ind_umap_plot(bill,   "Bill_HNSC"))) +
  plot_annotation(
    title = "Individual UMAPs — H&N Datasets",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  )
print(p_ind)
ggsave(file.path(OUTPUT_DIR, "01_individual_umaps.png"),
       p_ind, width = 18, height = 14, dpi = 300)
message("Saved 01_individual_umaps.png")

# ── 3. Merge ──────────────────────────────────────────────────────────────────
message("\nMerging Puram_HNSC, Choi_HNSC, Kurten_HNSC, Bill_HNSC ...")
merged <- merge(puram, y = list(choi, kurten, bill),
                add.cell.ids = c("Puram_HNSC", "Choi_HNSC", "Kurten_HNSC", "Bill_HNSC"))
message("Merged object: ", ncol(merged), " cells")

# ── 4. Normalize and select variable features (join layers first for v5) ──────
DefaultAssay(merged) <- "RNA"
if (SEURAT_V5) merged <- JoinLayers(merged)
merged <- NormalizeData(merged, verbose = FALSE)
merged <- FindVariableFeatures(merged, nfeatures = N_FEATURES, verbose = FALSE)
var_genes <- VariableFeatures(merged)
message("Variable features selected: ", length(var_genes))

# ── 5. Extract log-normalised matrix for ComBat ───────────────────────────────
# ComBat operates on log-normalised values; corrected matrix is stored as a
# separate 'ComBat' assay so the original RNA assay is preserved intact.
log_mat <- as.matrix(
  if (SEURAT_V5) GetAssayData(merged, assay = "RNA", layer = "data")[var_genes, ]
  else           GetAssayData(merged, assay = "RNA", slot  = "data")[var_genes, ]
)
batch_vec <- as.character(merged$source)
message("Batch sizes: ",
        paste(paste0(names(table(batch_vec)), "=", table(batch_vec)), collapse = ", "))

# Drop genes with zero variance in any single batch (ComBat cannot handle these)
batch_var <- do.call(cbind, lapply(unique(batch_vec), function(b)
  apply(log_mat[, batch_vec == b, drop = FALSE], 1, var)
))
keep        <- rowSums(batch_var == 0) == 0
var_genes_f <- var_genes[keep]
log_mat_f   <- log_mat[keep, ]
message("Removed ", sum(!keep), " zero-variance gene(s); ",
        sum(keep), " genes retained for ComBat.")

# ── 6. ComBat batch correction ────────────────────────────────────────────────
message("Running sva::ComBat (parametric prior) on ",
        nrow(log_mat_f), " genes x ", ncol(log_mat_f), " cells ...")
combat_mat <- sva::ComBat(dat        = log_mat_f,
                           batch      = batch_vec,
                           par.prior  = TRUE,
                           prior.plots = FALSE)
message("ComBat complete.")

# ── 7. Store corrected matrix as new 'ComBat' assay ───────────────────────────
sparse_cb          <- Matrix::Matrix(combat_mat, sparse = TRUE)
merged[["ComBat"]] <- tryCatch(
  CreateAssay5Object(data = sparse_cb),          # Seurat >= 5
  error = function(e) CreateAssayObject(data = sparse_cb)   # Seurat v4 fallback
)
DefaultAssay(merged)     <- "ComBat"
VariableFeatures(merged) <- var_genes_f

# ── 8. Scale → PCA → UMAP on ComBat-corrected assay ──────────────────────────
message("ScaleData → RunPCA → RunUMAP on ComBat assay ...")
merged <- ScaleData(merged,      features       = var_genes_f,       verbose = FALSE)
merged <- RunPCA(merged,         features       = var_genes_f,
                 npcs            = N_PCS,
                 reduction.name  = "pca_combat",                     verbose = FALSE)
merged <- FindNeighbors(merged,  reduction      = "pca_combat",
                         dims   = 1:N_PCS,                           verbose = FALSE)
merged <- FindClusters(merged,                                        verbose = FALSE)
merged <- RunUMAP(merged,        reduction      = "pca_combat",
                  dims           = 1:N_PCS,
                  reduction.name = "umap_combat",                    verbose = FALSE)
message("Dimensionality reduction complete.")

# ── 9. Integrated UMAP — Consensus Cell Type ─────────────────────────────────
ct_col_use <- if (CT_COL %in% colnames(merged@meta.data)) CT_COL else "seurat_clusters"
p_ct <- DimPlot(merged, reduction = "umap_combat", group.by = ct_col_use,
                label = TRUE, repel = TRUE, pt.size = 0.25) +
  ggtitle(paste("Integrated UMAP —", ct_col_use)) +
  theme(plot.title      = element_text(face = "bold"),
        legend.text     = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"))

# ── 10. Integrated UMAP — Source Dataset (batch-correction quality check) ─────
src_pal <- c(Puram_HNSC  = "#1B9E77",
             Choi_HNSC   = "#D95F02",
             Kurten_HNSC = "#7570B3",
             Bill_HNSC   = "#E7298A")
p_src <- DimPlot(merged, reduction = "umap_combat", group.by = "source",
                 cols = src_pal, pt.size = 0.25) +
  ggtitle("Integrated UMAP — Source Dataset") +
  theme(plot.title = element_text(face = "bold"))

p_combined <- (p_ct / p_src) +
  plot_annotation(
    title    = "ComBat-Corrected H&N Integration",
    subtitle = sprintf("%d cells | %d variable features | %d PCs",
                       ncol(merged), length(var_genes_f), N_PCS),
    theme    = theme(
      plot.title    = element_text(size = 14, face = "bold",  hjust = 0.5),
      plot.subtitle = element_text(size = 10, colour = "grey40", hjust = 0.5)
    )
  )

print(p_ct)
print(p_src)
print(p_combined)

ggsave(file.path(OUTPUT_DIR, "02_integrated_umap_celltype.png"),
       p_ct,       width = 12, height = 9,  dpi = 300)
ggsave(file.path(OUTPUT_DIR, "03_integrated_umap_source.png"),
       p_src,      width = 10, height = 8,  dpi = 300)
ggsave(file.path(OUTPUT_DIR, "04_integrated_umap_combined.png"),
       p_combined, width = 12, height = 18, dpi = 300)
message("Saved integrated UMAP plots (02, 03, 04).")

# ── 11. Save final ComBat-corrected Seurat object ─────────────────────────────
out_rds <- file.path(OUTPUT_DIR, "HNSC_ComBat_integrated.rds")
saveRDS(merged, out_rds)
message("\nFinal object saved to: ", out_rds)
message("Assays     : ", paste(Assays(merged),              collapse = ", "))
message("Reductions : ", paste(names(merged@reductions),    collapse = ", "))
message("Cells      : ", ncol(merged))
message("Metadata   : ", paste(colnames(merged@meta.data),  collapse = ", "))
```



#Check patient distribution
```{r}
library(tidyr)
# opt$column_id <- "ident_source"
merged.results <- helpers$add_malignant_gene_group(seurat_obj = HNSC.Atlas,
                                            gene = "CTLA4",
                                            cell_type_col = "Cell_Type",
                                            assay="RNA")

merged.results.ps <- merged.results$patient_summary

# This creates a 2D grid of counts
counts_table <- table(merged.results.ps$CTLA4_EXP, merged.results.ps$Project)
print(counts_table)

```



# Data validation: expression category distribution per patient

```{r gene-expression-category-distribution, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

# For each dataset, calculate and plot the distribution of HIGH/MID/LOW calls per patient
# across all analyzed genes.
BASE_RESULTS_DIR <- path.expand("~/CCC-PreProcess/results-patients-zscore/Patient-ZScore")


dataset_names <- c("Puram_HNSC", "Choi_HNSC", "Kurten_HNSC", "Bill_HNSC")

if (length(dataset_names) == 0) {
  message("No dataset names found (no *.AXL objects, and results dir missing). Skipping.")
} else {
  expr_levels <- c("LOW", "MID", "HIGH")
  expr_pal <- c(LOW = "#4E79A7", MID = "#F28E2B", HIGH = "#E15759")

  read_grouped_file <- function(rds_path, dataset_name) {
    df <- tryCatch(readRDS(rds_path), error = function(e) NULL)
    if (is.null(df) || !is.data.frame(df) || ncol(df) != 3) return(NULL)

    # Expect columns: patient_id, mean_expr, {gene}_exp (dynamic)
    exp_col <- colnames(df)[3]
    gene <- sub(paste0("_", dataset_name, "_grouped\\.rds$"), "", basename(rds_path))

    df %>%
      dplyr::select(patient_id, dplyr::all_of(exp_col)) %>%
      dplyr::rename(expression_level = dplyr::all_of(exp_col)) %>%
      dplyr::mutate(
        gene = gene,
        expression_level = as.character(expression_level)
      )
  }

  for (ds in dataset_names) {
    ds_dir <- file.path(BASE_RESULTS_DIR, ds)
    if (!dir.exists(ds_dir)) {
      message("Dataset directory not found: ", ds_dir, " (skipping)")
      next
    }

    rds_files <- list.files(
      ds_dir,
      pattern = paste0("_", ds, "_grouped\\.rds$"),
      full.names = TRUE
    )

    if (length(rds_files) == 0) {
      message("No grouped .rds files found for dataset: ", ds, " (skipping)")
      next
    }

    agg <- rds_files %>%
      lapply(read_grouped_file, dataset_name = ds) %>%
      Filter(Negate(is.null), .) %>%
      bind_rows()

    if (nrow(agg) == 0) {
      message("No readable/valid files for dataset: ", ds, " (skipping)")
      next
    }

    count_df <- agg %>%
      dplyr::filter(!is.na(patient_id), !is.na(expression_level)) %>%
      dplyr::mutate(
        patient_id = as.character(patient_id),
        expression_level = factor(expression_level, levels = expr_levels)
      ) %>%
      dplyr::count(patient_id, expression_level, name = "n") %>%
      tidyr::complete(patient_id, expression_level, fill = list(n = 0))

    p <- ggplot(count_df, aes(x = patient_id, y = n, fill = expression_level)) +
      geom_col(
        position = position_dodge2(width = 0.9, preserve = "single"),
        width = 0.8,
        color = "white",
        linewidth = 0.15
      ) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
      scale_fill_manual(values = expr_pal, drop = FALSE) +
      labs(
        title = paste0(ds, ": Gene expression category distribution per patient (Patients Z-Score)"),
        x = "patient_id",
        y = "Number of analyzed genes",
        fill = "Expression\ncategory"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.major.x = element_blank()
      )

    print(p)

    qc_plot_dir <- path.expand("~/Thesis/CCC/outputs/plots/QC")
    dir.create(qc_plot_dir, recursive = TRUE, showWarnings = FALSE)
    ggsave(
      filename = file.path(qc_plot_dir, paste0(ds, "_zscore_expression_category_distribution_per_patient.png")),
      plot = p,
      width = 10,
      height = 6,
      dpi = 300
    )
  }
}
```

# Gene dispersion: pseudobulk expression vs patient splits

Compare per-gene cross-patient dispersion in **continuous pseudobulk expression** vs **discrete HIGH/LOW/MID** patient classifications.

- **Expression metric:** row-wise SD of raw pseudobulk values (not SD of row z-scores, which is ~1 by construction).
- **Split metric:** row-wise SD of numeric-encoded labels (`LOW=1`, `MID=2`, `HIGH=3`).

`SPLIT_SOURCE` selects which classification pipeline to use (or both):

| Source | Meaning of split-space SD |
|--------|---------------------------|
| `patient_zscore` | Tertiles from row-wise z-scores of expression ([scDiffCom-Preprocess-PatientZScore.R](scDiffCom-Preprocess-PatientZScore.R)) — measures compression from tertiling expression |
| `rankgenes` | Tertiles from within-patient ranks ([scDiffCom-Preprocess-RankGenes.R](scDiffCom-Preprocess-RankGenes.R)) — reflects patient-specific ranking, not expression variance alone |

```{r gene-dispersion-expr-vs-split, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

# ── Config ────────────────────────────────────────────────────────────────────
SPLIT_SOURCE <- "both"   # "patient_zscore" | "rankgenes" | "both"
DATASETS <- c("Puram_HNSC", "Choi_HNSC", "Kurten_HNSC", "Bill_HNSC")
PSEUDOBULK_DIR <- path.expand("~/Thesis/CCC/outputs/RData_objects/pseudobulk_matrix")
ZSCORE_SPLITS_DIR <- path.expand("~/CCC-PreProcess/results-patients-zscore/Patient-ZScore")
RANKGENES_DIR <- path.expand("~/CCC-PreProcess/results-RankGenes")
QC_OUT_DIR <- path.expand("~/Thesis/CCC/outputs/plots/QC/gene_dispersion")

split_sources <- switch(
  SPLIT_SOURCE,
  patient_zscore = "patient_zscore",
  rankgenes = "rankgenes",
  both = c("patient_zscore", "rankgenes"),
  stop("SPLIT_SOURCE must be 'patient_zscore', 'rankgenes', or 'both'. Got: ", SPLIT_SOURCE)
)

split_source_labels <- c(
  patient_zscore = "Patient-ZScore tertiles",
  rankgenes = "RankGenes tertiles (within-patient rank)"
)

# ── Shared helpers (rankGenesSplitUtils.R) ────────────────────────────────────
script_dir <- tryCatch(
  dirname(knitr::current_input(dir = TRUE)),
  error = function(e) NULL
)
if (is.null(script_dir) || !nzchar(script_dir)) {
  script_dir <- normalizePath(".", winslash = "/")
}
source(file.path(script_dir, "rankGenesSplitUtils.R"))

LABEL_NUM <- c(LOW = 1, MID = 2, HIGH = 3)

row_dispersion <- function(mat) {
  if (requireNamespace("matrixStats", quietly = TRUE)) {
    return(matrixStats::rowSds(mat, na.rm = TRUE))
  }
  apply(mat, 1, stats::sd, na.rm = TRUE)
}

encode_splits <- function(splits_chr) {
  out <- matrix(NA_real_, nrow = nrow(splits_chr), ncol = ncol(splits_chr),
                dimnames = dimnames(splits_chr))
  for (lev in names(LABEL_NUM)) {
    out[splits_chr == lev] <- LABEL_NUM[[lev]]
  }
  storage.mode(out) <- "double"
  out
}

row_dispersion_expr <- function(mat) {
  row_dispersion(mat)
}

row_dispersion_split <- function(splits_chr) {
  row_dispersion(encode_splits(splits_chr))
}

align_matrices <- function(expr_mat, splits_mat) {
  genes <- intersect(rownames(expr_mat), rownames(splits_mat))
  patients <- intersect(colnames(expr_mat), colnames(splits_mat))
  if (length(genes) == 0L || length(patients) == 0L) {
    return(NULL)
  }
  list(
    expr = expr_mat[genes, patients, drop = FALSE],
    splits = splits_mat[genes, patients, drop = FALSE]
  )
}

compare_dispersion <- function(sd_expr, sd_split) {
  n <- length(sd_expr)
  if (is.null(names(sd_expr))) names(sd_expr) <- seq_len(n)
  valid <- is.finite(sd_expr) & is.finite(sd_split) & sd_expr > 0
  data.frame(
    gene = names(sd_expr)[valid],
    sd_expr = unname(sd_expr[valid]),
    sd_split = unname(sd_split[valid]),
    ratio = unname(sd_split[valid] / sd_expr[valid]),
    stringsAsFactors = FALSE
  )
}

load_splits_from_grouped_dir <- function(base_dir, dataset_name) {
  ds_dir <- file.path(path.expand(base_dir), dataset_name)
  if (!dir.exists(ds_dir)) {
    return(NULL)
  }

  files <- list.files(
    ds_dir,
    pattern = paste0("_", dataset_name, "_grouped\\.rds$"),
    full.names = TRUE
  )
  if (length(files) == 0L) {
    return(NULL)
  }

  patients <- NULL
  split_rows <- list()

  for (f in files) {
    df <- tryCatch(readRDS(f), error = function(e) NULL)
    if (is.null(df) || !is.data.frame(df)) next

    exp_col <- grep("_exp$", colnames(df), ignore.case = TRUE, value = TRUE)[1]
    if (is.na(exp_col)) next

    gene <- sub(paste0("_", dataset_name, "_grouped\\.rds$"), "", basename(f))
    if (is.null(patients)) {
      patients <- as.character(df$patient_id)
    }
    vec <- setNames(as.character(df[[exp_col]]), as.character(df$patient_id))
    split_rows[[gene]] <- vec[patients]
  }

  if (length(split_rows) == 0L) {
    return(NULL)
  }

  genes_found <- names(split_rows)
  splits <- matrix(NA_character_, nrow = length(genes_found), ncol = length(patients),
                   dimnames = list(genes_found, patients))
  for (g in genes_found) {
    splits[g, ] <- split_rows[[g]]
  }
  splits
}

load_splits_rankgenes <- function(dataset_name, rankgenes_dir, expr_mat) {
  ds_dir <- file.path(path.expand(rankgenes_dir), dataset_name)
  all_splits_path <- file.path(ds_dir, paste0(dataset_name, ALL_SPLITS_SUFFIX))

  if (file.exists(all_splits_path)) {
    message("  Loading RankGenes splits: ", all_splits_path)
    return(readRDS(all_splits_path))
  }

  splits <- tryCatch(
    load_splits_from_rankgenes_dir(rankgenes_dir, dataset_name),
    error = function(e) NULL
  )
  if (!is.null(splits)) {
    message("  Loaded RankGenes splits from grouped .rds files in ", ds_dir)
    return(splits)
  }

  message("  RankGenes splits not found; building from pseudobulk (column-wise rank tertiles).")
  rank_mat <- compute_rank_matrix(expr_mat)
  build_patient_splits_matrix(expr_mat, rank_mat = rank_mat)
}

load_splits_for_source <- function(source, dataset_name, expr_mat) {
  switch(
    source,
    patient_zscore = load_splits_from_grouped_dir(ZSCORE_SPLITS_DIR, dataset_name),
    rankgenes = load_splits_rankgenes(dataset_name, RANKGENES_DIR, expr_mat),
    stop("Unknown split source: ", source)
  )
}

plot_dispersion_scatter <- function(cmp_df, ds, source, source_label) {
  n_genes <- nrow(cmp_df)
  pct_split_lt <- 100 * mean(cmp_df$sd_split < cmp_df$sd_expr)
  pct_split_gt <- 100 * mean(cmp_df$sd_split > cmp_df$sd_expr)
  pct_split_zero <- 100 * mean(cmp_df$sd_split == 0)

  subtitle <- sprintf(
    "%d genes | split SD < expr SD: %.1f%% | split SD > expr SD: %.1f%% | split SD = 0: %.1f%%",
    n_genes, pct_split_lt, pct_split_gt, pct_split_zero
  )

  ggplot(cmp_df, aes(x = sd_expr, y = sd_split)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.35, size = 1.2, color = "#4E79A7") +
    scale_x_continuous(trans = "log1p") +
    scale_y_continuous(trans = "log1p") +
    labs(
      title = paste0(ds, ": Gene dispersion — expression vs split"),
      subtitle = subtitle,
      x = "Pseudobulk row SD (expression space)",
      y = "Encoded split row SD (split space)",
      caption = source_label
    ) +
    theme_minimal(base_size = 11) +
    theme(plot.caption = element_text(size = 9, color = "grey40"))
}

run_dispersion_stats <- function(cmp_df) {
  wt <- wilcox.test(cmp_df$sd_split, cmp_df$sd_expr, paired = TRUE, alternative = "two.sided")
  wt_greater <- wilcox.test(cmp_df$sd_split, cmp_df$sd_expr, paired = TRUE, alternative = "greater")
  wt_less <- wilcox.test(cmp_df$sd_split, cmp_df$sd_expr, paired = TRUE, alternative = "less")
  ct <- suppressWarnings(cor.test(cmp_df$sd_expr, cmp_df$sd_split, method = "spearman"))

  list(
    n_genes = nrow(cmp_df),
    median_sd_expr = median(cmp_df$sd_expr),
    median_sd_split = median(cmp_df$sd_split),
    median_ratio = median(cmp_df$ratio),
    wilcox_p = wt$p.value,
    wilcox_p_split_gt_expr = wt_greater$p.value,
    wilcox_p_split_lt_expr = wt_less$p.value,
    spearman_rho = unname(ct$estimate),
    spearman_p = ct$p.value
  )
}

dir.create(QC_OUT_DIR, recursive = TRUE, showWarnings = FALSE)

for (ds in DATASETS) {
  message("\n=== ", ds, " ===")

  pb_path <- tryCatch(
    resolve_pseudobulk_path(ds, PSEUDOBULK_DIR),
    error = function(e) {
      message("  Pseudobulk not found: ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(pb_path)) next

  message("  Loading pseudobulk: ", pb_path)
  expr_mat <- load_gene_patient_matrix(pb_path)

  for (src in split_sources) {
    message("  Split source: ", src, " (", split_source_labels[[src]], ")")

    splits_mat <- load_splits_for_source(src, ds, expr_mat)
    if (is.null(splits_mat)) {
      message("  No splits found for ", src, " — skipping.")
      next
    }

    aligned <- align_matrices(expr_mat, splits_mat)
    if (is.null(aligned)) {
      message("  No overlapping genes/patients between pseudobulk and splits — skipping.")
      next
    }

    message(sprintf(
      "  Aligned: %d genes x %d patients",
      nrow(aligned$expr), ncol(aligned$expr)
    ))

    sd_expr <- row_dispersion_expr(aligned$expr)
    sd_split <- row_dispersion_split(aligned$splits)
    names(sd_expr) <- rownames(aligned$expr)
    names(sd_split) <- rownames(aligned$splits)

    cmp_df <- compare_dispersion(sd_expr, sd_split)
    if (nrow(cmp_df) == 0L) {
      message("  No genes with valid paired dispersion — skipping.")
      next
    }

    stats <- run_dispersion_stats(cmp_df)
    message(sprintf(
      paste0(
        "  n=%d | median SD_expr=%.4f | median SD_split=%.4f | median ratio=%.4f\n",
        "  Wilcoxon (two-sided) p=%.2e | split>expr p=%.2e | split<expr p=%.2e\n",
        "  Spearman rho=%.3f (p=%.2e)"
      ),
      stats$n_genes, stats$median_sd_expr, stats$median_sd_split, stats$median_ratio,
      stats$wilcox_p, stats$wilcox_p_split_gt_expr, stats$wilcox_p_split_lt_expr,
      stats$spearman_rho, stats$spearman_p
    ))

    p <- plot_dispersion_scatter(cmp_df, ds, src, split_source_labels[[src]])
    print(p)

    out_png <- file.path(QC_OUT_DIR, paste0(ds, "_", src, "_expr_vs_split_dispersion.png"))
    ggsave(out_png, p, width = 8, height = 6, dpi = 300)
    message("  Saved: ", out_png)

    out_csv <- file.path(QC_OUT_DIR, paste0(ds, "_", src, "_expr_vs_split_dispersion.csv"))
    write.csv(cmp_df, out_csv, row.names = FALSE)
  }
}
```

# Gene patient-wise profiles: PCA in expression vs split space

Each **gene** is a point in patient space (dimensions = patients). Rows are row z-scored so profiles reflect **relative** highs/lows across patients, not absolute expression.

- **Expression PCA:** which genes have similar pseudobulk patterns across patients.
- **Split PCA:** which genes assign LOW/MID/HIGH to similar patients.
- **Profile correlation** (per gene): correlation between row z-scored expression and row z-scored encoded splits. **Low correlation** → split labels do not track that gene's expression patient pattern (less informative in split space for that gene).

```{r gene-profile-pca-expr-vs-split, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)

# ── Config (same split sources as dispersion section) ─────────────────────────
SPLIT_SOURCE <- "both"
DATASETS <- c("Puram_HNSC", "Choi_HNSC", "Kurten_HNSC", "Bill_HNSC")
PSEUDOBULK_DIR <- path.expand("~/Thesis/CCC/outputs/RData_objects/pseudobulk_matrix")
ZSCORE_SPLITS_DIR <- path.expand("~/CCC-PreProcess/results-patients-zscore/Patient-ZScore")
RANKGENES_DIR <- path.expand("~/CCC-PreProcess/results-RankGenes")
PROFILE_PCA_DIR <- path.expand("~/Thesis/CCC/outputs/plots/QC/gene_profile_pca")
N_PC <- 2L
MIN_PATIENTS <- 5L
MAX_PATIENT_NA_FRAC <- 0.25   # drop patient if >25% of genes are NA (incl. all-NA columns)
LABEL_LOW_COR_QUANTILE <- 0.10   # highlight bottom 10% expr–split profile correlation
N_LABEL_GENES <- 12L

split_sources <- switch(
  SPLIT_SOURCE,
  patient_zscore = "patient_zscore",
  rankgenes = "rankgenes",
  both = c("patient_zscore", "rankgenes"),
  stop("SPLIT_SOURCE must be 'patient_zscore', 'rankgenes', or 'both'.")
)

split_source_labels <- c(
  patient_zscore = "Patient-ZScore tertiles",
  rankgenes = "RankGenes tertiles (within-patient rank)"
)

script_dir <- tryCatch(
  dirname(knitr::current_input(dir = TRUE)),
  error = function(e) NULL
)
if (is.null(script_dir) || !nzchar(script_dir)) {
  script_dir <- normalizePath(".", winslash = "/")
}
source(file.path(script_dir, "rankGenesSplitUtils.R"))

LABEL_NUM <- c(LOW = 1, MID = 2, HIGH = 3)

encode_splits <- function(splits_chr) {
  out <- matrix(NA_real_, nrow = nrow(splits_chr), ncol = ncol(splits_chr),
                dimnames = dimnames(splits_chr))
  for (lev in names(LABEL_NUM)) {
    out[splits_chr == lev] <- LABEL_NUM[[lev]]
  }
  storage.mode(out) <- "double"
  out
}

align_matrices <- function(expr_mat, splits_mat) {
  genes <- intersect(rownames(expr_mat), rownames(splits_mat))
  patients <- intersect(colnames(expr_mat), colnames(splits_mat))
  if (length(genes) == 0L || length(patients) == 0L) return(NULL)
  list(
    expr = expr_mat[genes, patients, drop = FALSE],
    splits = splits_mat[genes, patients, drop = FALSE]
  )
}

row_zscore_matrix <- function(mat) {
  if (requireNamespace("matrixStats", quietly = TRUE)) {
    rm <- matrixStats::rowMeans2(mat, na.rm = TRUE)
    rsd <- matrixStats::rowSds(mat, na.rm = TRUE)
    rsd[!is.finite(rsd) | rsd == 0] <- NA_real_
    z <- sweep(mat, 1, rm, FUN = "-")
    z <- sweep(z, 1, rsd, FUN = "/")
    return(z)
  }
  t(scale(t(mat)))
}

genes_with_usable_rows <- function(mat, min_patients = MIN_PATIENTS) {
  apply(mat, 1, function(r) {
    ok <- is.finite(r)
    sum(ok) >= min_patients && stats::sd(r[ok]) > 0
  })
}

# Drop patients (columns) that are all-NA or mostly missing across genes.
drop_sparse_patients <- function(mat, max_na_frac = MAX_PATIENT_NA_FRAC,
                               min_patients = MIN_PATIENTS, label = "matrix") {
  if (ncol(mat) == 0L) return(mat)

  na_frac <- colMeans(!is.finite(mat))
  keep <- na_frac < max_na_frac
  dropped <- colnames(mat)[!keep]

  if (length(dropped) > 0L) {
    message(sprintf(
      "  Dropped %d patient(s) from %s (NA fraction >= %.0f%%): %s",
      length(dropped), label, 100 * max_na_frac,
      paste(dropped, collapse = ", ")
    ))
  }

  mat <- mat[, keep, drop = FALSE]
  if (ncol(mat) < min_patients) {
    warning(
      "Only ", ncol(mat), " patient(s) left in ", label,
      " after NA filtering (need >= ", min_patients, ")."
    )
  }
  mat
}

# Keep patients that pass coverage in both z-matrices (expr + split).
drop_sparse_patients_pair <- function(z_expr, z_split,
                                    max_na_frac = MAX_PATIENT_NA_FRAC,
                                    min_patients = MIN_PATIENTS) {
  patients <- intersect(colnames(z_expr), colnames(z_split))
  if (length(patients) == 0L) {
    stop("No shared patients between expression and split matrices.")
  }

  na_expr <- colMeans(!is.finite(z_expr[, patients, drop = FALSE]))
  na_split <- colMeans(!is.finite(z_split[, patients, drop = FALSE]))
  keep <- (na_expr < max_na_frac) & (na_split < max_na_frac)
  dropped <- patients[!keep]

  if (length(dropped) > 0L) {
    message(sprintf(
      "  Dropped %d patient(s) from paired PCA (NA in expr and/or split): %s",
      length(dropped), paste(dropped, collapse = ", ")
    ))
  }

  patients <- patients[keep]
  if (length(patients) < min_patients) {
    stop(
      "Fewer than ", min_patients, " patients after paired NA filtering (",
      length(patients), " left)."
    )
  }

  list(
    expr = z_expr[, patients, drop = FALSE],
    split = z_split[, patients, drop = FALSE],
    patients = patients
  )
}

# Subset genes/patients for PCA; impute remaining NA with 0 (neutral z-score) if needed.
prepare_matrix_for_pca <- function(z_mat, max_na_frac = MAX_PATIENT_NA_FRAC,
                                   min_patients = MIN_PATIENTS,
                                   impute_residual_na = TRUE, label = "PCA") {
  z <- drop_sparse_patients(z_mat, max_na_frac, min_patients, label)

  keep <- genes_with_usable_rows(z, min_patients)
  z <- z[keep, , drop = FALSE]

  gene_ok <- apply(z, 1, function(r) all(is.finite(r)) && stats::sd(r) > 0)
  z_complete <- z[gene_ok, , drop = FALSE]

  if (nrow(z_complete) >= 3L) {
    return(list(z = z_complete, imputed = FALSE, n_genes = nrow(z_complete)))
  }

  if (!impute_residual_na) {
    stop("Fewer than 3 genes with complete rows after patient filtering for ", label, ".")
  }

  n_imputed <- sum(!is.finite(z))
  z[!is.finite(z)] <- 0
  gene_ok <- apply(z, 1, function(r) stats::sd(r) > 0)
  z <- z[gene_ok, , drop = FALSE]

  if (nrow(z) < 3L) {
    stop("Fewer than 3 genes with usable z-scores for ", label, " after imputation.")
  }

  if (n_imputed > 0L) {
    message(sprintf(
      "  %s: imputed %d NA cell(s) with 0 (neutral z-score); %d genes for PCA.",
      label, n_imputed, nrow(z)
    ))
  }

  list(z = z, imputed = TRUE, n_genes = nrow(z))
}

pca_genes <- function(z_mat, n_pc = N_PC, label = "PCA") {
  prep <- prepare_matrix_for_pca(z_mat, label = label)
  z <- prep$z

  pc <- stats::prcomp(z, center = TRUE, scale. = FALSE)
  n_pc <- min(n_pc, ncol(pc$x))
  var_expl <- (pc$sdev^2) / sum(pc$sdev^2)
  list(
    coords = pc$x[, seq_len(n_pc), drop = FALSE],
    var_expl = var_expl,
    rotation = pc$rotation[, seq_len(n_pc), drop = FALSE],
    n_genes = nrow(z),
    n_patients = ncol(z),
    imputed = prep$imputed
  )
}

row_profile_correlation <- function(z_expr, z_split) {
  genes <- intersect(rownames(z_expr), rownames(z_split))
  vapply(genes, function(g) {
    xe <- z_expr[g, ]
    xs <- z_split[g, ]
    ok <- is.finite(xe) & is.finite(xs)
    if (sum(ok) < 3L) return(NA_real_)
    suppressWarnings(stats::cor(xe[ok], xs[ok], method = "pearson"))
  }, numeric(1))
}

load_splits_from_grouped_dir <- function(base_dir, dataset_name) {
  ds_dir <- file.path(path.expand(base_dir), dataset_name)
  if (!dir.exists(ds_dir)) return(NULL)
  files <- list.files(ds_dir, pattern = paste0("_", dataset_name, "_grouped\\.rds$"),
                      full.names = TRUE)
  if (length(files) == 0L) return(NULL)
  patients <- NULL
  split_rows <- list()
  for (f in files) {
    df <- tryCatch(readRDS(f), error = function(e) NULL)
    if (is.null(df) || !is.data.frame(df)) next
    exp_col <- grep("_exp$", colnames(df), ignore.case = TRUE, value = TRUE)[1]
    if (is.na(exp_col)) next
    gene <- sub(paste0("_", dataset_name, "_grouped\\.rds$"), "", basename(f))
    if (is.null(patients)) patients <- as.character(df$patient_id)
    vec <- setNames(as.character(df[[exp_col]]), as.character(df$patient_id))
    split_rows[[gene]] <- vec[patients]
  }
  if (length(split_rows) == 0L) return(NULL)
  genes_found <- names(split_rows)
  splits <- matrix(NA_character_, nrow = length(genes_found), ncol = length(patients),
                   dimnames = list(genes_found, patients))
  for (g in genes_found) splits[g, ] <- split_rows[[g]]
  splits
}

load_splits_rankgenes <- function(dataset_name, rankgenes_dir, expr_mat) {
  ds_dir <- file.path(path.expand(rankgenes_dir), dataset_name)
  all_splits_path <- file.path(ds_dir, paste0(dataset_name, ALL_SPLITS_SUFFIX))
  if (file.exists(all_splits_path)) return(readRDS(all_splits_path))
  splits <- tryCatch(
    load_splits_from_rankgenes_dir(rankgenes_dir, dataset_name),
    error = function(e) NULL
  )
  if (!is.null(splits)) return(splits)
  rank_mat <- compute_rank_matrix(expr_mat)
  build_patient_splits_matrix(expr_mat, rank_mat = rank_mat)
}

load_splits_for_source <- function(source, dataset_name, expr_mat) {
  switch(
    source,
    patient_zscore = load_splits_from_grouped_dir(ZSCORE_SPLITS_DIR, dataset_name),
    rankgenes = load_splits_rankgenes(dataset_name, RANKGENES_DIR, expr_mat),
    stop("Unknown split source: ", source)
  )
}

build_profile_pca_df <- function(pca_res, space_label) {
  df <- as.data.frame(pca_res$coords)
  colnames(df) <- paste0("PC", seq_len(ncol(df)))
  df$gene <- rownames(pca_res$coords)
  df$space <- space_label
  df$var_pc1 <- round(100 * pca_res$var_expl[1], 1)
  if (length(pca_res$var_expl) >= 2L) {
    df$var_pc2 <- round(100 * pca_res$var_expl[2], 1)
  } else {
    df$var_pc2 <- NA_real_
  }
  df
}

plot_profile_pca <- function(plot_df, df_expr, df_split, ds, source_label, cor_col = "profile_cor") {
  low_cor_thr <- stats::quantile(plot_df[[cor_col]], probs = LABEL_LOW_COR_QUANTILE, na.rm = TRUE)
  plot_df <- plot_df %>%
    dplyr::mutate(
      low_split_informativeness = !is.na(.data[[cor_col]]) &
        .data[[cor_col]] <= low_cor_thr
    )

  label_df <- plot_df %>%
    dplyr::filter(.data$space == "Expression", .data$low_split_informativeness) %>%
    dplyr::arrange(.data[[cor_col]]) %>%
    dplyr::slice_head(n = N_LABEL_GENES)

  p_expr <- ggplot(
    dplyr::filter(plot_df, .data$space == "Expression"),
    aes(x = PC1, y = PC2, color = .data[[cor_col]])
  ) +
    geom_point(alpha = 0.45, size = 1.4) +
    geom_text(
      data = label_df, aes(label = gene),
      size = 2.5, vjust = -0.6, check_overlap = TRUE, show.legend = FALSE
    ) +
    scale_color_viridis_c(option = "magma", name = "Expr–split\ncorrelation", na.value = "grey80") +
    labs(
      title = "Expression space",
      subtitle = sprintf("PC1 %.1f%% | PC2 %.1f%% variance", df_expr$var_pc1[1], df_expr$var_pc2[1]),
      x = "PC1", y = "PC2"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")

  p_split <- ggplot(
    dplyr::filter(plot_df, .data$space == "Split"),
    aes(x = PC1, y = PC2, color = .data[[cor_col]])
  ) +
    geom_point(alpha = 0.45, size = 1.4) +
    scale_color_viridis_c(option = "magma", name = "Expr–split\ncorrelation", na.value = "grey80") +
    labs(
      title = "Split space",
      subtitle = sprintf("PC1 %.1f%% | PC2 %.1f%% variance", df_split$var_pc1[1], df_split$var_pc2[1]),
      x = "PC1", y = "PC2"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")

  (p_expr | p_split) +
    plot_annotation(
      title = paste0(ds, ": Gene patient-wise profiles (row z-scored)"),
      subtitle = paste0(
        source_label, " | dark labels = bottom ", 100 * LABEL_LOW_COR_QUANTILE,
        "% expr–split correlation (least informative splits)"
      ),
      theme = theme(plot.title = element_text(face = "bold", size = 12))
    )
}

plot_profile_cor_hist <- function(metrics_df, ds, source_label) {
  ggplot(metrics_df, aes(x = profile_cor)) +
    geom_histogram(bins = 40, fill = "#4E79A7", color = "white", linewidth = 0.2) +
    geom_vline(
      xintercept = stats::quantile(metrics_df$profile_cor, LABEL_LOW_COR_QUANTILE, na.rm = TRUE),
      linetype = "dashed", color = "#E15759"
    ) +
    labs(
      title = paste0(ds, ": Expression vs split patient-profile correlation"),
      subtitle = source_label,
      x = "Per-gene correlation (row z expr vs row z split)",
      y = "Number of genes"
    ) +
    theme_minimal(base_size = 11)
}

dir.create(PROFILE_PCA_DIR, recursive = TRUE, showWarnings = FALSE)

for (ds in DATASETS) {
  message("\n=== Profile PCA: ", ds, " ===")

  pb_path <- tryCatch(
    resolve_pseudobulk_path(ds, PSEUDOBULK_DIR),
    error = function(e) NULL
  )
  if (is.null(pb_path)) next

  expr_mat <- load_gene_patient_matrix(pb_path)

  for (src in split_sources) {
    message("  Source: ", src)

    splits_mat <- load_splits_for_source(src, ds, expr_mat)
    if (is.null(splits_mat)) next

    aligned <- align_matrices(expr_mat, splits_mat)
    if (is.null(aligned)) next

    # Remove sparse / all-NA patients before z-scoring (same patients for expr + splits)
    aligned$expr <- drop_sparse_patients(aligned$expr, label = paste0(ds, " expression"))
    if (ncol(aligned$expr) < MIN_PATIENTS) {
      message("  Too few patients after expression NA filter — skipping.")
      next
    }
    aligned$splits <- aligned$splits[, colnames(aligned$expr), drop = FALSE]

    split_num <- encode_splits(aligned$splits)
    z_expr <- row_zscore_matrix(aligned$expr)
    z_split <- row_zscore_matrix(split_num)

    z_pair <- drop_sparse_patients_pair(z_expr, z_split)
    z_expr <- z_pair$expr
    z_split <- z_pair$split
    message(sprintf("  PCA / profiles use %d patients: %s",
                    length(z_pair$patients), paste(z_pair$patients, collapse = ", ")))

    profile_cor <- row_profile_correlation(z_expr, z_split)
    profile_cor <- profile_cor[is.finite(profile_cor)]

    pca_expr <- pca_genes(z_expr, label = paste0(ds, " expression"))
    pca_split <- pca_genes(z_split, label = paste0(ds, " split"))

    df_expr <- build_profile_pca_df(pca_expr, "Expression")
    df_split <- build_profile_pca_df(pca_split, "Split")
    plot_df <- dplyr::bind_rows(df_expr, df_split) %>%
      dplyr::left_join(
        data.frame(gene = names(profile_cor), profile_cor = unname(profile_cor),
                   stringsAsFactors = FALSE),
        by = "gene"
      )

    metrics_df <- data.frame(
      gene = names(profile_cor),
      profile_cor = unname(profile_cor),
      PC1_expr = df_expr$PC1[match(names(profile_cor), df_expr$gene)],
      PC2_expr = df_expr$PC2[match(names(profile_cor), df_expr$gene)],
      PC1_split = df_split$PC1[match(names(profile_cor), df_split$gene)],
      PC2_split = df_split$PC2[match(names(profile_cor), df_split$gene)],
      stringsAsFactors = FALSE
    ) %>%
      dplyr::mutate(
        pc_dist = sqrt((PC1_expr - PC1_split)^2 + (PC2_expr - PC2_split)^2),
        low_split_informativeness = profile_cor <= stats::quantile(
          profile_cor, LABEL_LOW_COR_QUANTILE, na.rm = TRUE
        )
      )

    message(sprintf(
      "  %d genes | median expr–split cor = %.3f | %.1f%% below Q%.0f",
      nrow(metrics_df),
      median(metrics_df$profile_cor),
      100 * mean(metrics_df$low_split_informativeness),
      100 * LABEL_LOW_COR_QUANTILE
    ))

    p_pca <- plot_profile_pca(
      plot_df, df_expr, df_split, ds, split_source_labels[[src]]
    )
    print(p_pca)

    p_hist <- plot_profile_cor_hist(metrics_df, ds, split_source_labels[[src]])
    print(p_hist)

    prefix <- file.path(PROFILE_PCA_DIR, paste0(ds, "_", src))
    ggsave(paste0(prefix, "_gene_profile_pca.png"), p_pca, width = 14, height = 6, dpi = 300)
    ggsave(paste0(prefix, "_profile_cor_hist.png"), p_hist, width = 8, height = 5, dpi = 300)
    write.csv(metrics_df, paste0(prefix, "_gene_profile_metrics.csv"), row.names = FALSE)
    write.csv(plot_df, paste0(prefix, "_gene_profile_pca_coords.csv"), row.names = FALSE)
    message("  Saved outputs under ", PROFILE_PCA_DIR)
  }
}
```

#Menachem Playground
```{r}
Datasets.table <- read.csv("/local/ofir/sklarz/sample_cell_summary_tirosh.tsv", sep = "\t")
```


