#installation
```{r}
devtools::install_github("CyrilLagger/scDiffCom")
devtools::install_github("DKFZ-E220/HNSCatlas")
library(scDiffCom)
library(Seurat)
library(dplyr)
# source("~/Thesis/CCC/scripts/utils/my.utils.R")
source("~/Thesis/CCC/scripts/utils/Utils.R")
DROPBOX_DIR_PATH <- file.path("/local/ofir/OCohenLab/Dropbox")

LOCAL_DROPBOX_DIR_PATH = file.path("~/Dropbox")
LOCAL_DIRECTORY_WORKING_DIR = file.path(LOCAL_DROPBOX_DIR_PATH, "gigiadir/CellChat/scripts")
```

# Example - Kurten.2021.Seurat
```{r}
library(future)
plan(sequential)  
options(future.globals.maxSize = Inf)    # disable the limit
options(future.globals.onReference = "ignore")


Kurten.2021.AXL.Expression <- add_gene_expression_group(Kurten.2021.Seurat, assay="alra", gene="AXL")
Kurten.2021.Seurat <- Kurten.2021.AXL.Expression$seurat_obj

Valid.Cell.Types.For.scDiffCom <- Kurten.2021.Seurat@meta.data %>% 
  filter(cell_type != "") %>%  
  group_by(cell_type) %>% 
  filter(all(c("HIGH", "LOW") %in% AXL_EXP)) %>% 
  pull(cell_type) %>% 
  unique()

Kurten.2021.Seurat.scDiffCom.Base <- subset(
    Kurten.2021.Seurat, 
    subset = cell_type %in% Valid.Cell.Types.For.scDiffCom & !is.na(AXL_EXP)
)


options(scDiffCom.no_parallel = TRUE)
options(future.globals.maxSize = 20 * 1024^3)   # 20 GB
Kurten.2021.scDiffCom.Object <- run_interaction_analysis(
  seurat_object = Kurten.2021.Seurat.scDiffCom.Base,
  seurat_assay="alra",
  LRI_species = "human",
  seurat_celltype_id = "cell_type",
  seurat_condition_id = list(
    column_name="AXL_EXP",
    cond1_name="HIGH",
    cond2_name="LOW"
  ),
  iterations = 1000
)

```
## View results on Shiny
```{r}
if (!require("shiny")) install.packages("shiny")
if (!require("shinyWidgets")) install.packages("shinyWidgets")
if (!require("shinythemes")) install.packages("shinythemes")
if (!require("DT")) install.packages("DT")
if (!require("kableExtra")) install.packages("kableExtra")
if (!require("RColorBrewer")) install.packages("RColorBrewer")
if (!require("igraph")) install.packages("igraph")
if (!require("visNetwork")) install.packages("visNetwork")
if (!require("plotly")) install.packages("plotly")
if (!require("GOSemSim")) BiocManager::install("GoSemSim")
if (!require("rrvgo")) BiocManager::install("rrvgo")
if (!require("org.Mm.eg.db")) BiocManager::install("org.Mm.eg.db")
```

```{r}
BuildShiny(Kurten.2021.scDiffCom.Object)
```

```{r}
CCI_detected <- GetTableCCI(Kurten.2021.scDiffCom.Object, type = "detected", simplified = TRUE)
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

ggplot(
  CCI_detected,
  aes(
     x = LOGFC,
     y = -log10(BH_P_VALUE_DE + 1E-2),
     colour = REGULATION
  )
) + geom_point(
) + scale_colour_manual(
  values = c("UP" = "red", "DOWN" = "blue", "FLAT" = "green", "NSC" = "grey")
) + xlab(
  "log(FC)"
) + ylab(
  "-log10(Adj. p-value)"
)

```

```{r}
PlotORA(
  object = Kurten.2021.scDiffCom.Object,
  category = "LRI",
  regulation = "UP"
) + theme(
  legend.position = c(0.85, 0.4),
  legend.key.size = unit(0.4, "cm")
)
```





# scDiffCom Pipeline

## Util Functions
```{r}
#
#  my.utils.R
#    - add_gene_expression_group 
library(Seurat)
library(dplyr)
library(scDiffCom)
source("~/Thesis/CellChat/scripts/utils/my.utils.R")


get_valid_cell_types <- function(seurat_obj) {
  valid_cell_types <- seurat_obj@meta.data %>%
    filter(cell_type != "" & AXL_EXP %in% c("HIGH", "LOW")) %>%
    group_by(cell_type) %>%
    filter(all(c("HIGH", "LOW") %in% AXL_EXP)) %>%
    pull(cell_type) %>%
    unique()
  
  message(sprintf("Found %d valid cell types present in both HIGH and LOW groups", 
                 length(valid_cell_types)))
  
  return(valid_cell_types)
}

get_patient_col <- function(seurat_obj, patient_col_options) {
  cell_metadata <- seurat_obj@meta.data
  available_cols <- colnames(cell_metadata)

  patient_col <- NULL
  for(col in patient_col_options) {
    if (col %in% available_cols) {
      patient_col <- col
      cat(sprintf("  Using patient column: %s\n", patient_col))
      break
    }
  }
  
  return(patient_col)
}


```



## Load Data
```{r}
SEURAT_OBJECTS_DIR <- file.path(DROPBOX_DIR_PATH, "hadasra/thesis/test_objects")
SEURAT_OBJECTS_OUTPUT_DIR <- file.path("~/Thesis/CellChat/outputs/scDiffCom/Seurats")

all_seurat_files <- list.files(SEURAT_OBJECTS_DIR, pattern = "\\.RDS$", full.names = TRUE, ignore.case = TRUE)

process_seurat_files <- function(
    file_paths,
    output_dir = "processed_seurat_objects",
    log_file = "skipped_datasets.txt",
    gene = "AXL",
    patient_col_options = c("orig.ident"),
    quantiles = c(1/3, 2/3),
    min_samples = 12,
    min_target_cells = 500
) {
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  writeLines(c(
    paste("Processing Log -", Sys.time()),
    "file_name\treason\tdetails",
    "---------------------------------------------------"
  ), con = log_file)
  
  log_skip <- function(name, reason, details) {
    line <- sprintf("%s\t%s\t%s", name, reason, details)
    write(line, file = log_file, append = TRUE)
    cat(sprintf("  -> SKIPPING: %s (%s)\n", reason, details))
  }
  
  
  cat(sprintf("Processing %d files...\n\n", length(file_paths)))
  cat(sprintf("Logging skipped datasets to: %s\n\n", log_file))
  
  for (i in seq_along(file_paths)) {
    file_path <- file_paths[i]
    file_name <- basename(file_path)
    object_name <- gsub("\\.RDS$", "", file_name, ignore.case = TRUE)
    
    cat(sprintf("[%d/%d] Processing %s...\n", i, length(file_paths), file_name))
    
    tryCatch({
      seurat_obj <- readRDS(file_path)
      cat(sprintf("  Loaded: %d cells, %d features\n", ncol(seurat_obj), nrow(seurat_obj)))
      
      patient_col <- get_patient_col(seurat_obj = seurat_obj, 
                                     patient_col_options = patient_col_options)
      
      # Validate Sample num
      n_samples <- length(unique(seurat_obj@meta.data[[patient_col]]))
      
      if (n_samples < min_samples) {
        log_skip(file_name, "Low Sample Count", sprintf("Found %d samples (Min: %d)", n_samples, min_samples))
        next 
      }
      
      # Validate more than N = 500 Malignant
      if (!"cell_type" %in% colnames(seurat_obj@meta.data)) {
        log_skip(file_name, "Missing Column", "'cell_type' column not found in metadata")
        next
      }

      cell_identities <- as.character(seurat_obj@meta.data$cell_type)
      
      # Regex to match Epithelial, Tumor, Cancer, or Malignant (case insensitive)
      target_pattern <- "Epithelial|Tumor|Cancer|Malignant"
      n_target_cells <- sum(grepl(target_pattern, cell_identities, ignore.case = TRUE))
      
      if (n_target_cells < min_target_cells) {
        log_skip(file_name, "Low Target Cells", sprintf("Found %d matching '%s' (Min: %d)", n_target_cells, target_pattern, min_target_cells))
        next 
      }
      
      result <- add_gene_expression_group(
        seurat_obj = seurat_obj,
        gene = gene,
        patient_col = patient_col,
        quantiles = quantiles
      )
      
      # Extract results
      updated_seurat <- result$seurat_obj
      patient_summary <- result$patient_summary
      
      cat(sprintf("  Added %s_EXP grouping: %d patients\n", gene, nrow(patient_summary)))
      
      output_file <- file.path(output_dir, paste0(object_name, "_processed.rds"))
      
      force(result) 
      
      saveRDS(result, file=output_file)
      cat(sprintf("  ✓ Saved to: %s\n", output_file))
      
    }, error = function(e) {
      cat(sprintf("  ✗ Error: %s\n", e$message))
    }, finally = {
      # --- FIX 3: Memory Cleanup ---
      # Remove large objects from the current environment to help R free memory
      if (exists("seurat_obj")) rm(seurat_obj)
      if (exists("result")) rm(result)
      if (exists("updated_seurat")) rm(updated_seurat)
      
      # Force garbage collection to return memory to OS (crucial for heavy Seurat loops)
      gc(verbose = FALSE)
      cat("\n") 
    })
  }
  cat("=== PROCESSING COMPLETE ===\n")
  cat(sprintf("All processed files saved to: %s\n", output_dir))
}
```


```{r}
selected_files <- all_seurat_files

process_seurat_files(selected_files, 
                     gene="AXL",
                     patient_col_options = c("sampleID", "orig.ident"),
                     output_dir = SEURAT_OBJECTS_OUTPUT_DIR)
```



## Main Pipeline
```{r}
Run.scDiffCom.Pipeline.For.Seurat <- function(seurat_obj) {
  valid_cell_types <- get_valid_cell_types(seurat_obj)
  
  seurat_obj_for_scDiffCom <- subset(
      seurat_obj, 
      subset = cell_type %in% valid_cell_types & !is.na(AXL_EXP)
  )
  scDiffCom_object <- run_interaction_analysis(
    seurat_object = seurat_obj_for_scDiffCom,
    LRI_species = scDiffCom.Pipeline.Config$LRI_species,
    seurat_celltype_id = scDiffCom.Pipeline.Config$celltype_column,
    seurat_condition_id = list(
      column_name="AXL_EXP",
      cond1_name="HIGHׁ",
      cond2_name="LOW"
    ),
    iterations = scDiffCom.Pipeline.Config$iterations
  )
  
  return(scDiffCom_object)
}

scDiffCom.Pipeline.Runner <- function(
    file_paths,
    output_dir = "scDiffCom_results"
) {
  # Create output directory if it doesn't exist
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  cat(sprintf("Processing %d files...\n\n", length(file_paths)))
  
  for (i in seq_along(file_paths)) {
    processed_file <- file_paths[i]
    
    if (!file.exists(processed_file)) {
      cat(sprintf("[%d/%d] ✗ File not found: %s\n\n", i, length(file_paths), processed_file))
      next
    }
    
    file_name <- basename(processed_file)
    object_name <- gsub("_processed\\.(RDS|rds)$", "", file_name, ignore.case = TRUE)

    cat(sprintf("[%d/%d] Processing %s...\n", i, length(file_paths), object_name))
    
    tryCatch({
      # Load processed file with grouped data
      cat("  Loading processed file...\n")
      processed_with_group_result <- readRDS(processed_file)
      
      # Extract Seurat object and patient summary from the result
      seurat_obj <- processed_with_group_result$seurat_obj
      patient_summary <- processed_with_group_result$patient_summary
      
      cat(sprintf("  Loaded: %d cells, %d features, %d patients\n", 
                  ncol(seurat_obj), nrow(seurat_obj), nrow(patient_summary)))
      
      # Check if AXL_EXP column exists
      if (!"AXL_EXP" %in% colnames(seurat_obj@meta.data)) {
        cat("  ✗ Warning: AXL_EXP column not found, skipping...\n\n")
        next
      }
      
      # Count cells with AXL_EXP annotation
      axl_annotated <- sum(!is.na(seurat_obj@meta.data$AXL_EXP))
      axl_high <- sum(seurat_obj@meta.data$AXL_EXP == "HIGH", na.rm = TRUE)
      axl_low <- sum(seurat_obj@meta.data$AXL_EXP == "LOW", na.rm = TRUE)
      
      cat(sprintf("  AXL_EXP annotated: %d cells (HIGH: %d, LOW: %d)\n", 
                  axl_annotated, axl_high, axl_low))
      
      if (axl_annotated == 0) {
        cat("  ✗ Warning: No cells with AXL_EXP annotation, skipping...\n\n")
        next
      }
      
      # Run scDiffCom pipeline
      cat("  Running scDiffCom pipeline...\n")
      scDiffCom_object <- Run.scDiffCom.Pipeline.For.Seurat(seurat_obj)
      
      cat("  scDiffCom analysis complete\n")
      # Save results
      output_file <- file.path(output_dir, paste0(object_name, "_scDiffCom.RDS"))
      
      cat("  Saving results...\n")
      force(scDiffCom_object)
      saveRDS(scDiffCom_object, output_file)
      
      cat(sprintf("  ✓ Saved to: %s\n\n", output_file))
      
    }, error = function(e) {
      cat(sprintf("  ✗ Error: %s\n\n", e$message))
    })
  }
  
  cat("=== PIPELINE COMPLETE ===\n")
  cat(sprintf("All scDiffCom results saved to: %s\n", output_dir))
  
}
```

## Pipeline Config
```{r}
scDiffCom.Pipeline.Config <- list (
  # QC
  gene = "AXL",
  assay = "RNA",
  min_samples = 12,
  min_epithelial_cells = 500,
  epithelial_cell_types = c("Epithelial", "Malignant", "Tumor", "Cancer"),
  
  # scDiffCom parameters
  iterations = 10000,
  LRI_species = "human",
  celltype_column = "cell_type",
  
  # Parallel processing
  n_workers = 1,
  
  # Output
  output_dir = "~/Thesis/CellChat/outputs/scDiffCom"
)
```
## Pipeline Runner
```{r}

all_processed_files <- list.files("~/Thesis/CellChat/outputs/scDiffCom/Seurats", full.names = TRUE)

selected_files <- c("/gpfs0/bgu-ofircohen/users/gigiadir/Thesis/CellChat/outputs/scDiffCom/Seurats/Kurten2021_Head_and_Neck_processed.rds")

scDiffCom.Pipeline.Runner(file_paths = selected_files,
                          output_dir="~/Thesis/CellChat/outputs/scDiffCom/scDiffComs")
```

## Inspect Shiny
```{r}

scDiffCom.Output.Path <- "~/Thesis/CellChat/outputs/scDiffCom/scDiffComs"

selected_scDiffCom_obj_path <- file.path(scDiffCom.Output.Path, "Kurten2021_Head_and_Neck_scDiffCom.RDS")

selected_scDiffCom_obj <- readRDS(selected_scDiffCom_obj_path)

BuildShiny(selected_scDiffCom_obj)
```



# scDiffCom Pipeline - V2

## Load Data
```{r}
V3.Pipeline.Config <- list(
  #Seurat Parsing Config
  n.var.genes = 5000,
  n.cancner.types = 4
)
CANCER_DB_DIR_PATH <- file.path(DROPBOX_DIR_PATH, "Cancer")

my.save.RData("Bassez.2021.Seurat", myfolder="~/Thesis/CellChat/outputs/data", is.load.obj = T)
my.save.RData("Bassez.2021.Seurat", myfolder=file.path(DROPBOX_DIR_PATH, "Cancer_DB/SC.3CA.Tirosh/Breast"), is.load.obj = T)

HNSCC.Atlas <- readRDS(file.path(CANCER_DB_DIR_PATH, "SingleCell_Data_Misc/HNSCC_Atlas.rds"))
DefaultAssay(HNSCC.Atlas) <- "RNA"
my.save.RData("HNSCC.Atlas", myfolder="~/Thesis/CCC/outputs/data", is.load.obj = T)


Chan.2021.Seurat <- readRDS(file.path(CANCER_DB_DIR_PATH,"SC.3CA.Tirosh/SeuratObjects_10x_cellnames_with_project/Chan2021_Lung.RDS"))

Chan.2021.Seurat <- subset(Chan.2021.Seurat, orig.ident != "PleuralEffusion")
Chan.2021.Seurat$orig.ident <- droplevels(Chan.2021.Seurat$orig.ident)

my.save.RData("Chan.2021.Seurat", myfolder="~/Thesis/CellChat/outputs/data")

Hadas.Directory <- file.path("~/../hadasra/thesis/new_objects")
Bassez.Breast.Seurat <- readRDS(file.path(Hadas.Directory, "Bassez2021_Breast.RDS"))
Chan.Lung.Seurat <- readRDS(file.path(Hadas.Directory, "Chan2021_Lung.RDS"))
Biermann.Melanoma.Seurat <- readRDS(file.path(Hadas.Directory, "Biermann2022_Skin.RDS"))
Chen.Prostate.Seurat <- readRDS(file.path(Hadas.Directory, "Chen2021_Prostate.RDS"))
Sharma.Liver.Seurat <- readRDS(file.path(Hadas.Directory, "Sharma2020_Liver_Biliary.RDS"))
Young.Kidney.Seurat <- readRDS(file.path(Hadas.Directory, "Young2018_Kidney.RDS"))
```

## FindVariableFeatures in Malignant
```{r}
HNSCC.Atlas.Malignant.Seurat <- subset(HNSCC.Atlas, Cell_Type == "Epithelial")
HNSCC.Atlas.Malignant.Seurat <- FindVariableFeatures(HNSCC.Atlas.Malignant.Seurat,
                                                     nfeatures = V3.Pipeline.Config$n.var.genes)
HNSCC.Atlas.Malignant.Top.5000.Genes <- VariableFeatures(HNSCC.Atlas.Malignant.Seurat)


Bassez.Breast.Malignant.Seurat <- subset(Bassez.Breast.Seurat, cell_type == "Malignant")
Bassez.Breast.Malignant.Seurat <- FindVariableFeatures(Bassez.Breast.Malignant.Seurat,
                                                     nfeatures = V3.Pipeline.Config$n.var.genes)
Bassez.Breast.Malignant.Top.5000.Genes <- VariableFeatures(Bassez.Breast.Malignant.Seurat)


Chan.Lung.Malignant.Seurat <- subset(Chan.Lung.Seurat, Consensus_Cell_Type == "Epithelial")
Chan.Lung.Malignant.Seurat <- FindVariableFeatures(Chan.Lung.Malignant.Seurat,
                                                   nfeatures = V3.Pipeline.Config$n.var.genes)
Chan.Lung.Malignant.Top.5000.Genes <- VariableFeatures(Chan.Lung.Malignant.Seurat)

Biermann.Melanoma.Malignant.Seurat <- subset(Biermann.Melanoma.Seurat, Consensus_Cell_Type == "Epithelial")
Biermann.Melanoma.Malignant.Seurat <- FindVariableFeatures(Biermann.Melanoma.Malignant.Seurat,
                                                           nfeatures = V3.Pipeline.Config$n.var.genes)
Biermann.Melanoma.Malignant.Top.5000.Genes <- VariableFeatures(Biermann.Melanoma.Malignant.Seurat)

Chen.Prostate.Malignant.Seurat <- subset(Chen.Prostate.Seurat, Consensus_Cell_Type == "Epithelial")
Chen.Prostate.Malignant.Seurat <- FindVariableFeatures(Chen.Prostate.Malignant.Seurat,
                                                           nfeatures = V3.Pipeline.Config$n.var.genes)
Chen.Prostate.Malignant.Top.5000.Genes <- VariableFeatures(Chen.Prostate.Malignant.Seurat)


Sharma.Liver.Malignant.Seurat <- subset(Sharma.Liver.Seurat, Consensus_Cell_Type == "Epithelial")
Sharma.Liver.Malignant.Seurat <- FindVariableFeatures(Sharma.Liver.Malignant.Seurat,
                                                           nfeatures = V3.Pipeline.Config$n.var.genes)
Sharma.Liver.Malignant.Top.5000.Genes <- VariableFeatures(Sharma.Liver.Malignant.Seurat)


Young.Kidney.Malignant.Seurat <- subset(Young.Kidney.Seurat, Consensus_Cell_Type == "Epithelial")
Young.Kidney.Malignant.Seurat <- FindVariableFeatures(Young.Kidney.Malignant.Seurat,
                                                           nfeatures = V3.Pipeline.Config$n.var.genes)
Young.Kidney.Malignant.Top.5000.Genes <- VariableFeatures(Young.Kidney.Malignant.Seurat)


```


# Try to get Genes to check
```{r}
library(readr)

# Define file paths
cosmic_path <- file.path(DROPBOX_DIR_PATH, "Cancer_DB/COSMIC/COSMIC.v103/Cosmic_CancerGeneCensus_v103_GRCh37.tsv")
ncg_path <- file.path(DROPBOX_DIR_PATH, "Cancer_DB/NetworkCancerGenes/NCG5.cancer.genes.tsv")
oncokb_path <- file.path(DROPBOX_DIR_PATH, "Cancer_DB/oncokb/OncoKB.2.2.2025.cancerGeneList.tsv")

# Read the files
cosmic_data <- read_tsv(cosmic_path)
oncokb_data <- read_tsv(oncokb_path)
ncg_data <- read_tsv(ncg_path)

library(dplyr)
library(tidyverse)

all_cancer_genes <- c(
  oncokb_data$`Hugo Symbol`,
  cosmic_data$GENE_SYMBOL,
  ncg_data$symbol
) %>%
  unique() %>%
  sort()

 # 1. Combine your 7 lists into one big vector
all_study_lists <- list(
  HNSCC.Atlas = HNSCC.Atlas.Malignant.Top.5000.Genes,
  Bassez.Breast = Bassez.Breast.Malignant.Top.5000.Genes,
  Chan.Lung = Chan.Lung.Malignant.Top.5000.Genes,
  Biermann.Melanoma = Biermann.Melanoma.Malignant.Top.5000.Genes,
  Chen.Prostate = Chen.Prostate.Malignant.Top.5000.Genes,
  Sharma.Liver = Sharma.Liver.Malignant.Top.5000.Genes,
  Young.Kidney = Young.Kidney.Malignant.Top.5000.Genes
)

Top.5000.Malignant.Genes.Per.Cancer <- enframe(all_study_lists, name = "Source", value = "Gene_name") %>%
  unnest(Gene_name)


# 3. Identify genes that appear in at least K = 4 sources
gene_counts <- table(Top.5000.Malignant.Genes.Per.Cancer$Gene_name)
frequent_study_genes <- names(gene_counts[gene_counts >= V3.Pipeline.Config$n.cancner.types])



# 4. Find the intersection with your COSMIC/OncoKB 'all_cancer_genes'
final_gene_selection <- intersect(all_cancer_genes, frequent_study_genes)
# I added some genes from final_gene_selection_ILANIT_version (such as AXL, ERBB2 which only appeared on 3 types and not 4).

# 5. Save the result
saveRDS(final_gene_selection, file.path("~/Thesis/CCC/outputs/data/final_gene_selection.rds"))

# Summary of results
print(paste("Total frequent genes found (K >= 4):", length(frequent_study_genes)))
print(paste("Final count after filtering by Cancer DBs:", length(final_gene_selection)))


Complexes.Genes <- unique(unlist(GSDB.list.of.lists$CORUM_core_complexes))
Complexes.Oncogenes.OncoKB.Cosmic.NCG <- intersect(Complexes.Genes, all_cancer_genes)
Complexes.Oncogenes.OncoKB <- intersect(Complexes.Genes, oncokb_data$`Hugo Symbol`)

saveRDS(Complexes.Oncogenes.OncoKB.Cosmic.NCG, "~/Thesis/CCC/outputs/data/Complexes.Oncogenes.OncoKB.Cosmic.NCG.rds")
my.save.RData("Complexes.Oncogenes.OncoKB", "~/Thesis/CCC/outputs/data")

```




# After Mass Run - Try to combine results for ILANIT
## Analysis Globals & Helpers
```{r}
source("~/Thesis/CCC/scripts/utils/Utils.R")
source("~/Thesis/CCC/scripts/utils/my.utils.R")

library(RColorBrewer)
library(pheatmap)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(scDiffCom)
library(ggplot2)
library(ggrepel)


BASE_RESULTS_DIR <- "/gpfs0/bgu-ofircohen/users/gigiadir/Thesis/CCC/outputs/scDiffCom/scDiffComs"
MALIGNANT_CELLTYPE <- "Epithelial"
USE_COSINE         <- TRUE           # TRUE → cosine distance instead of correlation
MIN_GENES_PER_CCI  <- 2              # CCIs must appear in ≥ N genes (filters sparse rows)
OUTPUT_DIR         <- "."            # where to save the heatmap PNGs

LRI_GO <- scDiffCom::LRI_human$LRI_curated_GO
BP_GO_INFO <- scDiffCom::gene_ontology_level %>% 
  filter(ASPECT == "biological_process") %>%
  select(ID, NAME)
LRI_GO_BP <- LRI_GO %>%
  inner_join(BP_GO_INFO, by = c("GO_ID" = "ID"))

```

## Analysis Utils
```{r}
switch.conditions.columns <- function(scDiffCom_obj) {
    #ONLY NEEDED ON THOSE OBJECTS WITHOUT _step4 suffix.
    message("switching conditions HIGH and LOW")
    cci_table_detected <- data.table::copy(scDiffCom_obj@cci_table_detected)
    
    # Multiply LOGFC by -1
    logfc_cols = c("LOGFC")
    cci_table_detected[, (logfc_cols) := lapply(.SD, function(x) x * -1), .SDcols = logfc_cols]
    
   # 2. Switch REGULATION values (UP <-> DOWN)
    cci_table_detected[REGULATION == "UP", REGULATION := "DOWN_temp"]
    cci_table_detected[REGULATION == "DOWN", REGULATION := "UP"]
    cci_table_detected[REGULATION == "DOWN_temp", REGULATION := "DOWN"]
    
    cci_table_detected$SWITCHED_COLUMNS <- TRUE
    # Write back to the object
    scDiffCom_obj@cci_table_detected <- cci_table_detected
    
    return(scDiffCom_obj)
}

load.dataset.scDiffComs <- function(dataset_name, results_dir = BASE_RESULTS_DIR) {
  #ONLY NEEDED ON THOSE OBJECTS WITHOUT _step4 suffix.
  message(".  Loading scDiffComs results for dataset: ", dataset_name)
  rds_dir <- file.path(results_dir, dataset_name)
  rds_files <- list.files(rds_dir, pattern=paste0("_", dataset_name, "_scDiffCom\\.rds$"), full.names = T)
  
  scDiffCom_list <- list()
  
  for (rds_path in rds_files) {
    fname <- basename(rds_path)
    gene <- str_remove(fname, paste0("_", dataset_name, "_scDiffCom\\.rds$"))
    message("  Loading gene: ", gene)
    scDiffCom_obj <- readRDS(rds_path)
    scDiffCom_obj <- switch.conditions.columns(scDiffCom_obj)
    scDiffCom_list[[gene]] <- scDiffCom_obj
  }
  
  return(scDiffCom_list)
}

filter.scDiffCom.cci_table_detected.for.malignant <- function(scDiffCom_obj, malignant_celltype = "Epithelial") {
  filtered_cci_table <- scDiffCom_obj@cci_table_detected %>%
    filter(
      (EMITTER_CELLTYPE == malignant_celltype | RECEIVER_CELLTYPE == malignant_celltype),
      IS_CCI_DE == TRUE,
      is.finite(LOGFC)
    )
  
  return (filtered_cci_table)
}

filter.scDiffCom.cci_table_detected.for.celltypes <- function(scDiffCom_obj, emitter_celltype, receiver_celltype) {
  filtered_cci_table <- scDiffCom_obj@cci_table_detected %>%
    filter(
      EMITTER_CELLTYPE == emitter_celltype,
      RECEIVER_CELLTYPE == receiver_celltype,
      IS_CCI_DE == TRUE,
      is.finite(LOGFC)
    )
  
  return (filtered_cci_table)
}
```

## Load Data
```{r}
DATASET <- "Bassez2021_Breast"
Bassez2021_Breast.scDiffComs <- load.dataset.scDiffComs("Bassez2021_Breast")
Chan2021_Lung.scDiffComs <- load.dataset.scDiffComs("Chan2021_Lung")
HNSCC_Atlas.scDiffComs <- load.dataset.scDiffComs("HNSCC_Atlas")


Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant <- lapply(Bassez2021_Breast.scDiffComs, filter.scDiffCom.cci_table_detected.for.malignant)
HNSCC_Atlas.malignant <- lapply(HNSCC_Atlas.scDiffComs, filter.scDiffCom.cci_table_detected.for.malignant)
Chan2021_Lung.malignant <- lapply(Chan2021_Lung.scDiffComs, filter.scDiffCom.cci_table_detected.for.malignant)
```


## GENE x GENE heatmap
```{r}
Bassez2021_Breast.LOGFC_list <- list()

for(gene in names(Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant)) {
  LOGFC_table <- Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant[[gene]] %>%
    select(CCI, LOGFC) %>%
    group_by(CCI) %>%
    summarise(LOGFC = mean(LOGFC, na.rm = TRUE), .groups = "drop")
  
  Bassez2021_Breast.LOGFC_list[[gene]] <- LOGFC_table
}

# prepare LOGFC matrix for heatmap
all_data <- bind_rows(
  lapply(names(Bassez2021_Breast.LOGFC_list), function(g) {
    Bassez2021_Breast.LOGFC_list[[g]] %>% mutate(Gene = g)
  })
)

mat_long <- all_data %>%
  pivot_wider(names_from = Gene, values_from = LOGFC)

mat <- as.matrix(mat_long[, -1])
rownames(mat) <- mat_long$CCI
  
message("  Matrix dimensions (CCIs x Genes): ", nrow(mat), " x ", ncol(mat))
  
n_non_na <- rowSums(!is.na(mat))
mat <- mat[n_non_na >= MIN_GENES_PER_CCI, , drop = FALSE]
message("  After filtering sparse CCIs: ", nrow(mat), " x ", ncol(mat))

sim_mat   <- cosine_sim_na(mat)
diag(sim_mat) <- NA

library(gplots)
# heatmap
sim_label <- "Cosine Similarity"
sim_file  <- paste0(DATASET, "_cosine_heatmap.png")
gene_gene_heatmap <- pheatmap(
    sim_mat,
    color          =  colorRampPalette(c("red", "white", "blue"))(100),
    breaks         = seq(-1, 1, length.out = 101),
    #na_col         = "grey90",
    main           = paste0(DATASET, " – Gene-Gene ", sim_label),
    fontsize_row   = 8,
    fontsize_col   = 8,
  )

n_genes <- nrow(sim_mat)
px_per_gene <- 20  # increase if still overlapping

png("gene-gene-heatmap2.png", width = max(4000, n_genes * px_per_gene), 
                               height = max(4000, n_genes * px_per_gene), res = 300)
gplots::heatmap.2(
    sim_mat,
    dendrogram = "row",
    trace      = "none",
    col        = redgreen(100),
    margins    = c(14, 14),
    cexRow     = 0.6,
    cexCol     = 0.6,
    srtCol     = 45
)
dev.off()

png("gene-gene-heatmap3.png", width = 4000, height = 4000, res = 300)
pheatmap(
    sim_mat,
    color        = colorRampPalette(c("red", "white", "blue"))(100),
    breaks       = seq(-1, 1, length.out = 101),
    fontsize_row = max(4, 200 / nrow(sim_mat)),  # auto-shrink based on count
    fontsize_col = max(4, 200 / nrow(sim_mat)),
    angle_col    = "45",
    main         = paste0(DATASET, " – Gene-Gene Cosine Similarity")
)
dev.off()
# ── UMAP of genes based on their LRI (CCI) LOGFC profiles ──────────────────
library(umap)
library(ggplot2)
library(ggrepel)

mat_genes <- t(mat)
mat_genes_imputed <- mat_genes
mat_genes_imputed[is.na(mat_genes_imputed)] <- 0

# ── Hierarchical clustering on cosine distance (matches your heatmap) ────────
k_clusters <- 5  # adjust to however many clusters make sense

dist_mat   <- as.dist(1 - sim_mat)   # cosine distance from your sim_mat
hclust_res <- hclust(dist_mat, method = "complete")
clusters   <- cutree(hclust_res, k = k_clusters)

# ── UMAP ─────────────────────────────────────────────────────────────────────
set.seed(42)
umap_config              <- umap.defaults
umap_config$n_neighbors  <- min(15, nrow(mat_genes_imputed) - 1)
umap_config$min_dist     <- 0.1
umap_config$metric       <- "cosine"

umap_result <- umap(mat_genes_imputed, config = umap_config)

umap_df <- data.frame(
  Gene    = rownames(mat_genes_imputed),
  UMAP1   = umap_result$layout[, 1],
  UMAP2   = umap_result$layout[, 2],
  n_CCIs  = rowSums(!is.na(mat_genes)),
  Cluster = factor(clusters[rownames(mat_genes_imputed)])  # match order safely
)

# ── Plot ──────────────────────────────────────────────────────────────────────
p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = Cluster, label = Gene)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    size          = 2.8,
    max.overlaps  = 40,
    segment.size  = 0.2,
    segment.alpha = 0.5,
    show.legend   = FALSE
  ) +
  scale_colour_brewer(palette = "Set1", name = "Cluster") +
  labs(
    title    = paste0(DATASET, " – Gene UMAP (LRI / CCI LOGFC space)"),
    subtitle = paste0("Hierarchical clustering (k = ", k_clusters, ", cosine distance)"),
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

ggsave(
  paste0(DATASET, "_gene_umap.png"),
  plot   = p_umap,
  width  = 10,
  height = 8,
  dpi    = 300
)

message("UMAP saved to ", DATASET, "_gene_umap.png")

# ── Print cluster membership so you can inspect them ─────────────────────────
cluster_df <- umap_df %>% 
  select(Gene, Cluster, n_CCIs) %>% 
  arrange(Cluster, Gene)
print(cluster_df)

cluster_summary <- cluster_df %>%
  group_by(Cluster) %>%
  summarise(
    n_genes = n(),
    genes   = paste(sort(Gene), collapse = ", ")
  ) %>%
  arrange(Cluster)

print(cluster_summary)
  
```



## GO x CCI - Cancer-T-Cell
```{r}
# ── 0.  CONFIG ────────────────────────────────────────────────
LOGFC_AGG   <- "mean"   # "mean" | "sum"
N_NEIGHBORS <- 30       # KNN for graph + UMAP
RESOLUTION  <- 1.2     # Leiden resolution parameter
MIN_DIST    <- 0.1      # UMAP min_dist
SEED        <- 42

# ── 1.  PACKAGES ──────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(tibble)
library(uwot)
library(FNN)
library(igraph)
library(ggplot2)
library(ggrepel)

# ── 2.  REFERENCE TABLES  ─────────────────────────────────────
message("Loading reference tables ...")

LRI_GO <- scDiffCom::LRI_human$LRI_curated_GO

GO_BP <- scDiffCom::gene_ontology_level %>%
  filter(ASPECT == "biological_process") %>%
  select(ID, NAME)

LRI_GO_BP <- LRI_GO %>%
  inner_join(GO_BP, by = c("GO_ID" = "ID")) %>%
  rename(GO_NAME = NAME)

# ── 3.  INTERACTION DATA ──────────────────────────────────────
Bassez2021_Breast.Malignant_Tcell.interactions <- lapply(
  Bassez2021_Breast.scDiffComs,
  filter.scDiffCom.cci_table_detected.for.celltypes,
  emitter_celltype = MALIGNANT_CELLTYPE,
  receiver_celltype = "T-cells"
)

Bassez2021_Breast.Malignant_Tcell.interactions.all_data <- bind_rows(
  lapply(names(Bassez2021_Breast.Malignant_Tcell.interactions), function(g) {
    Bassez2021_Breast.Malignant_Tcell.interactions[[g]] %>% mutate(Gene = g)
  })
)

# ── 4.  GENE × GO-BP MATRIX ───────────────────────────────────
agg_fn <- match.fun(LOGFC_AGG)

filt_go <- Bassez2021_Breast.Malignant_Tcell.interactions.all_data %>%
  select(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, LRI, LOGFC, Gene) %>%
  inner_join(LRI_GO_BP, by = "LRI", relationship = "many-to-many")

gene_go_matrix <- filt_go %>%
  group_by(Gene, GO_NAME) %>%
  summarise(
    agg_LOGFC = agg_fn(LOGFC, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = GO_NAME,
    values_from = agg_LOGFC,
    values_fill = 0
  )

gene_go_mat <- gene_go_matrix %>%
  column_to_rownames("Gene") %>%
  as.matrix()

# ── 5.  PRE-PROCESSING ────────────────────────────────────────
nz_cols <- apply(gene_go_mat, 2, var) > 0
message(sprintf("Keeping %d / %d GO-BP terms (non-zero variance)",
                sum(nz_cols), ncol(gene_go_mat)))

mat_scaled <- scale(gene_go_mat[, nz_cols])
mat_scaled[!is.finite(mat_scaled)] <- 0

# ── 6.  KNN GRAPH ─────────────────────────────────────────────
set.seed(SEED)
k   <- min(N_NEIGHBORS, nrow(mat_scaled) - 1)
knn <- FNN::get.knn(mat_scaled, k = k)

n     <- nrow(mat_scaled)
edges <- do.call(rbind, lapply(seq_len(n), function(i) {
  cbind(i, knn$nn.index[i, ])
}))
g <- igraph::graph_from_edgelist(edges, directed = FALSE) %>%
  igraph::simplify()

# ── 7.  LEIDEN CLUSTERING ─────────────────────────────────────
leiden_comm <- igraph::cluster_leiden(
  g,
  objective_function   = "modularity",
  resolution_parameter = RESOLUTION
)
clusters <- factor(leiden_comm$membership)
message(sprintf("Leiden found %d clusters", nlevels(clusters)))

# ── 8.  UMAP ──────────────────────────────────────────────────
set.seed(SEED)
umap_coords <- uwot::umap(
  mat_scaled,
  n_neighbors = k,
  min_dist    = MIN_DIST,
  metric      = "cosine"
)

umap_df <- data.frame(
  UMAP1  = umap_coords[, 1],
  UMAP2  = umap_coords[, 2],
  Leiden = clusters,
  Gene   = rownames(mat_scaled)
)

# ── 9.  PLOT ──────────────────────────────────────────────────
p <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2,
                          colour = Leiden, label = Gene)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(
    size         = 2.8,
    max.overlaps = 20,
    show.legend  = FALSE
  ) +
  scale_colour_brewer(palette = "Set2") +
  theme_classic(base_size = 13) +
  labs(
    title    = "Gene UMAP — Malignant → T-cell interactions",
    subtitle = sprintf("Leiden clustering (resolution = %.1f, k = %d)",
                       RESOLUTION, k),
    colour   = "Leiden\nCluster",
    x        = "UMAP 1",
    y        = "UMAP 2"
  )

print(p)
```



## GO x CCI
```{r}
# ══════════════════════════════════════════════════════════════════════════════
#  GENE REPRESENTATION IN EMITTER–RECEIVER–GO-BP SPACE
#  Bassez 2021 Breast Cancer · Malignant cell interactions
# ══════════════════════════════════════════════════════════════════════════════

# ── 0.  CONFIG ────────────────────────────────────────────────────────────────
LOGFC_AGG      <- "mean"       # "mean" | "sum"  — how to collapse LRIs per context
TOP_VAR_PCT    <- 0.05         # keep top X% GO-contexts by cross-gene variance (0.01 = 1%)
N_NEIGHBORS    <- 30           # k for KNN graph and UMAP
RESOLUTION     <- 1.2       # Leiden resolution
MIN_DIST       <- 0.1          # UMAP min_dist
SEED           <- 42
MALIGNANT_TYPE <- "Epithelial"

# ── 1.  PACKAGES ──────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(tibble)
library(uwot)
library(FNN)
library(igraph)
library(ggplot2)
library(ggrepel)
library(pheatmap)

# ── 2.  REFERENCE TABLES ──────────────────────────────────────────────────────
message("Loading reference tables ...")

LRI_GO    <- scDiffCom::LRI_human$LRI_curated_GO
GO_BP     <- scDiffCom::gene_ontology_level %>%
               filter(ASPECT == "biological_process") %>%
               select(ID, NAME)
LRI_GO_BP <- LRI_GO %>%
               inner_join(GO_BP, by = c("GO_ID" = "ID")) %>%
               rename(GO_NAME = NAME)

# ── 3.  INTERACTION DATA ──────────────────────────────────────────────────────
message("Extracting malignant interactions ...")

interactions_per_gene <- lapply(
  Bassez2021_Breast.scDiffComs,
  filter.scDiffCom.cci_table_detected.for.malignant,
  malignant_celltype = MALIGNANT_TYPE
)

interactions_all <- bind_rows(
  lapply(names(interactions_per_gene), function(g) {
    interactions_per_gene[[g]] %>% mutate(Gene = g)
  })
)

# ── 4.  BUILD EMITTER–RECEIVER–GO-BP CONTEXTS ─────────────────────────────────
# Key: GO_CONTEXT = GO_NAME__EMITTER__RECEIVER  (parsed later for display)
message("Joining with GO-BP annotations ...")

agg_fn <- match.fun(LOGFC_AGG)

GO_CCI <- interactions_all %>%
  select(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, LRI, LOGFC, Gene) %>%
  inner_join(LRI_GO_BP, by = "LRI", relationship = "many-to-many") %>%
  mutate(
    GO_CONTEXT = paste(GO_NAME, EMITTER_CELLTYPE, RECEIVER_CELLTYPE, sep = "__")
  )

# ── 5.  FILTER: TOP % HIGHLY VARIABLE GO-BP TERMS (on GO_CCI) ────────────────
message("Selecting top variable GO-BP terms ...")

# Step 1: aggregate per Gene × GO_NAME (collapsing all contexts/LRIs)
#         to get a pure GO-BP level signal per gene
gene_gobp_var <- GO_CCI %>%
  group_by(Gene, GO_NAME) %>%
  summarise(agg_LOGFC = agg_fn(LOGFC, na.rm = TRUE), .groups = "drop") %>%
  group_by(GO_NAME) %>%
  summarise(gobp_var = var(agg_LOGFC, na.rm = TRUE), .groups = "drop")

# Step 2: rank GO-BP terms by cross-gene variance, keep top %
var_cutoff    <- quantile(gene_gobp_var$gobp_var, probs = 1 - TOP_VAR_PCT, na.rm = TRUE)
top_gobp      <- gene_gobp_var %>%
  filter(gobp_var >= var_cutoff) %>%
  pull(GO_NAME)

message(sprintf(
  "Keeping %d / %d GO-BP terms (top %.0f%% by cross-gene variance)",
  length(top_gobp), nrow(gene_gobp_var), TOP_VAR_PCT * 100
))

# Step 3: filter GO_CCI to only rows whose GO_NAME is in the top set
GO_CCI_filt <- GO_CCI %>%
  filter(GO_NAME %in% top_gobp)

message(sprintf(
  "Retaining %d / %d GO-contexts belonging to those GO-BP terms",
  n_distinct(GO_CCI_filt$GO_CONTEXT), n_distinct(GO_CCI$GO_CONTEXT)
))

# ── 6.  GENE × GO-CONTEXT MATRIX (filtered) ───────────────────────────────────
message("Building gene × GO-context matrix ...")

gene_context_full <- GO_CCI_filt %>%
  group_by(Gene, GO_CONTEXT) %>%
  summarise(agg_LOGFC = agg_fn(LOGFC, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = GO_CONTEXT, values_from = agg_LOGFC, values_fill = 0)

gene_context_mat <- gene_context_full %>%
  column_to_rownames("Gene") %>%
  as.matrix()

# ── 7.  SCALE ─────────────────────────────────────────────────────────────────
mat_scaled <- scale(gene_context_mat)
mat_scaled[!is.finite(mat_scaled)] <- 0   # guard against zero-variance remnants

# ── 8.  KNN GRAPH ─────────────────────────────────────────────────────────────
set.seed(SEED)
k   <- min(N_NEIGHBORS, nrow(mat_scaled) - 1)
knn <- FNN::get.knn(mat_scaled, k = k)

n     <- nrow(mat_scaled)
edges <- do.call(rbind, lapply(seq_len(n), function(i) cbind(i, knn$nn.index[i, ])))

g <- igraph::graph_from_edgelist(edges, directed = FALSE) %>%
     igraph::simplify()

# ── 9.  LEIDEN CLUSTERING ─────────────────────────────────────────────────────
leiden_comm <- igraph::cluster_leiden(
  g,
  objective_function   = "modularity",
  resolution_parameter = RESOLUTION
)
clusters <- factor(leiden_comm$membership)
message(sprintf("Leiden found %d clusters", nlevels(clusters)))

# ── 10.  UMAP ─────────────────────────────────────────────────────────────────
set.seed(SEED)
umap_coords <- uwot::umap(
  mat_scaled,
  n_neighbors = k,
  min_dist    = MIN_DIST,
  metric      = "cosine"
)

umap_df <- data.frame(
  UMAP1  = umap_coords[, 1],
  UMAP2  = umap_coords[, 2],
  Leiden = clusters,
  Gene   = rownames(mat_scaled)
)

# ── 11.  UMAP PLOT ────────────────────────────────────────────────────────────
p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = Leiden, label = Gene)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(size = 2.8, max.overlaps = 20, show.legend = FALSE) +
  scale_colour_brewer(palette = "Set2") +
  theme_classic(base_size = 13) +
  labs(
    title    = sprintf("Gene UMAP — %s interactions", MALIGNANT_TYPE),
    subtitle = sprintf(
      "Emitter–Receiver–GO-BP space | top %.0f%% variable contexts | Leiden k=%d, res=%.1f",
      TOP_VAR_PCT * 100, k, RESOLUTION
    ),
    colour = "Leiden\nCluster",
    x = "UMAP 1", y = "UMAP 2"
  )
print(p_umap)
ggsave(
  filename = "gene-context-umap.png",
  plot     = p_umap,
  width    = 12,
  height   = 8,
  dpi      = 300
)

# ── 12.  HEATMAP ──────────────────────────────────────────────────────────────
library(gplots)
library(RColorBrewer)

# ── Parse GO_CONTEXT column names ─────────────────────────────────────────────
parse_context <- function(context_names) {
  parts <- strsplit(context_names, "__")
  data.frame(
    GO_NAME  = sapply(parts, `[`, 1),
    EMITTER  = sapply(parts, `[`, 2),
    RECEIVER = sapply(parts, `[`, 3),
    stringsAsFactors = FALSE
  )
}

col_meta <- parse_context(colnames(mat_scaled))

# ── Column labels: GO_NAME only, suppress repeated names ──────────────────────
n_col_labels <- 90
step         <- max(1, floor(ncol(mat_scaled) / n_col_labels))

col_labels <- rep("", ncol(mat_scaled))
# Among the thinned indices, show GO_NAME only on first occurrence
show_idx   <- seq(1, ncol(mat_scaled), by = step)
seen       <- c()
for (i in show_idx) {
  nm <- col_meta$GO_NAME[i]
  if (!nm %in% seen) {
    col_labels[i] <- nm
    seen <- c(seen, nm)
  }
}
# ── Single ColSideColors: Emitter→Receiver pair ────────────────────────────────
col_meta$ER_PAIR <- paste(col_meta$EMITTER, "→", col_meta$RECEIVER)

all_pairs <- unique(col_meta$ER_PAIR)
pair_pal  <- setNames(
  colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(all_pairs)),
  all_pairs
)
col_side_colors <- pair_pal[col_meta$ER_PAIR]   # single character vector ✓

# ── Row side bar: Leiden cluster ───────────────────────────────────────────────
n_clusters  <- nlevels(clusters)
cluster_pal <- setNames(
  RColorBrewer::brewer.pal(max(3, n_clusters), "Set2")[seq_len(n_clusters)],
  levels(clusters)
)
row_colors <- cluster_pal[as.character(clusters[rownames(mat_scaled)])]

# ── Colour scale ───────────────────────────────────────────────────────────────
heatmap_colors <- colorRampPalette(c("#2166ac", "white", "#d6604d"))(100)

# ── Draw ───────────────────────────────────────────────────────────────────────
png("gene-context-heatmap.png", width = 6000, height = 4000, res = 300)
par(oma = c(1, 1, 2, 1))

gplots::heatmap.2(
  mat_scaled,

  # ── labels ──────────────────────────────────────────────────
  labCol        = col_labels,
  labRow        = rownames(mat_scaled),
  cexCol        = 0.5,
  cexRow        = 0.65,
  srtCol        = 45,
  adjCol        = c(1, 0.5),

  # ── side bars ───────────────────────────────────────────────
  ColSideColors = col_side_colors,      # Emitter→Receiver pair colour bar
  RowSideColors = row_colors,           # Leiden cluster colour bar

  # ── clustering ──────────────────────────────────────────────
  dendrogram    = "both",
  distfun       = function(x) as.dist(1 - cor(t(x))),
  hclustfun     = function(d) hclust(d, method = "ward.D2"),

  # ── colour scale ────────────────────────────────────────────
  col           = heatmap_colors,
  symbreaks     = TRUE,
  breaks        = seq(-3, 3, length.out = 101),

  # ── chrome ──────────────────────────────────────────────────
  trace         = "none",
  density.info  = "none",
  margins       = c(16, 8),
  main          = sprintf(
    "Gene × Emitter–Receiver–GO-BP  |  top %.0f%% variable contexts",
    TOP_VAR_PCT * 100
  ),
  key.title     = "Scaled\nLOGFC",
  key.xlab      = "z-score"
)

# ── Legends ───────────────────────────────────────────────────────────────────
legend("topright",
  legend = all_pairs,
  fill   = pair_pal[all_pairs], border = NA,
  bty = "n", cex = 0.55, title = "Emitter → Receiver"
)

dev.off()
message("Heatmap written to gene-context-heatmap.png")

```


## GENES x LRI
```{r}
Bassez2021_Breast.LRI_LOGFC_list <- list()

for(gene in names(Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant)) {
  LOGFC_table <- Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant[[gene]] %>%
    select(LRI, LOGFC) %>%
    group_by(LRI) %>%
    summarise(LOGFC = mean(LOGFC, na.rm = TRUE), .groups = "drop")
  
  Bassez2021_Breast.LRI_LOGFC_list[[gene]] <- LOGFC_table
}

# prepare LOGFC matrix for heatmap
Bassez_LRI_LOGFC_all_data <- bind_rows(
  lapply(names(Bassez2021_Breast.LRI_LOGFC_list), function(g) {
    Bassez2021_Breast.LRI_LOGFC_list[[g]] %>% mutate(Gene = g)
  })
)

mat_long <- Bassez_LRI_LOGFC_all_data %>%
  pivot_wider(names_from = Gene, values_from = LOGFC)

mat <- as.matrix(mat_long[, -1])
rownames(mat) <- mat_long$CCI
  
message("  Matrix dimensions (CCIs x Genes): ", nrow(mat), " x ", ncol(mat))
  
n_non_na <- rowSums(!is.na(mat))
mat <- mat[n_non_na >= MIN_GENES_PER_CCI, , drop = FALSE]
message("  After filtering sparse CCIs: ", nrow(mat), " x ", ncol(mat))

sim_mat   <- cosine_sim_na(mat)
diag(sim_mat) <- NA

library(gplots)
# heatmap
sim_label <- "Cosine Similarity"
sim_file  <- paste0(DATASET, "_cosine_heatmap.png")
gene_gene_heatmap <- pheatmap(
    sim_mat,
    color          =  colorRampPalette(c("red", "white", "blue"))(100),
    breaks         = seq(-1, 1, length.out = 101),
    #na_col         = "grey90",
    main           = paste0(DATASET, " – Gene-Gene ", sim_label),
    fontsize_row   = 8,
    fontsize_col   = 8,
  )

n_genes <- nrow(sim_mat)
px_per_gene <- 20  # increase if still overlapping

png("gene-gene-lri-heatmap2.png", width = max(4000, n_genes * px_per_gene), 
                               height = max(4000, n_genes * px_per_gene), res = 300)
gplots::heatmap.2(
    sim_mat,
    dendrogram = "row",
    trace      = "none",
    col        = redgreen(100),
    margins    = c(14, 14),
    cexRow     = 0.6,
    cexCol     = 0.6,
    srtCol     = 45
)
dev.off()

png("gene-gene-lri-heatmap3.png", width = 4000, height = 4000, res = 300)
pheatmap(
    sim_mat,
    color        = colorRampPalette(c("red", "white", "blue"))(100),
    breaks       = seq(-1, 1, length.out = 101),
    fontsize_row = max(4, 200 / nrow(sim_mat)),  # auto-shrink based on count
    fontsize_col = max(4, 200 / nrow(sim_mat)),
    angle_col    = "45",
    main         = paste0(DATASET, " – Gene-Gene Cosine Similarity")
)
dev.off()
# ── UMAP of genes based on their LRI (CCI) LOGFC profiles ──────────────────
library(umap)
library(ggplot2)
library(ggrepel)

mat_genes <- t(mat)
mat_genes_imputed <- mat_genes
mat_genes_imputed[is.na(mat_genes_imputed)] <- 0

# ── Hierarchical clustering on cosine distance (matches your heatmap) ────────
k_clusters <- 5  # adjust to however many clusters make sense

dist_mat   <- as.dist(1 - sim_mat)   # cosine distance from your sim_mat
hclust_res <- hclust(dist_mat, method = "complete")
clusters   <- cutree(hclust_res, k = k_clusters)

# ── UMAP ─────────────────────────────────────────────────────────────────────
set.seed(42)
umap_config              <- umap.defaults
umap_config$n_neighbors  <- min(15, nrow(mat_genes_imputed) - 1)
umap_config$min_dist     <- 0.1
umap_config$metric       <- "cosine"

umap_result <- umap(mat_genes_imputed, config = umap_config)

umap_df <- data.frame(
  Gene    = rownames(mat_genes_imputed),
  UMAP1   = umap_result$layout[, 1],
  UMAP2   = umap_result$layout[, 2],
  n_CCIs  = rowSums(!is.na(mat_genes)),
  Cluster = factor(clusters[rownames(mat_genes_imputed)])  # match order safely
)

# ── Plot ──────────────────────────────────────────────────────────────────────
p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = Cluster, label = Gene)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    size          = 2.8,
    max.overlaps  = 40,
    segment.size  = 0.2,
    segment.alpha = 0.5,
    show.legend   = FALSE
  ) +
  scale_colour_brewer(palette = "Set1", name = "Cluster") +
  labs(
    title    = paste0(DATASET, " – Gene UMAP (LRI / CCI LOGFC space)"),
    subtitle = paste0("Hierarchical clustering (k = ", k_clusters, ", cosine distance)"),
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

ggsave(
  paste0(DATASET, "_lri_gene_umap.png"),
  plot   = p_umap,
  width  = 10,
  height = 8,
  dpi    = 300
)

message("UMAP saved to ", DATASET, "_lri_gene_umap.png")

# ── Print cluster membership so you can inspect them ─────────────────────────
cluster_df <- umap_df %>% 
  select(Gene, Cluster, n_CCIs) %>% 
  arrange(Cluster, Gene)
print(cluster_df)

cluster_summary <- cluster_df %>%
  group_by(Cluster) %>%
  summarise(
    n_genes = n(),
    genes   = paste(sort(Gene), collapse = ", ")
  ) %>%
  arrange(Cluster)

print(cluster_summary)
```

## AXL BOXPLOT
```{r}
library(ggplot2)
library(dplyr)

# 1. Prepare the specifically named dataframe
AXL_Breast_CCI_Boxplot_DF <- Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant[["AXL"]] %>% filter(
  EMITTER_CELLTYPE != "equivocal", RECEIVER_CELLTYPE != "equivocal"
) %>%
  mutate(ER_PAIR = paste(EMITTER_CELLTYPE, "→", RECEIVER_CELLTYPE))

# 2. Create the Plot
p_axl_boxplot <- ggplot(AXL_Breast_CCI_Boxplot_DF, aes(x = ER_PAIR, y = LOGFC, fill = ER_PAIR)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, color = "black") + 
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.6, aes(color = ER_PAIR)) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  theme_classic(base_size = 12) +
  labs(
    title = "AXL In Breast Cancer: L-R LogFC across Emitter-Receiver Pairs",
    subtitle = "Dataset: Bassez 2021 Breast Cancer",
    x = "Cell-Type Interaction (Emitter → Receiver)",
    y = "LogFC"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

# 3. Display
print(p_axl_boxplot)
file_name_boxplot <- "AXL_Breast_CCI_Boxplot.png"
ggsave(
  filename = file_name_boxplot,
  plot     = p_axl_boxplot,  # Using the plot object from the previous step
  width    = 10,             # Adjust width based on the number of ER pairs
  height   = 8,
  dpi      = 300,            # High resolution for thesis/presentation
  bg       = "white"         # Ensures background isn't transparent
)
```

## Cluster Genes x LRI, Combined
```{r}

library(harmony)
library(umap)

# --- 1. Helper Function ---
get_lri_matrix_long <- function(scDiffCom_list, dataset_name) {
  lapply(names(scDiffCom_list), function(gene) {
    scDiffCom_list[[gene]] %>%
      select(LRI, LOGFC) %>%
      group_by(LRI) %>%
      summarise(LOGFC = mean(LOGFC, na.rm = TRUE), .groups = "drop") %>%
      mutate(Gene = gene, Dataset = dataset_name, ID = paste0(gene, "_", dataset_name))
  }) %>% bind_rows()
}

# --- 2. Data Loading & Merging ---
df_breast <- get_lri_matrix_long(Bassez2021_Breast.scDiffComs.cci_table_detected.filtered.malignant, "Breast")
df_lung   <- get_lri_matrix_long(Chan2021_Lung.malignant, "Lung")
df_hnscc  <- get_lri_matrix_long(HNSCC_Atlas.malignant, "H&N")

# Find common LRIs (Features) across all cancer types
common_lris <- Reduce(intersect, list(
  unique(df_breast$LRI), 
  unique(df_lung$LRI), 
  unique(df_hnscc$LRI)
))

message("Number of common LRIs: ", length(common_lris))


# Combine and Filter
combined_long <- bind_rows(df_breast, df_lung, df_hnscc) %>%
  filter(LRI %in% common_lris) %>%
  mutate(ID = paste0(Gene, "_", Dataset))


library(tidyverse)
library(harmony)
library(umap)
library(ggrepel)
library(patchwork)

# --- 3. Scale and Compare Batches ---

# --- 3.1. Preparation & Baseline Matrix ---
# Assuming combined_long is already created from your previous steps
mat_wide <- combined_long %>%
  pivot_wider(id_cols = ID, names_from = LRI, values_from = LOGFC) %>%
  column_to_rownames("ID")

mat_final <- as.matrix(mat_wide)
mat_final[is.na(mat_final)] <- 0

meta_data <- data.frame(ID = rownames(mat_final)) %>%
  separate(ID, into = c("Gene", "Dataset"), sep = "_", remove = FALSE)

# Helper for UMAP plotting
plot_umap <- function(embeddings, meta, title) {
  df <- data.frame(
    UMAP1 = embeddings[,1], 
    UMAP2 = embeddings[,2], 
    ID = meta$ID, 
    Gene = meta$Gene, 
    Dataset = meta$Dataset
  )
  
  ggplot(df, aes(x = UMAP1, y = UMAP2, color = Dataset)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_text_repel(
      data = filter(df, Gene %in% c("AXL", "EGFR")),
      aes(label = ID), size = 3, fontface = "bold", color = "black", max.overlaps = 100
    ) +
    theme_minimal() +
    labs(title = title) +
    theme(legend.position = "bottom")
}

# --- METHOD 1: Scale only by LRI (Global Scaling) ---
mat_scaled_lri <- scale(mat_final)
umap_lri <- umap(mat_scaled_lri, config = umap.defaults)$layout
p1 <- plot_umap(umap_lri, meta_data, "1. Global LRI Scaling")

# --- METHOD 2: Scale by Dataset (Group-wise Scaling) ---
# This brings every dataset to the same distribution (mean 0, sd 1) independently
mat_scaled_ds <- do.call(rbind, lapply(unique(meta_data$Dataset), function(ds) {
  idx <- which(meta_data$Dataset == ds)
  # Scale the sub-matrix for this dataset
  scale(mat_final[idx, , drop = FALSE])
}))
# Re-order to match original metadata
mat_scaled_ds <- mat_scaled_ds[rownames(mat_final), ]
umap_ds <- umap(mat_scaled_ds, config = umap.defaults)$layout
p2 <- plot_umap(umap_ds, meta_data, "2. Per-Dataset Scaling")

# --- METHOD 3: Harmony Batch Correction ---
# Use the Globally Scaled matrix as input for Harmony
pca_res <- prcomp(mat_scaled_lri, rank. = 30)
harmony_emb <- HarmonyMatrix(
  pca_res$x, 
  meta_data, 
  "Dataset", 
  do_pca = FALSE, 
  verbose = FALSE
)
umap_harmony <- umap(harmony_emb, config = umap.defaults)$layout
p3 <- plot_umap(umap_harmony, meta_data, "3. Harmony Batch Correction")

# --- Combine Plots ---
(p1 | p2 | p3) + plot_layout(guides = 'collect') & theme(legend.position = 'bottom')
```
```{r}
# --- 1. Enhanced Plotting Function ---
plot_umap_clear <- function(embeddings, meta, title) {
  df <- data.frame(
    UMAP1 = embeddings[,1], 
    UMAP2 = embeddings[,2], 
    ID = meta$ID, 
    Gene = meta$Gene, 
    Dataset = meta$Dataset
  )
  
  ggplot(df, aes(x = UMAP1, y = UMAP2, color = Dataset)) +
    # Increased size and decreased alpha for better visibility of clusters
    geom_point(alpha = 0.6, size = 2) + 
    # Use geom_label_repel for a white background behind text (easier to read)
    geom_label_repel(
      data = filter(df, Gene %in% c("AXL", "EGFR")),
      aes(label = ID), 
      size = 4, 
      fontface = "bold",
      color = "black",
      box.padding = 0.8,      # More space around labels
      point.padding = 0.5,    # Space between point and label
      segment.color = 'grey50',
      max.overlaps = Inf,     # Force all labels to show
      fill = alpha(c("white"), 0.8)
    ) +
    theme_minimal(base_size = 14) + # Larger base font size
    labs(title = title) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      panel.grid.minor = element_blank() # Clean up the background
    )
}

# --- 2. Generate the Individual Plots ---
p1_v2 <- plot_umap_clear(umap_lri, meta_data, "1. Global LRI Scaling")
p2_v2 <- plot_umap_clear(umap_ds, meta_data, "2. Per-Dataset Scaling")
p3_v2 <- plot_umap_clear(umap_harmony, meta_data, "3. Harmony Integration")

# --- 3. Combine with Wide Layout ---
# Use plot_layout to give each plot plenty of horizontal breathing room
final_plot <- (p1_v2 | p2_v2 | p3_v2) + 
  plot_layout(guides = 'collect') & 
  theme(legend.position = 'bottom', legend.text = element_text(size = 14))

# --- 4. EXPORTING: The Secret to Clarity ---
# If you are in RStudio, the 'Plots' pane will always squash this.
# You MUST save it to a file or set chunk options to see the clarity.
ggsave(
  filename = "LRI_Integration_Comparison.png", 
  plot = final_plot, 
  width = 22,   # Very wide
  height = 8,   # Tall enough for labels
  dpi = 300,    # High resolution
  bg = "white"
)

library(plotly)

# 1. Create the dataframe specifically for the interactive plot
# Using the coordinates from the Harmony integration (umap_harmony)
df_interactive <- data.frame(
  UMAP1 = umap_harmony[, 1],
  UMAP2 = umap_harmony[, 2],
  ID = meta_data$ID,
  Gene = meta_data$Gene,
  Dataset = meta_data$Dataset
)

# 2. Build the ggplot object using this new dataframe
p_interact <- ggplot(df_interactive, aes(x = UMAP1, y = UMAP2, color = Dataset, text = ID)) +
  geom_point(alpha = 0.7, size = 1.5) +
  theme_minimal() +
  labs(
    title = "Interactive Harmony Integration",
    subtitle = "Hover over points to see the specific Gene_Dataset ID"
  )

# my.save.RData("p_interact", myfolder="~/Thesis/CCC/outputs/plots", is.load.obj = T)
# 3. Turn it into an interactive widget
# The 'tooltip = "text"' part tells plotly to show the 'ID' we defined in the aes()
ggplotly(p_interact, tooltip = "text")
```

## P-VALUE PER GENE, 
```{r}
# ══════════════════════════════════════════════════════════════════════════════
#  FISHER ENRICHMENT: UP/DOWN REGULATION PER GENE × EMITTER–RECEIVER–GO-BP
# ══════════════════════════════════════════════════════════════════════════════

# ── 0.  CONFIG ────────────────────────────────────────────────────────────────
MALIGNANT_TYPE  <- "Epithelial"
MIN_GO_SIZE     <- 5     # skip GO-BP with fewer than this many raw LRI-GO pairs
MIN_DETECTED    <- 2     # skip if fewer than this many DE interactions in GO
SAVE_EVERY      <- 1    # checkpoint every N genes
CHECKPOINT_FILE <- "fisher_results_checkpoint.rds"
FINAL_FILE      <- "fisher_results_final.rds"

# ── 1.  CORE FUNCTION: fisher for one gene ────────────────────────────────────
compute_fisher_for_gene <- function(sc_obj, gene_name, LRI_GO_BP,
                                    malignant_type  = MALIGNANT_TYPE,
                                    min_go_size     = MIN_GO_SIZE,
                                    min_detected    = MIN_DETECTED) {

  # ── 1a. Extract tables ─────────────────────────────────────────────────────
  # Use the shared helper for detected (handles IS_CCI_DE + finite LOGFC)
  detected <- filter.scDiffCom.cci_table_detected.for.malignant(
    sc_obj,
    malignant_celltype = malignant_type
  )

  # Mirror the same cell-type filter for raw (no DE/LOGFC conditions on raw)
  raw <- sc_obj@cci_table_raw %>%
    filter(
      EMITTER_CELLTYPE  == malignant_type |
      RECEIVER_CELLTYPE == malignant_type
    )

  if (nrow(detected) == 0) return(NULL)

  # ── 1b. Join GO annotations ────────────────────────────────────────────────
  raw_go <- raw %>%
    inner_join(LRI_GO_BP, by = "LRI", relationship = "many-to-many")

  detected_go <- detected %>%
    inner_join(LRI_GO_BP, by = "LRI", relationship = "many-to-many")

  if (nrow(detected_go) == 0) return(NULL)

  # ── 1c. Pre-compute background counts per (E, R, GO) ──────────────────────
  raw_counts <- raw_go %>%
    count(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, GO_NAME, name = "n_raw_in_go")

  raw_total_er <- raw_go %>%
    count(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, name = "n_raw_total")

  # ── 1d. Pre-compute detected counts per (E, R, GO, REGULATION) ────────────
  det_counts <- detected_go %>%
    count(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, GO_NAME, REGULATION,
          name = "n_det_in_go_reg")

  det_total_er_reg <- detected_go %>%
    count(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, REGULATION,
          name = "n_det_total_reg")

  # ── 1e. Build testable contexts ────────────────────────────────────────────
  contexts <- det_counts %>%
    filter(n_det_in_go_reg >= min_detected) %>%
    left_join(raw_counts,       by = c("EMITTER_CELLTYPE", "RECEIVER_CELLTYPE", "GO_NAME")) %>%
    left_join(raw_total_er,     by = c("EMITTER_CELLTYPE", "RECEIVER_CELLTYPE")) %>%
    left_join(det_total_er_reg, by = c("EMITTER_CELLTYPE", "RECEIVER_CELLTYPE", "REGULATION")) %>%
    filter(!is.na(n_raw_in_go), n_raw_in_go >= min_go_size)

  if (nrow(contexts) == 0) return(NULL)

  # ── 1f. Run Fisher per context ─────────────────────────────────────────────
  results <- vector("list", nrow(contexts))

  for (i in seq_len(nrow(contexts))) {
    a <- contexts$n_det_in_go_reg[i]
    b <- contexts$n_det_total_reg[i]   - a
    c <- contexts$n_raw_in_go[i]       - a
    d <- contexts$n_raw_total[i] - contexts$n_raw_in_go[i] - b

    if (any(c(a, b, c, d) < 0)) next

    ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2))

    results[[i]] <- data.frame(
      Gene              = gene_name,
      GO_NAME           = contexts$GO_NAME[i],
      EMITTER_CELLTYPE  = contexts$EMITTER_CELLTYPE[i],
      RECEIVER_CELLTYPE = contexts$RECEIVER_CELLTYPE[i],
      REGULATION        = contexts$REGULATION[i],
      odds_ratio        = ft$estimate,
      p_value           = ft$p.value,
      stringsAsFactors  = FALSE
    )
  }

  bind_rows(results)
}

# ── 2.  MAIN LOOP WITH CHECKPOINTING ─────────────────────────────────────────
gene_names  <- names(Bassez2021_Breast.scDiffComs)
n_genes     <- length(gene_names)

# Resume from checkpoint if it exists
if (file.exists(CHECKPOINT_FILE)) {
  checkpoint    <- readRDS(CHECKPOINT_FILE)
  completed     <- checkpoint$completed
  results_so_far <- checkpoint$results
  cat(sprintf("Resuming from checkpoint: %d / %d genes done\n",
              length(completed), n_genes))
} else {
  completed      <- character(0)
  results_so_far <- list()
}

genes_todo <- setdiff(gene_names, completed)
cat(sprintf("Genes remaining: %d\n", length(genes_todo)))

for (idx in seq_along(genes_todo)) {
  g <- genes_todo[idx]

  res <- tryCatch(
    compute_fisher_for_gene(
      sc_obj    = Bassez2021_Breast.scDiffComs[[g]],
      gene_name = g,
      LRI_GO_BP = LRI_GO_BP
    ),
    error = function(e) {
      cat(sprintf("  [!] Error on gene %s: %s\n", g, conditionMessage(e)))
      NULL
    }
  )

  if (!is.null(res) && nrow(res) > 0) {
    results_so_far[[g]] <- res
  }
  completed <- c(completed, g)

  # Progress (one line, not per-context noise)
  if (idx %% 10 == 0 || idx == length(genes_todo)) {
    cat(sprintf("  Progress: %d / %d genes  (%d total rows so far)\n",
                idx, length(genes_todo),
                sum(sapply(results_so_far, nrow))))
  }

  # Periodic checkpoint
  if (idx %% SAVE_EVERY == 0) {
    saveRDS(list(completed = completed, results = results_so_far),
            CHECKPOINT_FILE)
    cat(sprintf("  [checkpoint saved at gene %d]\n", idx))
  }
}

# ── 3.  COMBINE, BH-CORRECT, SAVE ─────────────────────────────────────────────
all_results <- bind_rows(results_so_far) %>%
  group_by(EMITTER_CELLTYPE, RECEIVER_CELLTYPE, GO_NAME, REGULATION) %>%
  mutate(bh_p_val = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  select(Gene, GO_NAME, EMITTER_CELLTYPE, RECEIVER_CELLTYPE,
         REGULATION, odds_ratio, p_value, bh_p_val) %>%
  arrange(bh_p_val)

saveRDS(all_results, FINAL_FILE)
cat(sprintf("Done. %d rows written to %s\n", nrow(all_results), FINAL_FILE))
```

## Volcano
```{r}

# ══════════════════════════════════════════════════════════════════════════════
#  VOLCANO PLOT: Emitter–Receiver–GO-BP  for selected genes
# ══════════════════════════════════════════════════════════════════════════════

# ── 0.  CONFIG ────────────────────────────────────────────────────────────────
SELECTED_GENES  <- c("ESR1","ERBB2", "HLA-A", "MKI67")   # <-- edit as needed
P_ADJ_CUTOFF    <- 0.05
OR_CUTOFF       <- 1.5    # odds ratio threshold for labelling
TOP_LABEL_N     <- 7     # max GO labels per facet to avoid overplotting
BOLD_GO_TERMS <- c("extracellular matrix organization", "angiogenesis", "	extracellular structure organization")

file_name <- paste(SELECTED_GENES, collapse="_")



# ── 1.  PREPARE DATA ──────────────────────────────────────────────────────────
fisher_res <- readRDS("~/Thesis/CellChat/scripts/fisher_results_final.rds")

fisher_res_sig <- fisher_res %>%
  filter(
    bh_p_val < P_ADJ_CUTOFF,
    is.finite(odds_ratio),
    is.finite(bh_p_val),
    REGULATION == "UP"
  ) %>%
  mutate(is_sig = bh_p_val < P_ADJ_CUTOFF)

# In volcano_df mutate — simplify is_sig to just p-value

cat(sprintf(
  "Significant results: %d / %d rows (%.1f%%)\n",
  nrow(fisher_res_sig), nrow(fisher_res),
  100 * nrow(fisher_res_sig) / nrow(fisher_res)
))


volcano_df <- fisher_res_sig %>%
  filter(
    Gene    %in% SELECTED_GENES,
  ) %>%
  mutate(
    ER_PAIR    = paste(EMITTER_CELLTYPE, "→", RECEIVER_CELLTYPE),
    log2_OR    = log2(odds_ratio),                 # ← compute here
    neg_log_p  = -log10(bh_p_val),
    label_text = ifelse(is_sig, GO_NAME, "")
  ) %>%
  group_by(Gene) %>%
  mutate(
    rank_p     = rank(-neg_log_p, ties.method = "first"),
    label_text = ifelse(is_sig & rank_p <= TOP_LABEL_N, GO_NAME, "")
  ) %>%
  ungroup()

volcano_df <- volcano_df %>%
  mutate(
    fontface = ifelse(GO_NAME %in% BOLD_GO_TERMS, "bold", "plain")
  )

# ── 2.  COLOUR PALETTE ────────────────────────────────────────────────────────
all_pairs <- unique(volcano_df$ER_PAIR)
n_pairs   <- length(all_pairs)
er_pal    <- setNames(
  colorRampPalette(RColorBrewer::brewer.pal(min(8, n_pairs), "Dark2"))(n_pairs),
  all_pairs
)

# ── 3.  PLOT ──────────────────────────────────────────────────────────────────
p_volcano <- ggplot(
    volcano_df,
    aes(x = log2_OR, y = neg_log_p, colour = ER_PAIR)
  ) +

  # ── reference line: significance only, no OR cutoff lines ─────────────────
  geom_hline(yintercept = -log10(P_ADJ_CUTOFF),
             linetype = "dashed", colour = "grey50", linewidth = 0.4) +

  # ── non-significant points ─────────────────────────────────────────────────
  geom_point(
    data    = filter(volcano_df, !is_sig),
    colour  = "grey80", size = 1, alpha = 0.5
  ) +

  # ── significant points ─────────────────────────────────────────────────────
  geom_point(
    data = filter(volcano_df, is_sig),
    aes(colour = ER_PAIR),
    size = 2.2, alpha = 0.85
  ) +

  # ── GO-BP labels ───────────────────────────────────────────────────────────
  ggrepel::geom_text_repel(

    data          = filter(volcano_df, label_text != ""),

    aes(label     = label_text, colour = ER_PAIR, fontface = fontface),

    size          = 2.4,

    max.overlaps  = 20,

    segment.size  = 0.3,

    segment.alpha = 0.5,

    show.legend   = FALSE,

    box.padding   = 0.3

  ) + 



facet_wrap(~ Gene, scales = "free_y", nrow = 2) +


  scale_colour_manual(values = er_pal) +

  scale_x_continuous(

    name   = expression(log[2]~"(Odds Ratio)"),

    labels = scales::number_format(accuracy = 0.1)

  ) +

  scale_y_continuous(

    name = expression(-log[10]~"(BH p-value)")

  ) +



  theme_classic(base_size = 12) +

  theme(

    strip.text       = element_text(face = "bold", size = 13),

    strip.background = element_blank(),

    legend.position  = "bottom",

    legend.title     = element_text(face = "bold"),

    legend.text      = element_text(size = 8),

    panel.grid.major = element_line(colour = "grey95")

  ) +

  guides(colour = guide_legend(

    title        = "Emitter → Receiver",

    nrow         = ceiling(n_pairs / 3),

    override.aes = list(size = 3, alpha = 1)

  )) +

  labs(

    title    = "GO-BP Enrichment Volcano",

    subtitle = sprintf("BH p < %.2f  |  top %d GO labels per gene",

                       P_ADJ_CUTOFF, TOP_LABEL_N)

  )
print(p_volcano)

# ── 4.  SAVE ──────────────────────────────────────────────────────────────────
ggsave(
  filename = paste0(file_name, "_volcano.png"),
  plot     = p_volcano,
  width    = 8,
  height   = 11,
  dpi      = 300
)

for (current_gene in SELECTED_GENES) {
  
  # Subset for the specific gene
  gene_data <- volcano_df %>% filter(Gene == current_gene)
  
  p_individual <- ggplot(
    gene_data, 
    aes(x = log2_OR, y = neg_log_p, colour = ER_PAIR)
  ) +
    # Reference line for significance
    geom_hline(yintercept = -log10(P_ADJ_CUTOFF), 
               linetype = "dashed", colour = "grey50", linewidth = 0.4) +
    
    # Points
    geom_point(size = 2.8, alpha = 0.85) +
    
    # Labels
    ggrepel::geom_text_repel(
      data = filter(gene_data, label_text != ""),
      aes(label = label_text, fontface = fontface),
      size = 3, max.overlaps = 20, segment.size = 0.3
    ) +
    
    # Apply global colors and hide the legend
    scale_colour_manual(values = er_pal) +
    
    labs(
      title = paste(current_gene),
      x = expression(log[2]~"(Odds Ratio)"),
      y = expression(-log[10]~"(BH p-value)")
    ) +
    
    theme_classic(base_size = 12) +
    theme(
      legend.position = "none",  # <--- This removes the legend
      plot.title = element_text(face = "bold", size = 14),
      panel.grid.major = element_line(colour = "grey95")
    )

  # Save individual file
  ggsave(
    filename = paste0(current_gene, "_volcano_no_legend.png"),
    plot     = p_individual,
    width    = 7,
    height   = 7,
    dpi      = 300
  )
}
```




# Playground - Check objects after running pipeline
```{r}

```

