---
title: "scDiffCom-Preprocessing-Checks"
output: html_document
---

# Load Seurats
```{r}
Load.Seurat.With.Patients.Data <- function(ds_path, ps_path, gene_col, patient_col_in_seurat) {
     message("Loading raw Seurat object...")
    ext <- tolower(tools::file_ext(ds_path))
    if (ext == "rds") {
      obj <- readRDS(ds_path)
    } else if (ext %in% c("rdata", "rda")) {
      miceadds::load.Rdata(ds_path, "obj")
    } else {
      stop(paste("Unsupported file format:", ext, ". Supported formats: .rds, .RData, .rda"))
    }
    
    message("Loading patient summary (preprocess output)...")
    ps <- readRDS(ps_path) 
    
    if (!patient_col_in_seurat %in% colnames(obj@meta.data)) {
      stop(sprintf("Error: Column '%s' not found in Seurat metadata.", patient_col_in_seurat))
    }
    
    message(sprintf("Merging %s classifications...", gene_col))
    
    # Map patient IDs to their HIGH/LOW/MID status
    mapping_vec <- setNames(ps[[gene_col]], ps$patient_id)
    
    # Apply mapping to the Seurat metadata
    obj@meta.data[[gene_col]] <- mapping_vec[as.character(obj@meta.data[[patient_col_in_seurat]])]
    
    return(list(
      seurat = obj, 
      patient_summary = as.data.frame(ps)
    ))
}
```

# Load Data
```{r}
# Qian.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Qian2020_Breast_step4.RDS",
#                                             ps_path = "~/outputs/Qian2020_Breast_step4/AXL_Qian2020_Breast_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")
# 
# Chan.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Chan2021_Lung_step4.RDS",
#                                             ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Chan2021_Lung_step4/AXL_Chan2021_Lung_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")
# 
# Laughney.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Laughney2020_Lung_step4.RDS",
#                                             ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Laughney2020_Lung_step4/AXL_Laughney2020_Lung_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")
# 
# Bassez.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Bassez2021_Breast_step4.RDS",
#                                             ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Bassez2021_Breast_step4/AXL_Bassez2021_Breast_step4_grouped.rds",
#                                             gene_col ="AXL_EXP",
#                                             patient_col_in_seurat = "ident")

Bill_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Bill_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Bill_HNSC/AXL_Bill_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")
Choi_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Choi_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Choi_HNSC/AXL_Choi_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")
Kurten_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Kurten_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Kurten_HNSC/AXL_Kurten_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")
Puram_HNSC.AXL <- Load.Seurat.With.Patients.Data(ds_path = "/gpfs0/bgu-ofircohen/users/gigiadir/scObjects/Puram_HNSC.RData",
                                              ps_path = "/gpfs0/bgu-ofircohen/users/gigiadir/CCC-PreProcess/results/Puram_HNSC/AXL_Puram_HNSC_grouped.rds",
                                              gene_col ="AXL_EXP",
                                              patient_col_in_seurat = "Patient")

```

# Check Seurat object Cell-Type-Distribution
```{r}

library(dplyr)
library(ggplot2)
library(Seurat)

plot_patient_composition <- function(obj, patient_summary, gene_col_exp_name, ds_name,
                                     cell_type_column = "Consensus_Cell_Type",
                                     patient_col_in_seurat = "ident",
                                     use_proportions = FALSE) {

  gene_col <- grep("_EXP$", colnames(patient_summary), value = TRUE)

  # Pull metadata, extract patient column
  meta <- obj@meta.data
  meta$patient <- as.character(meta[[patient_col_in_seurat]])
  meta <- meta[!is.na(meta$patient), ]

  # Order all patients by mean_expr descending; patients not in summary sort last
  all_patients <- unique(meta$patient)
  ordered_patients <- data.frame(patient = all_patients, stringsAsFactors = FALSE) %>%
    left_join(patient_summary[, c("patient_id", "mean_expr")],
              by = c("patient" = "patient_id")) %>%
    arrange(desc(mean_expr)) %>%
    pull(patient)

  # Build x-axis labels: "patient\nHIGH/MID/LOW" for labeled patients, else just patient
  label_map <- setNames(patient_summary[[gene_col]], patient_summary$patient_id)
  x_labels <- setNames(
    ifelse(ordered_patients %in% names(label_map),
           paste0(ordered_patients, "\n", label_map[ordered_patients]),
           ordered_patients),
    ordered_patients
  )

  cell_counts <- meta %>%
    dplyr::filter(!is.na(.data[[cell_type_column]])) %>%
    # Rename the dynamic column to "cell_type" to safely pass it down the pipe
    dplyr::rename(cell_type = dplyr::all_of(cell_type_column)) %>%
    # Explicitly use dplyr::count to prevent package masking
    dplyr::count(patient, cell_type) %>%
    dplyr::group_by(patient) %>%
    dplyr::mutate(prop = n / sum(n)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(cell_type = factor(cell_type,
                                     levels = c("Epithelial",
                                                    sort(setdiff(unique(cell_type), "Epithelial")))))

  y_var   <- if (use_proportions) "prop" else "n"
  y_label <- if (use_proportions) "Proportion" else "Cell Count"

  p <- ggplot(cell_counts, aes(x = patient, y = .data[[y_var]], fill = cell_type)) +
    geom_bar(stat = "identity")

  if (!use_proportions)
    p <- p + geom_hline(yintercept = 50, linetype = "dashed", alpha = 0.4) +
             scale_y_continuous(breaks = function(x) sort(unique(c(pretty(x), 50))))

  p +
    scale_x_discrete(limits = ordered_patients, labels = x_labels) +
    labs(x = "Patient", y = y_label,
         title = paste(ds_name, "Cell Type Distribution:", gene_col_exp_name),
         fill = "Cell Type") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.ticks.x = element_blank(),
      legend.title = element_text(size = 10),
      panel.grid = element_blank(),
      panel.background = element_blank()
    )
}

# Qian.AXL.plot <- plot_patient_composition(obj = Qian.AXL$seurat,
#                                            patient_summary =Qian.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Qian2020 - Breast",
#                                            patient_col_in_seurat = "ident",
#                                           cell_type_column = "cell_type")
# 
# Chan.AXL.plot <- plot_patient_composition(obj = Chan.AXL$seurat,
#                                            patient_summary =Chan.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Chan2021 - Lung",
#                                            patient_col_in_seurat = "ident",
#                                           cell_type_column = "cell_type")
# 
# Laughney.AXL.plot <- plot_patient_composition(obj = Laughney.AXL$seurat,
#                                            patient_summary =Laughney.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Laughney2020 - Lung",
#                                            patient_col_in_seurat = "ident",
#                                            cell_type_column = "cell_type")
# 
# Bassez.AXL.plot <- plot_patient_composition(obj = Bassez.AXL$seurat,
#                                            patient_summary =Bassez.AXL$patient_summary,
#                                            gene_col_exp_name = "AXL_EXP",
#                                            ds_name = "Bassez2021 - Breast",
#                                            patient_col_in_seurat = "ident",
#                                            cell_type_column = "cell_type",
#                                            use_proportions = TRUE)

Kurten_HNSC.AXL.plot <- plot_patient_composition(obj = Kurten_HNSC.AXL$seurat,
                                           patient_summary =Kurten_HNSC.AXL$patient_summary,
                                           gene_col_exp_name = "AXL_EXP",
                                           ds_name = "Kurten - H&N",
                                           patient_col_in_seurat = "Patient",
                                           cell_type_column = "Cell_Type",
                                           use_proportions = TRUE)

out_dir <- "~/Thesis/CCC/outputs/scDiffCom/QC"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ggsave(file.path(out_dir, "Qian2020_Breast_AXL_cell_type_distribution_original_cell_type.png"),    Qian.AXL.plot,    width = 20, height = 6)
# ggsave(file.path(out_dir, "Chan2021_Lung_AXL_cell_type_distribution.png"),      Chan.AXL.plot,    width = 20, height = 6)
# ggsave(file.path(out_dir, "Laughney2020_Lung_AXL_cell_type_distribution.png"),  Laughney.AXL.plot, width = 20, height = 6)
# ggsave(file.path(out_dir, "Bassez2021_Breast_AXL_cell_type_distribution.png"),  Bassez.AXL.plot,  width = 20, height = 6)
ggsave(file.path(out_dir, "Kurten_HNSC_AXL_cell_Type_distribution.png"), Kurten_HNSC.AXL.plot, width = 20, height = 6)
```


# Merge + ComBat Batch Correction + UMAP Integration
```{r merge-combat-umap, message=FALSE, warning=FALSE}
# Dependencies ─────────────────────────────────────────────────────────────────
if (!requireNamespace("sva",       quietly = TRUE)) stop("BiocManager::install('sva')")
if (!requireNamespace("patchwork", quietly = TRUE)) stop("install.packages('patchwork')")
if (!requireNamespace("Matrix",    quietly = TRUE)) stop("install.packages('Matrix')")

library(Seurat)
library(sva)
library(ggplot2)
library(dplyr)
library(patchwork)
library(Matrix)

# Config ───────────────────────────────────────────────────────────────────────
SEURAT_DIR <- path.expand("~/scObjects")
OUTPUT_DIR <- path.expand("~/Thesis/CCC/outputs/scDiffCom/integration")
N_FEATURES <- 3000
N_PCS      <- 30
CT_COL     <- "Cell_Type"
SEURAT_V5  <- utils::packageVersion("Seurat") >= "5.0.0"
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# ── 1. Load objects and stamp source label ────────────────────────────────────
stamp_source <- function(fname, label) {
   message("Loading raw Seurat object...")
    path <- file.path(SEURAT_DIR, fname)
    ext <- tolower(tools::file_ext(path))
    if (ext == "rds") { 
      obj <- readRDS(path)
    } else if (ext %in% c("rdata", "rda")) {
      miceadds::load.Rdata(path, "obj")
    } else {
      stop(paste("Unsupported file format:", ext, ". Supported formats: .rds, .RData, .rda"))
    }
  obj$source <- label
  message("Loaded ", fname, " — ", ncol(obj), " cells")
  obj
}

HNSC.Atlas <- stamp_source("HNSCC.Atlas.RData", "HNSC_Atlas")
puram <- stamp_source("Puram_HNSC.RData", "Puram_HNSC")
choi   <- stamp_source("Choi_HNSC.RData",    "Choi_HNSC")
kurten <- stamp_source("Kurten_HNSC.RData",  "Kurten_HNSC")
bill   <- stamp_source("Bill_HNSC.RData",    "Bill_HNSC")

# ── 2. Individual UMAPs (runs full Seurat pipeline if UMAP absent) ────────────
ensure_umap <- function(obj, label) {
  if ("umap" %in% names(obj@reductions)) {
    message(label, ": UMAP already present — skipping preprocessing.")
    return(obj)
  }
  message(label, ": no UMAP found — running NormalizeData → PCA → UMAP ...")
  obj <- NormalizeData(obj, verbose = FALSE)
  obj <- FindVariableFeatures(obj, nfeatures = N_FEATURES, verbose = FALSE)
  obj <- ScaleData(obj, verbose = FALSE)
  obj <- RunPCA(obj, npcs = N_PCS, verbose = FALSE)
  obj <- FindNeighbors(obj, dims = 1:N_PCS, verbose = FALSE)
  obj <- RunUMAP(obj, dims = 1:N_PCS, verbose = FALSE)
  obj
}
puram  <- ensure_umap(puram,  "Puram_HNSC")
choi   <- ensure_umap(choi,   "Choi_HNSC")
kurten <- ensure_umap(kurten, "Kurten_HNSC")
bill   <- ensure_umap(bill,   "Bill_HNSC")

ind_umap_plot <- function(obj, label) {
  col <- if (CT_COL %in% colnames(obj@meta.data)) CT_COL else "seurat_clusters"
  DimPlot(obj, reduction = "umap", group.by = col, label = TRUE, repel = TRUE,
          pt.size = 0.3) +
    ggtitle(label) +
    theme(plot.title      = element_text(face = "bold", size = 11),
          legend.text     = element_text(size = 7),
          legend.key.size = unit(0.35, "cm"))
}
p_ind <- ((ind_umap_plot(puram,  "Puram_HNSC")  |
           ind_umap_plot(choi,   "Choi_HNSC")) /
          (ind_umap_plot(kurten, "Kurten_HNSC") |
           ind_umap_plot(bill,   "Bill_HNSC"))) +
  plot_annotation(
    title = "Individual UMAPs — H&N Datasets",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  )
print(p_ind)
ggsave(file.path(OUTPUT_DIR, "01_individual_umaps.png"),
       p_ind, width = 18, height = 14, dpi = 300)
message("Saved 01_individual_umaps.png")

# ── 3. Merge ──────────────────────────────────────────────────────────────────
message("\nMerging Puram_HNSC, Choi_HNSC, Kurten_HNSC, Bill_HNSC ...")
merged <- merge(puram, y = list(choi, kurten, bill),
                add.cell.ids = c("Puram_HNSC", "Choi_HNSC", "Kurten_HNSC", "Bill_HNSC"))
message("Merged object: ", ncol(merged), " cells")

# ── 4. Normalize and select variable features (join layers first for v5) ──────
DefaultAssay(merged) <- "RNA"
if (SEURAT_V5) merged <- JoinLayers(merged)
merged <- NormalizeData(merged, verbose = FALSE)
merged <- FindVariableFeatures(merged, nfeatures = N_FEATURES, verbose = FALSE)
var_genes <- VariableFeatures(merged)
message("Variable features selected: ", length(var_genes))

# ── 5. Extract log-normalised matrix for ComBat ───────────────────────────────
# ComBat operates on log-normalised values; corrected matrix is stored as a
# separate 'ComBat' assay so the original RNA assay is preserved intact.
log_mat <- as.matrix(
  if (SEURAT_V5) GetAssayData(merged, assay = "RNA", layer = "data")[var_genes, ]
  else           GetAssayData(merged, assay = "RNA", slot  = "data")[var_genes, ]
)
batch_vec <- as.character(merged$source)
message("Batch sizes: ",
        paste(paste0(names(table(batch_vec)), "=", table(batch_vec)), collapse = ", "))

# Drop genes with zero variance in any single batch (ComBat cannot handle these)
batch_var <- do.call(cbind, lapply(unique(batch_vec), function(b)
  apply(log_mat[, batch_vec == b, drop = FALSE], 1, var)
))
keep        <- rowSums(batch_var == 0) == 0
var_genes_f <- var_genes[keep]
log_mat_f   <- log_mat[keep, ]
message("Removed ", sum(!keep), " zero-variance gene(s); ",
        sum(keep), " genes retained for ComBat.")

# ── 6. ComBat batch correction ────────────────────────────────────────────────
message("Running sva::ComBat (parametric prior) on ",
        nrow(log_mat_f), " genes x ", ncol(log_mat_f), " cells ...")
combat_mat <- sva::ComBat(dat        = log_mat_f,
                           batch      = batch_vec,
                           par.prior  = TRUE,
                           prior.plots = FALSE)
message("ComBat complete.")

# ── 7. Store corrected matrix as new 'ComBat' assay ───────────────────────────
sparse_cb          <- Matrix::Matrix(combat_mat, sparse = TRUE)
merged[["ComBat"]] <- tryCatch(
  CreateAssay5Object(data = sparse_cb),          # Seurat >= 5
  error = function(e) CreateAssayObject(data = sparse_cb)   # Seurat v4 fallback
)
DefaultAssay(merged)     <- "ComBat"
VariableFeatures(merged) <- var_genes_f

# ── 8. Scale → PCA → UMAP on ComBat-corrected assay ──────────────────────────
message("ScaleData → RunPCA → RunUMAP on ComBat assay ...")
merged <- ScaleData(merged,      features       = var_genes_f,       verbose = FALSE)
merged <- RunPCA(merged,         features       = var_genes_f,
                 npcs            = N_PCS,
                 reduction.name  = "pca_combat",                     verbose = FALSE)
merged <- FindNeighbors(merged,  reduction      = "pca_combat",
                         dims   = 1:N_PCS,                           verbose = FALSE)
merged <- FindClusters(merged,                                        verbose = FALSE)
merged <- RunUMAP(merged,        reduction      = "pca_combat",
                  dims           = 1:N_PCS,
                  reduction.name = "umap_combat",                    verbose = FALSE)
message("Dimensionality reduction complete.")

# ── 9. Integrated UMAP — Consensus Cell Type ─────────────────────────────────
ct_col_use <- if (CT_COL %in% colnames(merged@meta.data)) CT_COL else "seurat_clusters"
p_ct <- DimPlot(merged, reduction = "umap_combat", group.by = ct_col_use,
                label = TRUE, repel = TRUE, pt.size = 0.25) +
  ggtitle(paste("Integrated UMAP —", ct_col_use)) +
  theme(plot.title      = element_text(face = "bold"),
        legend.text     = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"))

# ── 10. Integrated UMAP — Source Dataset (batch-correction quality check) ─────
src_pal <- c(Puram_HNSC  = "#1B9E77",
             Choi_HNSC   = "#D95F02",
             Kurten_HNSC = "#7570B3",
             Bill_HNSC   = "#E7298A")
p_src <- DimPlot(merged, reduction = "umap_combat", group.by = "source",
                 cols = src_pal, pt.size = 0.25) +
  ggtitle("Integrated UMAP — Source Dataset") +
  theme(plot.title = element_text(face = "bold"))

p_combined <- (p_ct / p_src) +
  plot_annotation(
    title    = "ComBat-Corrected H&N Integration",
    subtitle = sprintf("%d cells | %d variable features | %d PCs",
                       ncol(merged), length(var_genes_f), N_PCS),
    theme    = theme(
      plot.title    = element_text(size = 14, face = "bold",  hjust = 0.5),
      plot.subtitle = element_text(size = 10, colour = "grey40", hjust = 0.5)
    )
  )

print(p_ct)
print(p_src)
print(p_combined)

ggsave(file.path(OUTPUT_DIR, "02_integrated_umap_celltype.png"),
       p_ct,       width = 12, height = 9,  dpi = 300)
ggsave(file.path(OUTPUT_DIR, "03_integrated_umap_source.png"),
       p_src,      width = 10, height = 8,  dpi = 300)
ggsave(file.path(OUTPUT_DIR, "04_integrated_umap_combined.png"),
       p_combined, width = 12, height = 18, dpi = 300)
message("Saved integrated UMAP plots (02, 03, 04).")

# ── 11. Save final ComBat-corrected Seurat object ─────────────────────────────
out_rds <- file.path(OUTPUT_DIR, "HNSC_ComBat_integrated.rds")
saveRDS(merged, out_rds)
message("\nFinal object saved to: ", out_rds)
message("Assays     : ", paste(Assays(merged),              collapse = ", "))
message("Reductions : ", paste(names(merged@reductions),    collapse = ", "))
message("Cells      : ", ncol(merged))
message("Metadata   : ", paste(colnames(merged@meta.data),  collapse = ", "))
```



#Check patient distribution
```{r}
library(tidyr)
# opt$column_id <- "ident_source"
merged.results <- helpers$add_malignant_gene_group(seurat_obj = HNSC.Atlas,
                                            gene = "CTLA4",
                                            cell_type_col = "Cell_Type",
                                            assay="RNA")

merged.results.ps <- merged.results$patient_summary

# This creates a 2D grid of counts
counts_table <- table(merged.results.ps$CTLA4_EXP, merged.results.ps$Project)
print(counts_table)

```



## Data validation: expression category distribution per patient

```{r gene-expression-category-distribution, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

# For each dataset, calculate and plot the distribution of HIGH/MID/LOW calls per patient
# across all analyzed genes.
BASE_RESULTS_DIR <- path.expand("~/CCC-PreProcess/results")


dataset_names <- c("Puram_HNSC", "Choi_HNSC", "Kurten_HNSC", "Bill_HNSC")

if (length(dataset_names) == 0) {
  message("No dataset names found (no *.AXL objects, and results dir missing). Skipping.")
} else {
  expr_levels <- c("LOW", "MID", "HIGH")
  expr_pal <- c(LOW = "#4E79A7", MID = "#F28E2B", HIGH = "#E15759")

  read_grouped_file <- function(rds_path, dataset_name) {
    df <- tryCatch(readRDS(rds_path), error = function(e) NULL)
    if (is.null(df) || !is.data.frame(df) || ncol(df) != 3) return(NULL)

    # Expect columns: patient_id, mean_expr, {gene}_exp (dynamic)
    exp_col <- colnames(df)[3]
    gene <- sub(paste0("_", dataset_name, "_grouped\\.rds$"), "", basename(rds_path))

    df %>%
      dplyr::select(patient_id, dplyr::all_of(exp_col)) %>%
      dplyr::rename(expression_level = dplyr::all_of(exp_col)) %>%
      dplyr::mutate(
        gene = gene,
        expression_level = as.character(expression_level)
      )
  }

  for (ds in dataset_names) {
    ds_dir <- file.path(BASE_RESULTS_DIR, ds)
    if (!dir.exists(ds_dir)) {
      message("Dataset directory not found: ", ds_dir, " (skipping)")
      next
    }

    rds_files <- list.files(
      ds_dir,
      pattern = paste0("_", ds, "_grouped\\.rds$"),
      full.names = TRUE
    )

    if (length(rds_files) == 0) {
      message("No grouped .rds files found for dataset: ", ds, " (skipping)")
      next
    }

    agg <- rds_files %>%
      lapply(read_grouped_file, dataset_name = ds) %>%
      Filter(Negate(is.null), .) %>%
      bind_rows()

    if (nrow(agg) == 0) {
      message("No readable/valid files for dataset: ", ds, " (skipping)")
      next
    }

    count_df <- agg %>%
      dplyr::filter(!is.na(patient_id), !is.na(expression_level)) %>%
      dplyr::mutate(
        patient_id = as.character(patient_id),
        expression_level = factor(expression_level, levels = expr_levels)
      ) %>%
      dplyr::count(patient_id, expression_level, name = "n") %>%
      tidyr::complete(patient_id, expression_level, fill = list(n = 0))

    p <- ggplot(count_df, aes(x = patient_id, y = n, fill = expression_level)) +
      geom_col(
        position = position_dodge2(width = 0.9, preserve = "single"),
        width = 0.8,
        color = "white",
        linewidth = 0.15
      ) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
      scale_fill_manual(values = expr_pal, drop = FALSE) +
      labs(
        title = paste0(ds, ": Gene expression category distribution per patient (Expression Tertile)"),
        x = "patient_id",
        y = "Number of analyzed genes",
        fill = "Expression\ncategory"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.major.x = element_blank()
      )

    print(p)

    qc_plot_dir <- path.expand("~/Thesis/CCC/outputs/plots/QC")
    dir.create(qc_plot_dir, recursive = TRUE, showWarnings = FALSE)
    ggsave(
      filename = file.path(qc_plot_dir, paste0(ds, "_expression_category_distribution_per_patient.png")),
      plot = p,
      width = 10,
      height = 6,
      dpi = 300
    )
  }
}
```




# Build pseudobulk matrices (malignant cells, detection-filtered genes)

For each dataset, **malignant cells** are defined by a configurable metadata column and value set (e.g. HNSC: `Cell_Type == "Tumor"`; other cohorts: e.g. `Consensus_Cell_Type` in `"Epithelial"`). The object is restricted to those cells with **`subset()`** (S3 method for Seurat objects; `cells = ...`) before computing detection and means. Genes kept are those **detected** (non-zero **counts**) in **more than 5%** of malignant cells (threshold configurable). The matrix entry is the **mean normalized expression** (`data` layer / slot) of that gene **within malignant cells only**, averaged separately per patient. Rows = genes passing the filter, columns = patients. Objects are loaded from **`~/scObjects`** via **`MATRIX_DATASETS`**.

```{r build-gene-patient-matrices, message=TRUE, warning=FALSE}
library(dplyr)
library(purrr)
library(Seurat)
library(Matrix)

SAVE_MATRICES <- TRUE
SAVE_PER_DATASET_RDS <- TRUE
SAVE_ALL_AS_RDATA <- TRUE
OUT_DIR <- path.expand("~/Thesis/CCC/outputs/RData_objects/pseudobulk_matrix")

# One file per dataset under ~/scObjects: {dataset_name}.rds / .RDS / .RData / .rda
SCOBJECTS_DIR <- path.expand("~/scObjects")
MATRIX_DATASETS <- c(
  "Kurten_HNSC",
  "Bill_HNSC",
  "Puram_HNSC",
  "Choi_HNSC"
)

# Malignant definition per dataset (metadata column + value(s) matched with %in%, as character).
# Add or edit entries when cell-type column / malignant label differs (e.g. Consensus_Cell_Type + Epithelial).
MALIG_CELL_CONFIG <- list(
  Kurten_HNSC = list(metadata_col = "Cell_Type", malignant_values = "Tumor"),
  Bill_HNSC   = list(metadata_col = "Cell_Type", malignant_values = "Tumor"),
  Puram_HNSC  = list(metadata_col = "Cell_Type", malignant_values = "Tumor"),
  Choi_HNSC   = list(metadata_col = "Cell_Type", malignant_values = "Tumor")
  # Example (other cohorts):
  # Some_dataset = list(metadata_col = "Consensus_Cell_Type", malignant_values = c("Epithelial"))
)

# Minimum fraction of malignant cells with counts > 0 required to retain a gene
MIN_DETECT_FRAC_IN_MALIGNANT <- 0.05

SEURAT_BY_DATASET <- list(
  Kurten_HNSC = Kurten_HNSC,
  Bill_HNSC = Bill_HNSC,
  Puram_HNSC = Puram_HNSC,
  Choi_HNSC = Choi_HNSC
)

PATIENT_COL_BY_DATASET <- c(
  Kurten_HNSC = "Patient",
  Bill_HNSC   = "Patient",
  Puram_HNSC  = "Patient",
  Choi_HNSC   = "Patient"
)

get_assay_layer_or_slot <- function(obj, layer, slot) {
  assay <- Seurat::DefaultAssay(obj)
  mat <- tryCatch(
    SeuratObject::LayerData(obj, layer = layer, assay = assay),
    error = function(e) {
      Seurat::GetAssayData(obj, assay = assay, slot = slot)
    }
  )
  mat
}

build_maligexpr_patient_matrix <- function(
    obj,
    dataset_name,
    patient_col,
    malig_cfg,
    min_detect_frac = MIN_DETECT_FRAC_IN_MALIGNANT
) {
  if (is.null(malig_cfg) || is.null(malig_cfg$metadata_col) || is.null(malig_cfg$malignant_values)) {
    stop("MALIG_CELL_CONFIG missing or incomplete for ", dataset_name)
  }
  if (!patient_col %in% colnames(obj@meta.data)) {
    stop("Column '", patient_col, "' not found in metadata for ", dataset_name)
  }
  mc <- malig_cfg$metadata_col
  if (!mc %in% colnames(obj@meta.data)) {
    stop("Malignant metadata column '", mc, "' not found for ", dataset_name)
  }

  malig_vals <- as.character(malig_cfg$malignant_values)
  malig_cell_ids <- rownames(obj@meta.data)[as.character(obj@meta.data[[mc]]) %in% malig_vals]
  malig_cell_ids <- intersect(malig_cell_ids, colnames(obj))
  if (length(malig_cell_ids) == 0L) {
    stop("No malignant cells for ", dataset_name, " (", mc, " %in% ", paste(malig_vals, collapse = ", "), ")")
  }

  obj_malignant <- subset(x = obj, cells = malig_cell_ids)
  malig_cells <- colnames(obj_malignant)
  meta_m <- obj_malignant@meta.data
  meta_full <- obj@meta.data

  counts_mat <- get_assay_layer_or_slot(obj_malignant, layer = "counts", slot = "counts")
  data_mat <- get_assay_layer_or_slot(obj_malignant, layer = "data", slot = "data")

  if (length(malig_cells) == 0L) {
    stop("Malignant subset has no cells for ", dataset_name)
  }

  C <- counts_mat[, malig_cells, drop = FALSE]
  n_m <- ncol(C)
  detect_frac <- Matrix::rowSums(C > 0) / n_m
  genes_keep <- rownames(C)[detect_frac > min_detect_frac]
  genes_keep <- intersect(genes_keep, rownames(data_mat))
  if (length(genes_keep) == 0L) {
    stop("No genes with detection fraction > ", min_detect_frac, " in malignant cells for ", dataset_name)
  }

  D <- data_mat[genes_keep, malig_cells, drop = FALSE]
  pat_by_cell <- setNames(as.character(meta_m[[patient_col]]), rownames(meta_m))
  patients <- sort(unique(c(
    as.character(meta_full[[patient_col]]),
    as.character(pat_by_cell)
  )))
  patients <- patients[!is.na(patients) & patients != ""]

  mat <- matrix(
    NA_real_,
    nrow = length(genes_keep),
    ncol = length(patients),
    dimnames = list(genes_keep, patients)
  )

  for (p in patients) {
    cc <- names(pat_by_cell)[pat_by_cell == p]
    if (length(cc) == 0L) next
    mat[, p] <- Matrix::rowMeans(D[, cc, drop = FALSE], na.rm = TRUE)
  }

  mat <- as.matrix(mat)
  storage.mode(mat) <- "double"
  mat
}

write_matrix_dims <- function(mat, path) {
  n_patients <- ncol(mat)
  n_genes <- nrow(mat)
  lines <- c(
    paste0("n_patients\t", n_patients),
    paste0("n_genes\t", n_genes),
    "layout\trows = genes (detected in > min fraction of malignant cells), columns = patients; values = mean data layer in malignant cells"
  )
  writeLines(lines, path)
  message(
    "Dimensions: ", n_patients, " patients (columns) × ", n_genes,
    " genes (rows); wrote ", path
  )
}

if (isTRUE(SAVE_MATRICES)) {
  dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
}

expr_mats <- purrr::imap(
  SEURAT_BY_DATASET,
  function(obj, ds) {
    patient_col <- unname(PATIENT_COL_BY_DATASET[[ds]])
    if (is.null(patient_col) || (length(patient_col) == 1L && is.na(patient_col))) {
      message("No PATIENT_COL_BY_DATASET entry for ", ds, "; skipping.")
      return(NULL)
    }
    malig_cfg <- MALIG_CELL_CONFIG[[ds]]
    tryCatch(
      build_maligexpr_patient_matrix(obj, ds, patient_col, malig_cfg),
      error = function(e) {
        message("Failed to build matrix for ", ds, ": ", conditionMessage(e))
        NULL
      }
    )
  }
)
expr_mats <- purrr::compact(expr_mats)

message(
  "Built malignant-cell mean-expression × patient matrices for ", length(expr_mats), " dataset(s): ",
  paste(names(expr_mats), collapse = ", ")
)

if (isTRUE(SAVE_MATRICES) && length(expr_mats) > 0) {
  if (isTRUE(SAVE_PER_DATASET_RDS)) {
    purrr::iwalk(expr_mats, function(mat, ds) {
      rds_path <- file.path(OUT_DIR, paste0(ds, "_pseudobulk_matrix.rds"))
      saveRDS(mat, file = rds_path)
      dims_path <- file.path(OUT_DIR, paste0(ds, "_pseudobulk_matrix.rds_dims.txt"))
      write_matrix_dims(mat, dims_path)
    })
    message("Saved per-dataset matrices and dimension summaries to: ", OUT_DIR)
  }

  if (isTRUE(SAVE_ALL_AS_RDATA)) {
    save(expr_mats, file = file.path(OUT_DIR, "expr_mats_gene_by_patient_mean_expr.RData"))
    message("Saved all matrices as .RData to: ", OUT_DIR)
  }
}

# Example access:
# expr_mats[["Kurten_HNSC"]][1:5, 1:5]
```

#Menachem Playground
```{r}
Datasets.table <- read.csv("/local/ofir/sklarz/sample_cell_summary_tirosh.tsv", sep = "\t")
```


